---
title: 'Lab 02: Lasso Illustration'
output:
  html_document:
    mathjax: "default" # to display LaTeX equations correctly in external browsers
---

```{r}
suppressMessages({
library(foreach)
library(glmnet)
library(kableExtra)
library(stargazer)
})
```

We illustrate Lasso using simulated data.

# DGP

We use the same polynomial regression model as before:

Consider a polynomial regression model:

\begin{align}
Y_i & = \beta_0+\beta_1 X_{i}+\beta_2 X_i^2+\ldots+\beta_{p_0} X_i^{p_0}+U_i,\\
U_i &\sim N(0,1),\\
X_i &\sim N(0,1).
\end{align}

A custom function to generate data:
```{r}
DGP<-function(n,p0){
  #p0 = the number of polynomial terms
  #n = the number of observations
  X<-rnorm(n)
  U<-rnorm(n)
  Y<-1+U
  for (j in 1:p0){
    Y<-Y+X^j
  }
  return(list(y=Y,x=X))
}
```

Generate data with four polynomial terms:
```{r}
n=30
p0=4
set.seed(42)
MyData<-DGP(n=n,p0=p0)
```


# Lasso

One of the R functions for Lasso is `glmnet()` from the package `glmnet()`:

* `glmnet(x,y,alpha=1)` where
  - `x` = matrix of observations on the regressors
  - `y` = vector of observations on the dependent variable
  - `alpha=1` instructs to run Lasso; `alpha=0` runs Ridge
* We will use the `model.matrix()` function to create the matrix of observations on the regressors. 
  - The option `[,-1]` instructs it to drop the first column in the matrix, which is the intercept.

Let's run lasso with 10 polynomial terms:
```{r}
p=10
X=model.matrix(MyData$y ~ poly(MyData$x,p=p,raw=TRUE))[,-1]
Lasso<-glmnet(x=X,
              y=MyData$y,
              #standardize = F,
              alpha=1)
names(Lasso)
```

* We named the output object from `glmnet()` as `Lasso`.
* `Lasso$lambda` contains the vector of values for the penalty parameter generated automatically by `glmnet()`. 
  - You can also supply your own `lambda` values by using `lambda=` option in `glmnet()`.
* `Lasso$beta` contains the vector of estimated coefficients for each value of `Lasso$lambda`.

For example:
```{r}
Lasso$lambda[7]
```

```{r}
Lasso$beta[,7]
```


Let's compare the coefficients for different values of the penalty parameter `lambda`. We use the `kable()` function from the package `kableExtra` to construct a table:
```{r}
#indexes of lambda's to pick
lambda_ind_1=1
lambda_ind_2=10
lambda_ind_3=25
lambda_ind_4=length(Lasso$lambda)

ind=seq(from=1,to=p, by=1)
cbind(ind,Lasso$beta[,lambda_ind_1],Lasso$beta[,lambda_ind_2],Lasso$beta[,lambda_ind_3],Lasso$beta[,lambda_ind_4]) %>%
  kable(digits=3,
        align=c(rep('c',times=4)),
        caption="Estimated Lasso coefficients for different $\\lambda$ penalty parameter values",
        row.names = FALSE,
        col.names=c("Polynom. terms",paste0("$\\lambda=$ ",round(Lasso$lambda[lambda_ind_1],digits=2)),paste0("$\\lambda=$ ",round(Lasso$lambda[lambda_ind_2],digits=2)),paste0("$\\lambda=$ ",round(Lasso$lambda[lambda_ind_3],digits=2)),paste0("$\\lambda=$ ",round(Lasso$lambda[lambda_ind_4],digits=2)))
        ) %>%
  kable_classic(full_width = F, html_font = "Cambria")
```



*  **Note** that by default, `glmnet()` standardizes all regressors to have unit variances. More on that later.
  - For our purposes, we only care if the coefficient is zero or not.
*  Smaller `lambda` tends to produce more non-zero coefficients.


```{r}
Lasso_s<-glmnet(x=X,
              y=MyData$y,
              lambda=0.001,
              alpha=1)
Lasso_s$beta
```

* Now all ten regressors are included.
* Choosing the penalty parameter properly is crucial.
* `glmnet()` comes with a set of tools (cross-validation) designed to choose the best `lambda` for **prediction**.
  - May not be always good for regressors' selection.


# Lasso selection with estimation-targeted penalty parameter `lambda` 


```{r}
Lasso_s<-glmnet(x=X,
              y=MyData$y,
              lambda=2*sqrt(2*log(n*p)/n),
              alpha=1)
Lasso_s$beta
```



# MC simulations for Lasso's accuracy

A custom function to generate data, run Lasso, and check included regressors:
```{r}
Selected<-function(n,lambda){
  #n=sample size
  MyData<-DGP(n=n,p0=4)
  X=model.matrix(MyData$y ~ poly(MyData$x,p=10,raw=TRUE))[,-1]
  Lasso.mc<-glmnet(x=X,
              y=MyData$y,
              alpha=1,
              lambda=lambda,
              )
  IN_r<-sum(Lasso.mc$beta[1:4] != 0)
  IN_w<-sum(Lasso.mc$beta[5:10] !=0)
  return(list(right=IN_r,wrong=IN_w))
}
```


Let's run the simulations 100 times: 
```{r}
set.seed(42)
OUT<-foreach(r=1:100, .combine='rbind') %do% Selected(n=1000,lambda=2*sqrt(log(2*n*10)/n))
cat('Ave number of correct regressors included = ',mean(unlist(lapply(OUT[,1],as.vector,))),'\n')
cat('Ave number of wrong regressors included   = ',mean(unlist(lapply(OUT[,2],as.vector))))
```

Very accurate selection in large samples!


# Small coefficients, Lasso, bias

Let's change the DGP:

\begin{align}
Y_i & = \beta_0+ D_i+\beta X_{i} +U_i,\\
D_i &=X_i+ \rho\cdot N(0,1), \\
X_i &\sim N(0,1),\\
U_i &\sim N(0,0.01).
\end{align}

* $D_i$ is the main regressor of interest.
* The control $X_i$ is generated to be correlated with $D_i$ depending on the value of $\rho$.
  - Smaller $\rho$ implies a stronger relationship.
* The coefficient on $X_i$ may be small. 

A new custom function to generate data:
```{r}
DGP2<-function(n,rho,beta){
  #n = number of observations
  #rho= determines the relationship between the main regressor D and control X. Small rho -> strong relationship
  #beta= coefficient on the control X.
  X<-rnorm(n)
  D<-X+rho*rnorm(n)
  U<-rnorm(n)
  Y<-D+beta*X+0.1*U
  return(list(Y=Y,X=X,D=D))
}
```


Let's generate data with a small $\rho=0.1$, a small $\beta =2/\sqrt{n}$, and run Lasso:
```{r}
set.seed(42)
MyData<-DGP2(n=100,rho=.1,beta=2/sqrt(n))
X=model.matrix(MyData$Y ~ MyData$D+MyData$X)[,-1]
Lasso2<-glmnet(x=X,
              y=MyData$Y,
              alpha=1,
              lambda=2*sqrt(log(2*n*10)/n),
              )
coef(Lasso2)
```


* $X_i$ is dropped because it has a small coefficient!


Let's see if omitting $X_i$ causes bias:
```{r}
model.long<-lm(MyData$Y~MyData$D+MyData$X)
model.short<-lm(MyData$Y~MyData$D)
stargazer(model.long,model.short,
          type="text",
          omit.stat=c("ll","ser","f","n")
)
```

* When the control $X_i$ is omitted, we see a substantial bias in the estimated coefficient on the main regressor $D_i$: approx 35%!
  - This despite $X_i$ having a small coefficient.
  - The reason: a strong relationship between $D_i$ and $X_i$.
  - Conclusion: We should be more careful with controls that have a strong relationship with the main regressor.



---