---
title: "Lab 01: BIC illustration"
output:
  html_document:
    mathjax: "default" # to display LaTeX equations correctly in external browsers
---


```{r}
suppressMessages({
library(foreach) #flexible looping construct
library(stargazer) # for tables
})
```

We illustrate the use of the BIC model selection procedures using simulated data from a polynomial regression model.

# DGP

Consider a polynomial regression model with a polynomial of order $p_0$:

\begin{align}
Y_i & = \beta_0+\beta_1 X_{i}+\beta_2 X_i^2+\ldots+\beta_{p_0} X_i^{p_0}+U_i,\\
U_i &\sim N(0,1),\\
X_i &\sim N(0,1).
\end{align}

Let's define a custom R function to generate the data. Here, all $\beta$'s are set to one:
```{r}
DGP<-function(n,p0){
  #p0 = the number of polynomial terms
  #n = the number of observations
  X<-rnorm(n)
  U<-rnorm(n)
  Y<-1+U
  for (j in 1:p0){
    Y<-Y+X^j
  }
  return(list(y=Y,x=X))
}
```


Let's simulate some data. To make the results reproducible, we will use a specified state/seed of the random number generator. Each time you run the code, the same sequence of ovservations is simulated.
```{r}
n=30
p0=4
set.seed(42)
MyData<-DGP(n=n,p0=p0)
```


# Polynomial regression

We can obtain the OLS estimator for the polynomial regression using the function `poly(x,p,raw=TRUE)`, where:

- `x` is the data vector used to construct the polynomial terms.
- `p` is the degree of the polynomial.
- `raw=TRUE` instructs R to **not** use orthogonal polynomial terms.

Below, we estimate the model with the true number of the polynomial terms `p=p0`:

```{r, results="asis"}
model4<-lm(MyData$y ~ poly(MyData$x,p=p0,raw=TRUE))
stargazer(model4,
          covariate.labels=c(as.character(seq(from=1,to=4,by=1))),
          omit.stat=c("ll","f","ser","n","adj.rsq")
          ,type="html",notes.append = FALSE,notes = c("<sup>&sstarf;</sup>p<0.1; <sup>&sstarf;&sstarf;</sup>p<0.05; <sup>&sstarf;&sstarf;&sstarf;</sup>p<0.01")
          #,type="text"
          )
```
<br>

```{r}
model4<-lm(MyData$y ~ poly(MyData$x,p=p0,raw=TRUE))
summary(model4)
```


* All estimates are significant.
* All estimates are reasonably close to one (the true value)


Now let's increase the number of regressors (polynomial terms) beyond what was used to generate the data: `p=10`

```{r, results="asis"}
model10<-lm(MyData$y ~ poly(MyData$x,p=10,raw=TRUE))
stargazer(model10,
          covariate.labels=c(as.character(seq(from=1,to=10,by=1)),"Intercept"),
          omit.stat=c("ll","f","ser","n","adj.rsq")
          ,type="html",notes.append = FALSE,notes = c("<sup>&sstarf;</sup>p<0.1; <sup>&sstarf;&sstarf;</sup>p<0.05; <sup>&sstarf;&sstarf;&sstarf;</sup>p<0.01")
          #,type="text"
          )
```
<br>

* No significant estimates.
* The estimates for the first four parameters can substantially deviate from the true values due to the large variances.
* The variances (and the standard errors) are large because the extra regressors are correlated with the relevant regressors.

Let's try a misspecified model:

```{r, results="asis"}
model3<-lm(MyData$y ~ poly(MyData$x,p=3,raw=TRUE))
stargazer(model3,
          covariate.labels=c(as.character(seq(from=1,to=3,by=1)),"Intercept"),
          omit.stat=c("ll","f","ser","n","adj.rsq")
          ,type="html",notes.append = FALSE,notes = c("<sup>&sstarf;</sup>p<0.1; <sup>&sstarf;&sstarf;</sup>p<0.05; <sup>&sstarf;&sstarf;&sstarf;</sup>p<0.01")
          #,type="text"
          )
```
<br>

* Biased estimates due to the omitted variables

# BIC

- R `BIC()` function computes BIC


```{r}
cat("BIC for: \n",
   '- correct model         =', BIC(model4),'\n', 
   '- overspecified model   =', BIC(model10),'\n',
   '- under-specified model =', BIC(model3),'\n')
```
 
 
The custom function below runs all polynomial regression up to the `max_p` order, and picks the one with the lowest BIC value:
```{r}
MySelect<-function(MyData,max_p){
  Powers<-seq(from=1,to=max_p,by=1)
  myBICs<-foreach(j=Powers) %do% BIC(lm(MyData$y ~ poly(MyData$x,p=j,raw=TRUE)))
    p_hat<-which.min(myBICs)
  return(p_hat)
}
```

Let's try BIC selection:
```{r}
MySelect(MyData,max_p=10)
```



# Monter Carlo simulations for the accuracy of BIC in selecting the correct model

Let's check how likely BIC to select the right model using MC simulations.

- `n` is the sample size.
- `p0` is the true degree.
- `max_p` is the largest degree to try.
- `R` is the number of MC repetitions.
- `set.seed(seed)` for reproducability

```{r}
MC_BIC<-function(n,p0,max_p,R,seed){
    set.seed(seed)
    Success<-foreach(r=1:R, .combine='rbind')  %do% {
      MyData<-DGP(n,p0=p0)
      p_hat<-MySelect(MyData,max_p=max_p)
      (p_hat==p0)
  }
  cat("With n=",n, "the probability of BIC selecting the true model is",mean(Success))
}
```


```{r}
MC_BIC(n=30,p0=4,max_p=10,R=10^2,seed=42)
```

```{r}
MC_BIC(n=100,p0=4,max_p=10,R=10^2,seed=42)
```


# What if some of the coefficients are small?

We re-define the function that generates the coefficients:

- The coefficient on the term of order $p_0$ is now **small**:

```{r}
DGP<-function(n,p0){
  X<-rnorm(n)
  U<-rnorm(n)
  Y<-U+0.01*X^p0
  for (j in 1:p0-1){
    Y<-Y+X^j
  }
  return(list(y=Y,x=X))
}
```


```{r}
MC_BIC(n=1000,p0=4,max_p=10,R=10^2,seed=42)
```


- Now, the probability of selecting the true order $p_0$ is much smaller.
- BIC thinks the $p_0$ term is irrelevant.

# Parallel loops

#### Without parallel loops

Execution time:
```{r}
execution_time <- system.time({
  # Code to measure
  result <- MC_BIC(n=100,p0=4,max_p=10,R=10^3,seed=42)
})
cat("\n Execution time:\n")
print(execution_time)
```
* Over 3 secs sequential execution of the loop


#### Set up parallel loops
```{r}
library(doParallel) 
library(doRNG)

cl <- makeCluster(8) #Use 8 workers/cores
registerDoParallel(cl)
invisible(clusterEvalQ(cl, {
  suppressPackageStartupMessages(library(foreach, quietly = TRUE))
})) #load library foreach for all workers
```


Parallel version of the MC function: 
- use `%dopar%` instead.
- `export=c("DGP","MySelect"))` to ensure that workers see custom functions 
```{r}
MC_BIC_par<-function(n,p0,max_p,R){
    Success<-foreach(r=1:R, .combine='rbind',.export=c("DGP","MySelect"))  %dopar% {
      MyData<-DGP(n,p0=p0)
      p_hat<-MySelect(MyData,max_p=max_p)
      (p_hat==p0)
  }
  cat("With n=",n, "the probability of BIC selecting the true model is",mean(Success))
}
```

Run parallel loop:
```{r}
registerDoRNG(42)
execution_time <- system.time({
  # Code to measure
  result <- MC_BIC_par(n=100,p0=4,max_p=10,R=10^2)
})
cat("\n Execution time:\n")
print(execution_time)
```

* Execution time: 0.025 sec!

---