---
title: "Lecture 14: Treatment effects and difference-in-differences"
subtitle: "Economics 326 — Introduction to Econometrics II"
author:
  - name: "Vadim Marmer, UBC"
format:
  html:
    output-file: 326_14_treatment_did.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: katex
  pdf:
    output-file: 326_14_treatment_did.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
  revealjs:
    output-file: 326_14_treatment_did_slides.html
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: katex
    scrollable: true
    chalkboard: false
    self-contained: true
    transition: none
---

## Motivation: causal questions

::: {.hidden}
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
\gdef\Vhat#1{\widehat{\mathrm{Var}}\left(#1\right)}
\gdef\se#1{\mathrm{se}\left(#1\right)}
:::

- Many important questions in economics are **causal**:
  - Does job training increase earnings?
  - Does a new drug improve health outcomes?
  - Does building an incinerator reduce nearby house prices?

- The **fundamental challenge**: we can never observe the same individual both with and without treatment at the same time.

## Potential outcomes

- For each individual $i$, define two **potential outcomes**:
  - $Y_i(1)$: outcome if individual $i$ receives treatment ($D_i = 1$),
  - $Y_i(0)$: outcome if individual $i$ does not receive treatment ($D_i = 0$).

- The **individual treatment effect** for person $i$ is:

  $$
  Y_i(1) - Y_i(0).
  $$

- Example: if $Y_i$ is earnings and $D_i$ indicates job training, then $Y_i(1) - Y_i(0)$ is the causal effect of training on earnings for person $i$.

## The fundamental problem

- Let $D_i \in \{0, 1\}$ denote treatment status. The **observed outcome** is:

  $$
  Y_i = D_i \, Y_i(1) + (1 - D_i) \, Y_i(0).
  $$

- If $D_i = 1$, we observe $Y_i(1)$ but not $Y_i(0)$.
- If $D_i = 0$, we observe $Y_i(0)$ but not $Y_i(1)$.

- We can never observe both potential outcomes for the same individual. This is the **fundamental problem of causal inference**.

## Average treatment effects

- Since individual treatment effects $Y_i(1) - Y_i(0)$ are unobservable, we focus on averages.

- **Average Treatment Effect (ATE):**

  $$
  \text{ATE} = \E{Y_i(1) - Y_i(0)}.
  $$

- **Average Treatment Effect on the Treated (ATT):**

  $$
  \text{ATT} = \E{Y_i(1) - Y_i(0) \mid D_i = 1}.
  $$

- ATE averages over the entire population; ATT averages only over those who actually receive treatment.

## Selection bias

- A naive comparison of treated and untreated outcomes yields:

  \begin{align*}
  &\E{Y_i \mid D_i = 1} - \E{Y_i \mid D_i = 0} \\
  &= \E{Y_i(1) \mid D_i = 1} - \E{Y_i(0) \mid D_i = 0}.
  \end{align*}

- Add and subtract $\E{Y_i(0) \mid D_i = 1}$:

  \begin{align*}
  &= \underbrace{\E{Y_i(1) - Y_i(0) \mid D_i = 1}}_{\text{ATT}} \\
  &\quad + \underbrace{\E{Y_i(0) \mid D_i = 1} - \E{Y_i(0) \mid D_i = 0}}_{\text{Selection bias}}.
  \end{align*}

- **Selection bias** arises when the treatment and control groups differ in their baseline outcomes $Y_i(0)$. For example, people who choose to enroll in job training may differ systematically from those who do not.

## Random assignment

- Under **random assignment**, treatment $D_i$ is independent of potential outcomes:

  $$
  \E{Y_i(0) \mid D_i = 1} = \E{Y_i(0) \mid D_i = 0} = \E{Y_i(0)}.
  $$

- The selection bias vanishes, and the simple difference in means equals the ATE:

  $$
  \E{Y_i \mid D_i = 1} - \E{Y_i \mid D_i = 0} = \text{ATE} = \text{ATT}.
  $$

- Randomized experiments (like the National Supported Work program) achieve this by randomly assigning individuals to treatment and control groups.

## Regression with a treatment dummy

- With random assignment, the ATE can be estimated by regressing $Y_i$ on a treatment dummy:

  $$
  Y_i = \alpha + \tau D_i + U_i,
  $$

  where $\E{U_i \mid D_i} = 0$ under randomization.

- The OLS estimate of $\tau$ equals the difference in sample means:

  $$
  \hat{\tau} = \bar{Y}_1 - \bar{Y}_0,
  $$

  where $\bar{Y}_1$ and $\bar{Y}_0$ are the sample averages for the treated and control groups.

## Example: Lalonde data

- The **National Supported Work (NSW)** program randomly assigned disadvantaged workers to receive job training.

- We use the `jtrain2` dataset from the `wooldridge` package, based on the Lalonde (1986) study:

  ```{r}
  #| echo: true
  #| message: false
  library(wooldridge)
  data(jtrain2)
  head(jtrain2[, c("train", "re78", "educ", "age", "black", "married")], n = 10)
  ```

- `train`: 1 if assigned to job training, 0 otherwise.
- `re78`: real earnings in 1978 (in dollars).

## Estimating the ATE

- Since treatment was randomly assigned, the coefficient on `train` estimates the ATE:

  ```{r}
  #| echo: true
  summary(lm(re78 ~ train, data = jtrain2))$coefficients
  ```

- The estimated ATE is approximately \$1,794: on average, participation in the job training program increased 1978 earnings by about \$1,794.

## Observational studies

- In many settings, treatment is **not randomly assigned**. Individuals self-select into treatment.

- Example: workers who voluntarily enroll in job training may be more motivated or have lower baseline earnings than those who do not enroll.

- Self-selection generates **selection bias**, and the simple difference in means no longer estimates the ATE.

- To estimate treatment effects from observational data, we need to **control for covariates** that affect both treatment selection and outcomes.

## Potential outcomes with a covariate

- Suppose the potential outcomes depend linearly on a covariate $X_i$:

  \begin{align*}
  Y_i(0) &= \alpha_0 + \beta_0 X_i + U_i(0), \\
  Y_i(1) &= \alpha_1 + \beta_1 X_i + U_i(1),
  \end{align*}

  where $\E{U_i(0) \mid X_i, D_i} = 0$ and $\E{U_i(1) \mid X_i, D_i} = 0$.

- These assumptions mean that, after conditioning on $X_i$, treatment assignment $D_i$ is as good as random. This is called **conditional mean independence** (or "selection on observables").

## The ATE with a covariate

- Taking expectations of the potential outcomes:

  \begin{align*}
  \E{Y_i(1)} &= \alpha_1 + \beta_1 \E{X_i}, \\
  \E{Y_i(0)} &= \alpha_0 + \beta_0 \E{X_i}.
  \end{align*}

- The ATE is:

  \begin{align*}
  \text{ATE} &= \E{Y_i(1) - Y_i(0)} \\
  &= (\alpha_1 - \alpha_0) + (\beta_1 - \beta_0)\E{X_i}.
  \end{align*}

- The ATE depends on the difference in intercepts **and** the difference in slopes, weighted by the population mean of $X_i$.

## Two separate regressions

- One approach: estimate separate regressions for each group.

- **Control group** ($D_i = 0$): $Y_i = \alpha_0 + \beta_0 X_i + U_i(0).$

- **Treatment group** ($D_i = 1$): $Y_i = \alpha_1 + \beta_1 X_i + U_i(1).$

- The estimated ATE is:

  \begin{align*}
  \widehat{\text{ATE}} &= (\hat{\alpha}_1 - \hat{\alpha}_0) \\
  &\quad + (\hat{\beta}_1 - \hat{\beta}_0)\bar{X},
  \end{align*}

  where $\bar{X}$ is the overall sample mean of $X_i$.

## Combined regression with interactions

- The observed outcome $Y_i = D_i Y_i(1) + (1 - D_i) Y_i(0)$ can be written as a single regression. Expanding:

  \begin{align*}
  Y_i &= \alpha_0(1 - D_i) + \alpha_1 D_i \\
  &\quad + \beta_0 X_i(1 - D_i) + \beta_1 X_i D_i + \tilde{U}_i \\
  &= \alpha_0 + (\alpha_1 - \alpha_0)D_i \\
  &\quad + \beta_0 X_i + (\beta_1 - \beta_0)X_i D_i + \tilde{U}_i,
  \end{align*}

  where $\tilde{U}_i = (1 - D_i)U_i(0) + D_i U_i(1)$.

- This is a regression of $Y_i$ on $D_i$, $X_i$, and the interaction $X_i D_i$.

- The coefficient on $D_i$ is $\alpha_1 - \alpha_0$, which is **not** the ATE (unless $\beta_1 = \beta_0$).

## The demeaning trick

- Write $X_i = \E{X_i} + (X_i - \E{X_i})$. Then the interaction term becomes:

  \begin{align*}
  (\beta_1 - \beta_0)X_i D_i &= (\beta_1 - \beta_0)\E{X_i} D_i \\
  &\quad + (\beta_1 - \beta_0)(X_i - \E{X_i})D_i.
  \end{align*}

- Substituting into the regression:

  \begin{align*}
  Y_i &= \alpha_0 + \Big[(\alpha_1 - \alpha_0) + (\beta_1 - \beta_0)\E{X_i}\Big] D_i \\
  &\quad + \beta_0 X_i + (\beta_1 - \beta_0)(X_i - \E{X_i})D_i + \tilde{U}_i.
  \end{align*}

- Define $\tau = (\alpha_1 - \alpha_0) + (\beta_1 - \beta_0)\E{X_i}$ and $\delta = \beta_1 - \beta_0$:

  \begin{align*}
  Y_i &= \alpha_0 + \tau D_i + \beta_0 X_i \\
  &\quad + \delta(X_i - \E{X_i})D_i + \tilde{U}_i.
  \end{align*}

- The coefficient $\tau$ on $D_i$ is exactly the **ATE**.

## Estimating the ATE with covariates

- In practice, replace $\E{X_i}$ with the sample mean $\bar{X}$ and run the regression:

  \begin{align*}
  Y_i &= \hat{\alpha}_0 + \hat{\tau} D_i + \hat{\beta}_0 X_i \\
  &\quad + \hat{\delta}(X_i - \bar{X})D_i + \text{residual}.
  \end{align*}

- The coefficient $\hat{\tau}$ on $D_i$ directly estimates the ATE.

- This works because demeaning the interaction "absorbs" the $(\beta_1 - \beta_0)\E{X_i}$ part of the ATE into the coefficient on $D_i$.

## Why not regress $Y_i$ on $D_i$ and $X_i$?

- A simpler regression omits the interaction:

  $$
  Y_i = a + \tau D_i + b X_i + V_i.
  $$

- This is valid if $\beta_1 = \beta_0$ (the covariate has the same effect in both groups). Then $\delta = 0$, the interaction drops out, and the coefficient on $D_i$ is the ATE.

- If $\beta_1 \neq \beta_0$, the omitted interaction creates bias. The regression with the demeaned interaction nests the simpler model as a special case.

## Example: separate regressions

- Estimate separate regressions of `re78` on `educ` for the treatment and control groups:

  ```{r}
  #| echo: true
  reg0 <- lm(re78 ~ educ, data = jtrain2, subset = (train == 0))
  reg1 <- lm(re78 ~ educ, data = jtrain2, subset = (train == 1))
  cbind(Control = coef(reg0), Treatment = coef(reg1))
  ```

- Compute the estimated ATE:

  ```{r}
  #| echo: true
  xbar <- mean(jtrain2$educ)
  a0 <- coef(reg0)[1]; b0 <- coef(reg0)[2]
  a1 <- coef(reg1)[1]; b1 <- coef(reg1)[2]
  ATE_manual <- (a1 - a0) + (b1 - b0) * xbar
  cat("Sample mean of educ:", round(xbar, 2), "\n")
  cat("Estimated ATE:", round(ATE_manual, 2), "\n")
  ```

## Example: demeaned regression

- Create the demeaned interaction and run the combined regression:

  ```{r}
  #| echo: true
  jtrain2$educ_dm <- jtrain2$educ - mean(jtrain2$educ)
  reg_dm <- lm(re78 ~ train + educ + I(train * educ_dm), data = jtrain2)
  summary(reg_dm)$coefficients
  ```

- The coefficient on `train` directly estimates the ATE, matching the result from the separate regressions approach.

## Example: regression lines

- The two regression lines, with a vertical line at $\bar{X}$ and the ATE marked as the gap:

  ```{r}
  #| echo: false
  #| fig-align: center
  #| fig-width: 8
  #| fig-height: 5.5

  educ_range <- seq(5, 18, length.out = 100)
  yhat0 <- coef(reg0)[1] + coef(reg0)[2] * educ_range
  yhat1 <- coef(reg1)[1] + coef(reg1)[2] * educ_range

  plot(educ_range, yhat0, type = "l", lwd = 2, col = "blue",
       xlab = "Education (years)", ylab = "Earnings in 1978 ($)",
       ylim = range(c(yhat0, yhat1)),
       main = "Separate regression lines by treatment group")
  lines(educ_range, yhat1, lwd = 2, col = "red", lty = 2)

  # Vertical line at xbar
  xbar <- mean(jtrain2$educ)
  abline(v = xbar, lty = 3, col = "gray50")

  # ATE arrow at xbar
  y0_at_xbar <- coef(reg0)[1] + coef(reg0)[2] * xbar
  y1_at_xbar <- coef(reg1)[1] + coef(reg1)[2] * xbar
  arrows(xbar + 0.3, y0_at_xbar, xbar + 0.3, y1_at_xbar,
         code = 3, length = 0.1, lwd = 1.5, col = "black")
  text(xbar + 0.7, (y0_at_xbar + y1_at_xbar) / 2, "ATE", cex = 1.1)
  text(xbar, min(yhat0, yhat1) - 300, expression(bar(X)), cex = 1.1)

  legend("topleft", legend = c("Control (train = 0)", "Treatment (train = 1)"),
         col = c("blue", "red"), lty = c(1, 2), lwd = 2)
  ```

## Example: adding more covariates

- So far, we used only `educ` as the control. The dataset contains more pre-treatment characteristics: `age`, `black`, `married`.

- With random assignment, slopes are approximately equal across groups, so we can add controls directly without demeaned interactions:

  ```{r}
  #| echo: true
  reg_X <- lm(re78 ~ train + educ + age + black + married, data = jtrain2)
  round(summary(reg_X)$coefficients, 1)
  ```

## Comparison: simple vs. controlled

- Compare the estimated ATE and its standard error across specifications:

  ```{r}
  #| echo: true
  m1 <- lm(re78 ~ train, data = jtrain2)
  m2 <- lm(re78 ~ train + educ + age + black + married, data = jtrain2)
  tab <- rbind(
    "No controls"   = c(Estimate = coef(m1)["train"],
                        SE       = summary(m1)$coef["train", "Std. Error"]),
    "With controls" = c(Estimate = coef(m2)["train"],
                        SE       = summary(m2)$coef["train", "Std. Error"])
  )
  round(tab, 0)
  ```

- The estimated ATE barely changes. The standard error shrinks when controls are added.

## Why do controls change little here?

- Under **random assignment**, $D_i$ is independent of all covariates: the treatment and control groups are balanced in expectation.

- Because of balance, including $X_i$ does not change the coefficient on $D_i$: there is no selection bias to remove.

## Why does the standard error shrink?

- Recall that the variance of the OLS estimator depends on the residual variance $\hat{\sigma}^2$:

  $$
  \Vhat{\hat{\tau}} = \frac{\hat{\sigma}^2}{\sum_{i=1}^{n}(D_i - \bar{D})^2 (1 - R_D^2)},
  $$

  where $R_D^2$ is the $R^2$ from regressing $D_i$ on the other regressors.

- With randomization, $D_i$ is nearly uncorrelated with covariates, so $R_D^2 \approx 0$: the denominator stays roughly the same.

- However, covariates that predict earnings absorb variation in $Y_i$, reducing $\hat{\sigma}^2$. Since $\se{\hat{\tau}} \propto \hat{\sigma}$, the standard error shrinks.

- In short: **adding covariates under randomization buys precision without changing the point estimate**.

## Example: observational data (jtrain3)

- The `jtrain3` dataset from the `wooldridge` package combines the same NSW trainees with a **non-experimental** control group drawn from the Current Population Survey (CPS).

- In this dataset, treatment is **not** randomly assigned: CPS respondents differ systematically from NSW participants.

  ```{r}
  #| echo: true
  data(jtrain3)
  cat("n =", nrow(jtrain3), " (train=1:", sum(jtrain3$train == 1),
      ", train=0:", sum(jtrain3$train == 0), ")\n")
  ```

## Observational: without controls

- A simple regression of earnings on the training dummy:

  ```{r}
  #| echo: true
  obs1 <- lm(re78 ~ train, data = jtrain3)
  round(summary(obs1)$coefficients, 3)
  ```

- The estimated "effect" is **large and negative**: trainees appear to earn far less than the CPS comparison group.

- This is pure **selection bias**: individuals selected into the NSW program were disadvantaged, with much lower baseline earnings.

## Observational: with controls

- Adding covariates dramatically changes the estimate:

  ```{r}
  #| echo: true
  obs2 <- lm(re78 ~ train + educ + age + black + hisp + married + re74 + re75,
              data = jtrain3)
  round(summary(obs2)$coefficients, 3)
  ```

- The coefficient on `train` shifts from about $-$15,200 to about $+$900, much closer to the experimental benchmark of $+$1,794.

## Experimental vs. observational

- Side-by-side comparison:

  ```{r}
  #| echo: true
  make_row <- function(m) c(Estimate = coef(m)["train"],
                            SE = summary(m)$coef["train", "Std. Error"])
  tab <- rbind(
    "Experimental: no controls"   = make_row(m1),
    "Experimental: with controls" = make_row(m2),
    "Observational: no controls"  = make_row(obs1),
    "Observational: with controls"= make_row(obs2)
  )
  round(tab, 3)
  ```

- **Experimental data:** controls barely change the estimate (1.79 vs. 1.68); standard errors shrink.

- **Observational data:** controls remove most of the selection bias ($-$15.2 → $+$0.9); covariates are essential for a credible estimate.

## From cross-sections to panel data

- So far, we used **cross-sectional** data and assumed selection on observables: after controlling for $X_i$, treatment is as good as random.

- In some settings, this assumption is hard to justify. An alternative approach exploits **panel data** (repeated observations on the same units over time).

- The **difference-in-differences (DID)** method compares changes over time between a treatment group and a control group.

## DID setup

- Two time periods: $t \in \{0, 1\}$ (before and after treatment).

- Two groups: $D_i \in \{0, 1\}$ (control and treatment).

- Treatment occurs between periods 0 and 1, and only the treatment group ($D_i = 1$) is affected.

- We observe $Y_{it}$: the outcome for individual $i$ at time $t$.

## DID regression model

- The DID regression is:

  $$
  Y_{it} = \alpha + \delta \cdot t + \gamma D_i + \beta(t \cdot D_i) + U_{it},
  $$

  where $\E{U_{it} \mid D_i} = 0$.

- The 2×2 table of conditional means:

  | | $D_i = 0$ (Control) | $D_i = 1$ (Treatment) |
  |---|---|---|
  | $t = 0$ | $\alpha$ | $\alpha + \gamma$ |
  | $t = 1$ | $\alpha + \delta$ | $\alpha + \delta + \gamma + \beta$ |

## Interpreting the coefficients

- $\alpha$: baseline expected outcome (control group, before treatment).

- $\delta$: **time effect** — the change in the control group's outcome from $t = 0$ to $t = 1$. This captures common trends (e.g., inflation, economic growth).

- $\gamma$: **group difference** at baseline — the pre-existing difference between treatment and control groups at $t = 0$.

- $\beta$: **DID estimand** — the additional change in the treatment group's outcome, beyond what the control group experienced.

## Deriving the DID estimand

- For each combination of $t$ and $D_i$, take expectations using $\E{U_{it} \mid D_i} = 0$:

  \begin{align*}
  t = 0,\ D_i = 0: \quad & \E{Y_{i0} \mid D_i = 0} = \alpha, \\
  t = 1,\ D_i = 0: \quad & \E{Y_{i1} \mid D_i = 0} = \alpha + \delta, \\
  t = 0,\ D_i = 1: \quad & \E{Y_{i0} \mid D_i = 1} = \alpha + \gamma, \\
  t = 1,\ D_i = 1: \quad & \E{Y_{i1} \mid D_i = 1} = \alpha + \delta + \gamma + \beta.
  \end{align*}

- Subtracting the control group change from the treatment group change:

  \begin{align*}
  \beta &= \Big(\E{Y_{i1} \mid D_i = 1} - \E{Y_{i0} \mid D_i = 1}\Big) \\
  &\quad - \Big(\E{Y_{i1} \mid D_i = 0} - \E{Y_{i0} \mid D_i = 0}\Big).
  \end{align*}

- The DID estimand is the **difference in within-group changes** over time:

  \begin{align*}
  \beta &= \E{Y_{i1} - Y_{i0} \mid D_i = 1} \\
  &\quad - \E{Y_{i1} - Y_{i0} \mid D_i = 0}.
  \end{align*}

## DID diagram

- The classic DID diagram shows the control group, treatment group, and counterfactual:

  ```{r}
  #| echo: false
  #| fig-align: center
  #| fig-width: 8
  #| fig-height: 5.5

  alpha <- 5; delta <- 2; gamma <- 1; beta_did <- 3

  y_ctrl <- c(alpha, alpha + delta)
  y_treat <- c(alpha + gamma, alpha + delta + gamma + beta_did)
  y_cf <- c(alpha + gamma, alpha + delta + gamma)

  plot(c(0, 1), y_ctrl, type = "b", pch = 16, lwd = 2, col = "blue",
       xlab = "Time period", ylab = "Outcome",
       xlim = c(-0.1, 1.3), ylim = c(3, 13),
       xaxt = "n", main = "Difference-in-differences")
  axis(1, at = c(0, 1), labels = c("t = 0\n(Before)", "t = 1\n(After)"))
  lines(c(0, 1), y_treat, type = "b", pch = 16, lwd = 2, col = "red")
  lines(c(0, 1), y_cf, type = "b", pch = 1, lwd = 2, col = "gray50", lty = 2)

  arrows(1.05, y_cf[2], 1.05, y_treat[2],
         code = 3, length = 0.08, lwd = 1.5)
  text(1.15, (y_cf[2] + y_treat[2]) / 2, expression(beta), cex = 1.3)

  text(1.05, y_ctrl[2], "Control", pos = 4, col = "blue", cex = 0.9)
  text(1.05, y_treat[2] + 0.3, "Treatment", pos = 4, col = "red", cex = 0.9)
  text(1.05, y_cf[2] - 0.3, "Counterfactual", pos = 4, col = "gray50", cex = 0.9)

  arrows(-0.05, y_ctrl[1], -0.05, y_ctrl[2],
         code = 3, length = 0.08, lwd = 1, col = "blue")
  text(-0.1, (y_ctrl[1] + y_ctrl[2]) / 2, expression(delta), cex = 1,
       col = "blue", pos = 2)
  ```

## Example: incinerator and house prices

- Kiel and McClain (1995) studied how the construction of a garbage incinerator affected nearby house prices in North Andover, Massachusetts.

- We use the `kielmc` dataset from the `wooldridge` package:

  ```{r}
  #| echo: true
  data(kielmc)
  head(kielmc[, c("rprice", "y81", "nearinc", "y81nrinc", "age")], n = 10)
  ```

- `rprice`: house price in 1978 dollars.
- `y81`: 1 if year is 1981 (after incinerator announced), 0 if 1978.
- `nearinc`: 1 if house is near the incinerator site.
- `y81nrinc`: interaction `y81 * nearinc`.

## The 2×2 table of means

- Compute the four group means:

  ```{r}
  #| echo: true
  means <- tapply(kielmc$rprice, list(kielmc$y81, kielmc$nearinc), mean)
  colnames(means) <- c("Far (nearinc=0)", "Near (nearinc=1)")
  rownames(means) <- c("1978 (y81=0)", "1981 (y81=1)")
  round(means)
  ```

- Computing the DID by hand:

  ```{r}
  #| echo: true
  diff_near <- means[2, 2] - means[1, 2]
  diff_far  <- means[2, 1] - means[1, 1]
  DID <- diff_near - diff_far
  cat("Change (near):", round(diff_near), "\n")
  cat("Change (far): ", round(diff_far), "\n")
  cat("DID:          ", round(DID), "\n")
  ```

## DID regression

- The DID regression:

  ```{r}
  #| echo: true
  reg_did <- lm(rprice ~ y81 + nearinc + y81nrinc, data = kielmc)
  summary(reg_did)$coefficients
  ```

- The coefficient on `y81nrinc` matches the DID computed from the 2×2 table.

- The estimated effect is negative (incinerator reduced nearby prices), but the p-value is around 0.11, so it is not statistically significant at the 5% level.

## DID with covariates

- Adding house characteristics as controls can improve precision:

  ```{r}
  #| echo: true
  reg_did_cov <- lm(rprice ~ y81 + nearinc + y81nrinc + age + I(age^2),
                     data = kielmc)
  summary(reg_did_cov)$coefficients
  ```

- After controlling for house age, the DID estimate becomes larger in magnitude and statistically significant.

- Controlling for covariates reduces residual variance, leading to more precise estimates.

## DID and potential outcomes

- To connect DID with the potential outcomes framework, define **panel potential outcomes**: $Y_{it}(d)$ is the outcome for individual $i$ at time $t$ if assigned to group $d \in \{0, 1\}$.

- The observed outcome is:

  $$
  Y_{it} = D_i \, Y_{it}(1) + (1 - D_i) \, Y_{it}(0).
  $$

- What we observe for each group:

  | | Control ($D_i = 0$) | Treatment ($D_i = 1$) |
  |---|---|---|
  | $t = 0$ | $Y_{i0}(0)$ | $Y_{i0}(1)$ |
  | $t = 1$ | $Y_{i1}(0)$ | $Y_{i1}(1)$ |

- The treatment effect at time $t = 1$ for the treated group is:

  $$
  \text{ATT} = \E{Y_{i1}(1) - Y_{i1}(0) \mid D_i = 1}.
  $$

  The counterfactual $Y_{i1}(0)$ is unobserved for the treated group.

## DID as a treatment effect

- Recall that $\beta = \E{Y_{i1} - Y_{i0} \mid D_i = 1} - \E{Y_{i1} - Y_{i0} \mid D_i = 0}$.

- Substituting observed outcomes with potential outcomes:

  \begin{align*}
  \beta &= \E{Y_{i1}(1) - Y_{i0}(1) \mid D_i = 1} \\
  &\quad - \E{Y_{i1}(0) - Y_{i0}(0) \mid D_i = 0}.
  \end{align*}

- To relate $\beta$ to the ATT, add and subtract $\E{Y_{i1}(0) - Y_{i0}(0) \mid D_i = 1}$:

  \begin{align*}
  \beta &= \underbrace{\E{Y_{i1}(1) - Y_{i1}(0) \mid D_i = 1}}_{\text{ATT}} \\
  &\quad + \E{Y_{i1}(0) - Y_{i0}(1) \mid D_i = 1} \\
  &\quad - \E{Y_{i1}(0) - Y_{i0}(0) \mid D_i = 0}.
  \end{align*}

- For $\beta$ to equal the ATT, we need two additional assumptions.

## Assumption 1: no anticipation

- **No anticipation:** the outcome at $t = 0$ (before treatment) is not affected by future treatment group assignment:

  $$
  \E{Y_{i0}(1) \mid D_i = 1} = \E{Y_{i0}(0) \mid D_i = 1}.
  $$

- Being assigned to the treatment group does not change pre-treatment outcomes in expectation.

- In the incinerator example: before the incinerator was announced, living near the future site did not affect house prices (relative to what they would have been otherwise).

- Under no anticipation, we can replace $Y_{i0}(1)$ with $Y_{i0}(0)$ in the expression for $\beta$:

  \begin{align*}
  \beta &= \text{ATT} + \E{Y_{i1}(0) - Y_{i0}(0) \mid D_i = 1} \\
  &\quad - \E{Y_{i1}(0) - Y_{i0}(0) \mid D_i = 0}.
  \end{align*}

## Assumption 2: parallel trends

- **Parallel trends:** in the absence of treatment, both groups would have experienced the same change over time:

  $$
  \E{Y_{i1}(0) - Y_{i0}(0) \mid D_i = 0} = \E{Y_{i1}(0) - Y_{i0}(0) \mid D_i = 1}.
  $$

- Under both no anticipation and parallel trends:

  $$
  \beta = \text{ATT}.
  $$

- The parallel trends assumption cannot be directly tested because $Y_{i1}(0)$ is unobserved for the treated group. However, if pre-treatment data for multiple periods exist, one can check whether trends were parallel before treatment.

## Parallel trends diagram

- Illustrating the parallel trends assumption:

  ```{r}
  #| echo: false
  #| fig-align: center
  #| fig-width: 8
  #| fig-height: 5.5

  alpha <- 5; delta <- 2; gamma <- 1.5; beta_did <- 2.5

  y_ctrl <- c(alpha, alpha + delta)
  y_treat <- c(alpha + gamma, alpha + delta + gamma + beta_did)
  y_cf <- c(alpha + gamma, alpha + delta + gamma)

  plot(c(0, 1), y_ctrl, type = "b", pch = 16, lwd = 2, col = "blue",
       xlab = "Time period", ylab = "Outcome",
       xlim = c(-0.2, 1.5), ylim = c(3, 13),
       xaxt = "n", main = "Parallel trends assumption")
  axis(1, at = c(0, 1), labels = c("t = 0", "t = 1"))
  lines(c(0, 1), y_treat, type = "b", pch = 16, lwd = 2, col = "red")
  lines(c(0, 1), y_cf, type = "b", pch = 1, lwd = 2, col = "gray50", lty = 2)

  arrows(0.45, y_ctrl[1], 0.45, y_ctrl[2],
         code = 3, length = 0.06, lwd = 1, col = "blue")
  text(0.35, (y_ctrl[1] + y_ctrl[2]) / 2, expression(delta), cex = 1, col = "blue")

  arrows(0.55, y_cf[1], 0.55, y_cf[2],
         code = 3, length = 0.06, lwd = 1, col = "gray50")
  text(0.65, (y_cf[1] + y_cf[2]) / 2, expression(delta), cex = 1, col = "gray50")

  arrows(1.08, y_cf[2], 1.08, y_treat[2],
         code = 3, length = 0.08, lwd = 2)
  text(1.25, (y_cf[2] + y_treat[2]) / 2,
       expression(beta == ATT), cex = 1.1)

  text(-0.15, y_ctrl[1], "Control", col = "blue", cex = 0.9)
  text(-0.15, y_treat[1], "Treatment", col = "red", cex = 0.9)

  legend("topleft", legend = c("Control", "Treatment (actual)",
         "Treatment (counterfactual)"),
         col = c("blue", "red", "gray50"), lty = c(1, 1, 2),
         pch = c(16, 16, 1), lwd = 2, cex = 0.85)
  ```

## Summary

- **Potential outcomes** $Y_i(1)$ and $Y_i(0)$ formalize causal effects. The individual treatment effect $Y_i(1) - Y_i(0)$ is never fully observed.

- The **ATE** and **ATT** are population-level summaries of treatment effects.

- With **random assignment**, a simple regression of $Y_i$ on $D_i$ estimates the ATE.

- With **observational data**, controlling for covariates and using the **demeaning trick** (interacting $D_i$ with $X_i - \bar{X}$) allows the coefficient on $D_i$ to estimate the ATE.

- **Difference-in-differences** uses panel data to compare changes over time between groups:

  \begin{align*}
  \beta &= \E{Y_{i1} - Y_{i0} \mid D_i = 1} \\
  &\quad - \E{Y_{i1} - Y_{i0} \mid D_i = 0}.
  \end{align*}

- Under **no anticipation** and **parallel trends**, the DID estimand $\beta$ equals the ATT.
