---
title: "Lecture 16: Difference-in-differences"
subtitle: "Economics 326 — Introduction to Econometrics II"
author:
  - name: "Vadim Marmer, UBC"
format:
  html:
    output-file: 326_16_did.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: katex
    embed-resources: true
  pdf:
    output-file: 326_16_did.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
    include-in-header:
      text: |
        \usepackage{cancel}
  revealjs:
    output-file: 326_16_did_slides.html
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: katex
    scrollable: true
    chalkboard: false
    self-contained: true
    transition: none
---

## Motivation

::: {.hidden}
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
\gdef\Vhat#1{\widehat{\mathrm{Var}}\left(#1\right)}
\gdef\se#1{\mathrm{se}\left(#1\right)}
:::

- Previous lecture: treatment effects from **cross-sectional** data, assuming selection on observables (treatment is as good as random after controlling for covariates).

- Often hard to justify. Alternative: exploit data over time:
  - **Panel data**: repeated observations on the exact same units over time.
  - **Repeated cross-sections**: observations on different units from the same populations at different points in time.

- **Difference-in-differences (DID)**: compare changes over time between a treatment group and a control group.

## DID basic setup

- Two time periods: $t \in \{0, 1\}$ (before and after treatment).

- Two groups: $D_i \in \{0, 1\}$ (control and treatment).

- Treatment occurs between periods 0 and 1; only the treatment group ($D_i = 1$) is affected.

- $Y_{it}$: outcome for individual $i$ at time $t$.

## DID regression model

- DID regression:

  $$
  Y_{it} = \alpha + {\color{blue}\delta} \cdot t + {\color{purple}\gamma} D_i + {\color{teal}\beta}(t \cdot D_i) + U_{it},
  $$

  where $\E{U_{it} \mid D_i} = 0$.

- Regressors:
  - $t$: time (0 = before, 1 = after).
  - $D_i$: group (0 = control, 1 = treatment).
  - $t \cdot D_i$: interaction, equals 1 only for the treatment group after treatment.

## Interpreting the coefficients

- $$
  Y_{it} = \alpha + {\color{blue}\delta} \cdot t + {\color{purple}\gamma} D_i + {\color{teal}\beta}(t \cdot D_i) + U_{it},
  $$

- Evaluate $\E{Y_{it} \mid D_i}$ for each combination of $t$ and $D_i$:

  \begin{align*}
  t = 0,\ D_i = 0: \quad & \E{Y_{i0} \mid D_i = 0} = \alpha, \\
  t = 0,\ D_i = 1: \quad & \E{Y_{i0} \mid D_i = 1} = \alpha + {\color{purple}\gamma}, \\
  t = 1,\ D_i = 0: \quad & \E{Y_{i1} \mid D_i = 0} = \alpha + {\color{blue}\delta}, \\
  t = 1,\ D_i = 1: \quad & \E{Y_{i1} \mid D_i = 1} = \alpha + {\color{blue}\delta} + {\color{purple}\gamma} + {\color{teal}\beta}.
  \end{align*}

- Summarized as a 2×2 table:

  | | $D_i = 0$ (Control) | $D_i = 1$ (Treatment) |
  |---|---|---|
  | $t = 0$ | $\alpha$ | $\alpha + {\color{purple}\gamma}$ |
  | $t = 1$ | $\alpha + {\color{blue}\delta}$ | $\alpha + {\color{blue}\delta} + {\color{purple}\gamma} + {\color{teal}\beta}$ |

- $\alpha$: baseline (control group, $t = 0$).
- ${\color{purple}\gamma}$: pre-existing **group difference** at $t = 0$.
- ${\color{blue}\delta}$: **time effect** — change in the control group from $t = 0$ to $t = 1$ (common trend).

- Change over time for each group:
  - **Treatment:** $\E{Y_{i{\color{red}1}} \mid D_i = 1} - \E{Y_{i{\color{red}0}} \mid D_i = 1} = {\color{blue}\delta} + {\color{teal}\beta}$.
  - **Control:** $\E{Y_{i{\color{red}1}} \mid D_i = 0} - \E{Y_{i{\color{red}0}} \mid D_i = 0} = {\color{blue}\delta}$.

- Subtract control's change from treatment's change:

  $$
  ({\color{blue}\delta} + {\color{teal}\beta}) - {\color{blue}\delta} = {\color{teal}\beta}.
  $$

- DID estimand as a double difference:

  \begin{align*}
  {\color{teal}\beta} &= \E{Y_{i1} - Y_{i0} \mid D_i = 1} - \E{Y_{i1} - Y_{i0} \mid D_i = 0}.
  \end{align*}

- The common trend ${\color{blue}\delta}$ cancels, isolating ${\color{teal}\beta}$.

## DID diagram

- DID regression: $Y_{it} = \alpha + {\color{blue}\delta} \cdot t + {\color{purple}\gamma} D_i + {\color{teal}\beta}(t \cdot D_i) + U_{it}$.

- DID diagram — control, treatment, and counterfactual:

  ```{r}
  #| echo: false
  #| fig-align: center
  #| fig-width: 8
  #| fig-height: 5

  alpha <- 5; delta <- 2; gamma <- 1; beta_did <- 3

  y_ctrl <- c(alpha, alpha + delta)
  y_treat <- c(alpha + gamma, alpha + delta + gamma + beta_did)
  y_cf <- c(alpha + gamma, alpha + delta + gamma)

  plot(c(0, 1), y_ctrl, type = "b", pch = 16, lwd = 2, col = "blue",
       xlab = "Time period", ylab = "Average outcome",
       xlim = c(-0.2, 1.7), ylim = c(3, 13),
       xaxt = "n", main = "Difference-in-differences")
  axis(1, at = c(0, 1), labels = c("t = 0\n(Before)", "t = 1\n(After)"))
  lines(c(0, 1), y_treat, type = "b", pch = 16, lwd = 2, col = "red")
  lines(c(0, 1), y_cf, type = "b", pch = 1, lwd = 2, col = "gray50", lty = 2)

  arrows(1.05, y_cf[2], 1.05, y_treat[2],
         code = 3, length = 0.08, lwd = 1.5, col = "#008080")
  text(1.15, (y_cf[2] + y_treat[2]) / 2, expression(beta), cex = 1.3, col = "#008080")

  text(1.2, y_ctrl[2], "Control", pos = 4, col = "blue", cex = 0.9)
  text(1.2, y_treat[2], "Treatment", pos = 4, col = "red", cex = 0.9)
  text(1.2, y_cf[2], "Counterfactual", pos = 4, col = "gray50", cex = 0.9)

  # gamma: baseline group difference at t = 0
  arrows(-0.05, y_ctrl[1], -0.05, y_treat[1],
         code = 3, length = 0.08, lwd = 1, col = "purple")
  text(-0.1, (y_ctrl[1] + y_treat[1]) / 2, expression(gamma), cex = 1,
       col = "purple", pos = 2)

  # delta: time effect (control group change, under beta arrow)
  segments(0, alpha, 1, alpha, lty = 3, col = "blue", lwd = 0.5)
  arrows(1.05, y_ctrl[1], 1.05, y_ctrl[2],
         code = 3, length = 0.08, lwd = 1, col = "blue")
  text(1.15, (y_ctrl[1] + y_ctrl[2]) / 2, expression(delta), cex = 1,
       col = "blue")
  ```

- Counterfactual (dashed gray): treatment group's baseline $\alpha + {\color{purple}\gamma}$ plus the control group's change ${\color{blue}\delta}$.

## DID and potential outcomes

- **Panel potential outcomes**: $Y_{it}(d)$ — outcome for individual $i$ at time $t$ if assigned to group $d \in \{0, 1\}$.

- Four potential outcomes: $Y_{i0}(0)$, $Y_{i1}(0)$, $Y_{i0}(1)$, $Y_{i1}(1)$.

- The observed outcome is:

  $$
  Y_{it} = D_i \, Y_{it}(1) + (1 - D_i) \, Y_{it}(0).
  $$

- What we observe for each group:

  | | Control ($D_i = 0$) | Treatment ($D_i = 1$) |
  |---|---|---|
  | $t = 0$ | $Y_{i0}(0)$ | $Y_{i0}(1)$ |
  | $t = 1$ | $Y_{i1}(0)$ | $Y_{i1}(1)$ |

- Treatment effect on the treated at $t = 1$:

  $$
  {\color{teal}\text{ATT}} = {\color{teal}\E{Y_{i1}(1) - Y_{i1}(0) \mid D_i = 1}}.
  $$

  [The counterfactual $Y_{i1}(0)$ is unobserved for the treated group.]{style="color: red;"}

## DID as a treatment effect

- ${\color{teal}\beta} = \E{Y_{i1} - Y_{i0} \mid D_i = 1} - \E{Y_{i1} - Y_{i0} \mid D_i = 0}$.

- Substituting observed outcomes with potential outcomes:

  \begin{align*}
  \beta &= \E{{\color{teal}Y_{i1}(1)} {\color{red}- Y_{i0}(1)} \mid D_i = 1} \\
  &\quad {\color{blue}- \E{Y_{i1}(0) - Y_{i0}(0) \mid D_i = 0}}.
  \end{align*}

- To relate $\beta$ to the ATT, add and subtract $Y_{i1}(0)$ and $Y_{i0}(0)$ inside the first expectation:

  \begin{align*}
  \beta &= \E{{\color{teal}Y_{i1}(1)} {\color{red}- Y_{i0}(1)} \mid D_i = 1} {\color{blue}- \E{Y_{i1}(0) - Y_{i0}(0) \mid D_i = 0}} \\
  &= \E{{\color{teal}Y_{i1}(1)} \underbrace{{\color{teal}- Y_{i1}(0)} {\color{blue}+ Y_{i1}(0)}}_{= \, 0} \underbrace{{\color{blue}- Y_{i0}(0)} {\color{red}+ Y_{i0}(0)}}_{= \, 0} {\color{red}- Y_{i0}(1)} \mid D_i = 1} \\
  &\quad {\color{blue}- \E{Y_{i1}(0) - Y_{i0}(0) \mid D_i = 0}}.
  \end{align*}

- Rearranging and splitting the expectation:

  \begin{align*}
  \beta &= \underbrace{{\color{teal}\E{Y_{i1}(1) - Y_{i1}(0) \mid D_i = 1}}}_{\color{teal}\text{ATT}} \\
  &\quad + \underbrace{{\color{blue}\E{Y_{i1}(0) - Y_{i0}(0) \mid D_i = 1} - \E{Y_{i1}(0) - Y_{i0}(0) \mid D_i = 0}}}_{\color{blue}\text{difference in trends}} \\
  &\quad + \underbrace{{\color{red}\E{Y_{i0}(0) - Y_{i0}(1) \mid D_i = 1}}}_{\color{red}\text{anticipation effect}}.
  \end{align*}

- For ${\color{teal}\beta}$ to equal the ${\color{teal}\text{ATT}}$, the ${\color{blue}\text{difference in trends}}$ and ${\color{red}\text{anticipation effect}}$ must be zero. This requires two assumptions.

## Assumption 1: Parallel trends

- Difference in trends is zero if both groups would have experienced the same change over time absent treatment:

  $$
  {\color{blue}\E{Y_{i{\color{red}1}}(0) - Y_{i{\color{red}0}}(0) \mid D_i = 1} = \E{Y_{i{\color{red}1}}(0) - Y_{i{\color{red}0}}(0) \mid D_i = 0}}.
  $$

- Under parallel trends, the decomposition reduces to:

  $$
  {\color{teal}\beta} = {\color{teal}\text{ATT}} + \underbrace{{\color{red}\E{Y_{i0}(0) - Y_{i0}(1) \mid D_i = 1}}}_{\color{red}\text{anticipation effect}}.
  $$

- Cannot be directly tested: $Y_{i1}(0)$ is unobserved for the treated group.
- With multiple pre-treatment periods, can check whether trends were parallel before treatment.

## Assumption 2: No anticipation

- Anticipation effect is zero if [pre-treatment]{style="color: red;"} outcomes are not affected by future treatment assignment:

  $$
  \E{Y_{i{\color{red}0}}(1) \mid D_i = 1} = \E{Y_{i{\color{red}0}}(0) \mid D_i = 1}.
  $$

- Treatment assignment does not change [pre-treatment]{style="color: red;"} outcomes in expectation.

- Under both parallel trends and no anticipation:

  $$
  {\color{teal}\beta} = {\color{teal}\text{ATT}}.
  $$

## Example: incinerator and house prices

- Kiel and McClain (1995): did a garbage incinerator in North Andover, MA reduce nearby house prices?

- This is a **repeated cross-section**: the houses sold in 1981 (after) are different from the houses sold in 1978 (before).

- Data: `kielmc` from the `wooldridge` package:

  ```{r}
  #| echo: true
  #| message: false
  library(wooldridge)
  data(kielmc)
  # Show 2 observations from each of the 4 groups
  rows <- c(
    head(which(kielmc$y81 == 0 & kielmc$nearinc == 0), 2),
    head(which(kielmc$y81 == 0 & kielmc$nearinc == 1), 2),
    head(which(kielmc$y81 == 1 & kielmc$nearinc == 0), 2),
    head(which(kielmc$y81 == 1 & kielmc$nearinc == 1), 2)
  )
  kielmc[rows, c("rprice", "y81", "nearinc", "age")]
  ```

- `rprice`: house price in 1978 dollars ($Y_{it}$).
- `y81`: 1 if year is 1981 (after incinerator announced), 0 if 1978 ($t = 1$ if 1981, 0 if 1978).
- `nearinc`: 1 if house is near the incinerator site ($D_i$).

## The 2×2 table of means

- Compute the four group means:

  ```{r}
  #| echo: true
  means <- tapply(kielmc$rprice, list(kielmc$y81, kielmc$nearinc), mean)
  colnames(means) <- c("Far (nearinc=0)", "Near (nearinc=1)")
  rownames(means) <- c("1978 (y81=0)", "1981 (y81=1)")
  round(means, 2)
  ```

- Computing the DID by hand:

  ```{r}
  #| echo: true
  diff_near <- means[2, 2] - means[1, 2]
  diff_far  <- means[2, 1] - means[1, 1]
  DID <- diff_near - diff_far
  cat("Change (near):", round(diff_near, 2), "\n")
  cat("Change (far): ", round(diff_far, 2), "\n")
  cat("DID:          ", round(DID, 2), "\n")
  ```

## DID regression

- The DID regression, where `y81nrinc` $= \text{y81} \times \text{nearinc}$ is the interaction term:

  ```{r}
  #| echo: true
  options(scipen = 999)
  reg_did <- lm(rprice ~ y81 + nearinc + y81nrinc, data = kielmc)
  round(summary(reg_did)$coefficients, 4)
  ```

- Coefficient on `y81nrinc` matches the DID from the 2×2 table.

- $\hat{\beta} = -\$11{,}864$ (SE $= 7{,}457$, $p = 0.113$): negative but not significant at 5%.

## Assumptions in the incinerator example

- **Parallel trends:** without the incinerator, house prices near and far from the site would have followed the same trend over time.

- **No anticipation:** before the incinerator was announced, living near the future site did not affect house prices.

## Why add covariates? Compositional changes

- Because this is a **repeated cross-section**, the houses sold in 1981 are different from those sold in 1978.

- What if the *mix* of houses sold changed over time? Look at the average age of houses sold:

  ```{r}
  #| echo: true
  means_age <- tapply(kielmc$age, list(kielmc$y81, kielmc$nearinc), mean)
  colnames(means_age) <- c("Far (nearinc=0)", "Near (nearinc=1)")
  rownames(means_age) <- c("1978 (y81=0)", "1981 (y81=1)")
  round(means_age, 2)
  ```

- **Compositional change**: In the "Far" group, houses sold in 1981 were 4.25 years newer than in 1978. But in the "Near" group, they were **11.84 years newer**.

- The "Near" group got disproportionately newer. Since newer houses generally sell for more, this drastic change in composition artificially pushes the "Near" average prices up.

- This artificial price bump from the change in age composition partially **masked** the negative effect of the incinerator in our simple DID regression.

## DID regression with covariates

- To remove this bias, we must control for age. This adjusts for the fact that the two groups experienced different compositional changes over time.

  ```{r}
  #| echo: true
  reg_did_cov <- lm(rprice ~ y81 + nearinc + y81nrinc + age + I(age^2),
                     data = kielmc)
  round(summary(reg_did_cov)$coefficients, 4)
  ```

- With age controls: $\hat{\beta} = -\$21{,}920$ (SE $= 6{,}360$, $p = 0.0006$) — nearly twice as large, highly significant.

- Estimated bias from compositional change: $-\$11{,}864 - (-\$21{,}920) \approx +\$10{,}056$. The simple DID was severely biased upward because the "Near" houses sold in 1981 were unusually new.

## Detecting anticipation and pre-trends

- With only two time periods, anticipation effects **cannot** be separately identified.

- With **multiple pre-treatment periods**, use an **event study** design. Data span $t = -T, \ldots, -1, 0, 1, \ldots, T'$, where ${\color{red}t = 0}$ **is now the treatment date** (not "before" as in the two-period model).

- Replace the single interaction $\beta(t \cdot D_i)$ with a separate coefficient per period. The baseline is $t = -1$ (last pre-treatment period):

  \begin{align*}
  Y_{it} &= \alpha + \sum_{s \neq -1} {\color{blue}\delta_s} \cdot \mathbb{1}[t = s] + {\color{purple}\gamma} D_i \\
  &\quad + \sum_{s \neq -1} {\color{teal}\beta_s} \cdot D_i \cdot \mathbb{1}[t = s] + U_{it}.
  \end{align*}

- Expected outcomes for $t = -2$ and $t = -1$:

  | | $D_i = 0$ (Control) | $D_i = 1$ (Treatment) |
  |---|---|---|
  | $t = -1$ | $\alpha$ | $\alpha + {\color{purple}\gamma}$ |
  | $t = -2$ | $\alpha + {\color{blue}\delta_{-2}}$ | $\alpha + {\color{blue}\delta_{-2}} + {\color{purple}\gamma} + {\color{teal}\beta_{-2}}$ |
  | Backward trend ($t\!:\! -1 \to -2$) | ${\color{blue}\delta_{-2}}$ | ${\color{blue}\delta_{-2}} + {\color{teal}\beta_{-2}}$ |

- ${\color{teal}\beta_{-2}}$ is the difference in backward trends from $t = -1$ to $t = -2$ between treatment and control: $({\color{blue}\delta_{-2}} + {\color{teal}\beta_{-2}}) - {\color{blue}\delta_{-2}} = {\color{teal}\beta_{-2}}$.

- More generally, ${\color{teal}\beta_s}$ for $s \leq -2$ is the difference in backward trends from $t = -1$ to $t = s$ between treatment and control.

- No anticipation + parallel pre-trends $\implies$ ${\color{teal}\beta_s} = 0$ for all $s \leq -2$. However, we never observe $Y_{it}(0)$ for the treated group, so nonzero ${\color{teal}\beta_s}$ tests the **joint** hypothesis: we cannot determine which assumption failed.

- The critical point: with multiple pre-treatment periods, we can **compare the two groups before treatment**. Since ${\color{teal}\beta_s}$ for $s \leq -2$ are estimable from pre-treatment data, ${\color{teal}\beta_s} = 0$ is a **testable** implication of the joint hypothesis.

- ${\color{teal}\beta_s}$ all equal and nonzero **suggests anticipation** at $t = -1$: backward trends are the same for all $s \leq -2$, but something shifts at the baseline period.

- ${\color{teal}\beta_s}$ nonzero and unequal **suggests a parallel trends violation**: groups were already diverging before treatment.

- For $s \geq 0$: under parallel trends and no anticipation, ${\color{teal}\beta_s}$ measures the **treatment effect** at period $s$ relative to $t = -1$.

## Time fixed effects

- In the two-period model, the term ${\color{blue}\delta} \cdot t$ creates two intercepts: $\alpha$ at $t = 0$ and $\alpha + {\color{blue}\delta}$ at $t = 1$.

- With multiple periods $t = -T, \ldots, -1, 0, 1, \ldots, T'$, we cannot use ${\color{blue}\delta} \cdot t$ because that forces a **linear** trend: the time effect at period $s$ would be $\delta \cdot s$, with no flexibility.

- Instead, include a separate dummy for each period. The regression actually estimated is:

  \begin{align*}
  Y_{it} &= \alpha + \cdots + {\color{blue}\delta_{-2}} \cdot \mathbb{1}[t = -2] + {\color{blue}\delta_0} \cdot \mathbb{1}[t = 0] + {\color{blue}\delta_1} \cdot \mathbb{1}[t = 1] + \cdots \\
  &\quad + {\color{purple}\gamma} D_i + \cdots + U_{it}.
  \end{align*}

- The key property: $\mathbb{1}[t = s]$ equals $1$ when $t = s$ and $0$ otherwise. At $t = -1$ all dummies are zero (baseline). At any other $t$, exactly one dummy equals $1$:

  | Period | Active dummy | Intercept |
  |---|---|---|
  | $t = -1$ | none (baseline) | $\alpha$ |
  | $t = 0$ | $\mathbb{1}[t = 0] = 1$ | $\alpha + {\color{blue}\delta_0}$ |
  | $t = 1$ | $\mathbb{1}[t = 1] = 1$ | $\alpha + {\color{blue}\delta_1}$ |

- Each period gets its own intercept. The compact notation $\sum_{s \neq -1} {\color{blue}\delta_s} \cdot \mathbb{1}[t = s]$ writes the time dummies as a sum. The event study model with all the dummies written out:

  \begin{align*}
  Y_{it} &= \alpha + \cdots + {\color{blue}\delta_{-2}} \mathbb{1}[t\!=\!-2] + {\color{blue}\delta_0} \mathbb{1}[t\!=\!0] + {\color{blue}\delta_1} \mathbb{1}[t\!=\!1] + \cdots \\
  &\quad + {\color{purple}\gamma} D_i + \cdots + {\color{teal}\beta_{-2}} D_i \mathbb{1}[t\!=\!-2] + {\color{teal}\beta_0} D_i \mathbb{1}[t\!=\!0] + {\color{teal}\beta_1} D_i \mathbb{1}[t\!=\!1] + \cdots + U_{it}.
  \end{align*}

- At each $t$, exactly one time dummy survives (and at $t = -1$ none do). So $\alpha + {\color{blue}\delta_t}$ is the intercept at time $t$. Define ${\color{blue}\lambda_t} = \alpha + {\color{blue}\delta_t}$ (with $\delta_{-1} = 0$):

  $$
  \ldots, \quad {\color{blue}\lambda_{-1}} = \alpha, \quad {\color{blue}\lambda_0} = \alpha + {\color{blue}\delta_0}, \quad {\color{blue}\lambda_1} = \alpha + {\color{blue}\delta_1}, \quad \ldots
  $$

  The ${\color{blue}\lambda_t}$'s are called **time fixed effects**. This is the regression we actually run in OLS, with a dummy for each period:

  \begin{align*}
  Y_{it} &= \cdots + {\color{blue}\lambda_{-1}} \mathbb{1}[t\!=\!-1] + {\color{blue}\lambda_0} \mathbb{1}[t\!=\!0] + {\color{blue}\lambda_1} \mathbb{1}[t\!=\!1] + \cdots \\
  &\quad + {\color{purple}\gamma} D_i + \cdots + {\color{teal}\beta_{-2}} D_i \mathbb{1}[t\!=\!-2] + {\color{teal}\beta_0} D_i \mathbb{1}[t\!=\!0] + {\color{teal}\beta_1} D_i \mathbb{1}[t\!=\!1] + \cdots + U_{it}.
  \end{align*}

- At each $t$, exactly one $\lambda$-dummy equals $1$ and all others are zero:

  $$
  \underbrace{\cdots + {\color{blue}\lambda_t} \cdot 1 + \cdots}_{\text{only } {\color{blue}\lambda_t} \text{ survives}} = {\color{blue}\lambda_t}.
  $$

  The notation ${\color{blue}\lambda_t}$ is shorthand for this — it represents whichever $\lambda$ is active at time $t$:

  $$
  Y_{it} = {\color{blue}\lambda_t} + {\color{purple}\gamma} D_i + \sum_{s \neq -1} {\color{teal}\beta_s} \cdot D_i \cdot \mathbb{1}[t = s] + U_{it}.
  $$

## Individual fixed effects

- **Note**: The following requires **panel data** (observing the exact same individuals over time). It cannot be used with repeated cross-sections like the incinerator example.

- The same idea can be applied to individuals. In the event study model, the intercept for individual $i$ is $\alpha + {\color{purple}\gamma} D_i$. This allows only **two** values:

  | Group | Intercept |
  |---|---|
  | Control ($D_i = 0$) | $\alpha$ |
  | Treatment ($D_i = 1$) | $\alpha + {\color{purple}\gamma}$ |

  All individuals within the same group share the same baseline. But individuals may differ even within a group (e.g., houses near the incinerator differ in size, age, neighborhood quality).

- **Individual fixed effects** give each individual its own dummy, just like time fixed effects give each period its own dummy. The regression is:

  \begin{align*}
  Y_{it} &= {\color{purple}\alpha_1} \mathbb{1}[i\!=\!1] + \cdots + {\color{purple}\alpha_n} \mathbb{1}[i\!=\!n] \\
  &\quad + \text{(time and treatment terms)} + U_{it}.
  \end{align*}

  Only the dummy for individual $i$ is active, so the intercept is ${\color{purple}\alpha_i}$. Using time fixed effects ${\color{blue}\lambda_t}$ from the previous slide:

  \begin{align*}
  Y_{it} &= {\color{purple}\alpha_i} + {\color{blue}\lambda_t} + {\color{red}\cancel{{\color{purple}\gamma} D_i}} \\
  &\quad + \sum_{s \neq -1} {\color{teal}\beta_s} \cdot D_i \cdot \mathbb{1}[t = s] + U_{it}.
  \end{align*}

- Since $D_i$ does **not change over time** for any individual, it is already captured by ${\color{purple}\alpha_i}$. Including both ${\color{purple}\alpha_i}$ and ${\color{purple}\gamma} D_i$ would be **perfect multicollinearity**, so we drop ${\color{purple}\gamma} D_i$:

  \begin{align*}
  Y_{it} &= {\color{purple}\alpha_i} + {\color{blue}\lambda_t} \\
  &\quad + \sum_{s \neq -1} {\color{teal}\beta_s} \cdot D_i \cdot \mathbb{1}[t = s] + U_{it}.
  \end{align*}

- The interaction terms ${\color{teal}\beta_s} \cdot D_i \cdot \mathbb{1}[t = s]$ survive because they vary across **both** individuals and time.

## Two-way fixed effects

- Combining individual and time fixed effects gives the **two-way fixed effects** (TWFE) model:

  \begin{align*}
  Y_{it} &= {\color{purple}\alpha_i} + {\color{blue}\lambda_t} \\
  &\quad + \sum_{s \neq -1} {\color{teal}\beta_s} \cdot D_i \cdot \mathbb{1}[t = s] + U_{it},
  \end{align*}

  where ${\color{purple}\alpha_i}$ is the individual fixed effect and ${\color{blue}\lambda_t}$ is the time fixed effect.

- This is the standard **event study** specification used in the literature for DID with multiple periods of **panel data**.
