---
title: "Lecture 3: Simple Linear Regression and OLS"
subtitle: "Economics 326 — Introduction to Econometrics II"
author:
  - name: "Vadim Marmer, UBC"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    output-file: 326_03_simple.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: katex
    include-in-header:
      text: |
        <style>
          .tight-figure p {
            margin-top: 0;
            margin-bottom: 0.75rem;
          }
          .tight-figure img {
            display: block;
            margin-left: auto;
            margin-right: auto;
          }
          .figure-crop {
            overflow: hidden;
            margin-bottom: 0.75rem;
          }
          .figure-crop p {
            margin: 0;
          }
          .figure-crop img {
            width: 100%;
            height: auto;
          }
          .figure-crop-51 {
            aspect-ratio: 1.582;
          }
          .figure-crop-50 {
            aspect-ratio: 1.556;
          }
          .figure-crop-51 img {
            transform: translateY(-50.9%);
          }
          .figure-crop-50 img {
            transform: translateY(-50.3%);
          }
          .figure-crop-small {
            width: 80%;
            margin-left: auto;
            margin-right: auto;
          }
        </style>
  pdf:
    output-file: 326_03_simple.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
  revealjs:
    output-file: 326_03_simple_slides.html
    date: ""
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: katex
    scrollable: true
    chalkboard: false
    self-contained: true
    transition: none
    include-in-header:
      text: |
        <style>
          .tight-figure p {
            margin-top: 0;
            margin-bottom: 0.75rem;
          }
          .tight-figure img {
            display: block;
            margin-left: auto;
            margin-right: auto;
          }
          .figure-crop {
            overflow: hidden;
            margin-bottom: 0.75rem;
          }
          .figure-crop p {
            margin: 0;
          }
          .figure-crop img {
            width: 100%;
            height: auto;
          }
          .figure-crop-51 {
            aspect-ratio: 1.582;
          }
          .figure-crop-50 {
            aspect-ratio: 1.556;
          }
          .figure-crop-51 img {
            transform: translateY(-50.9%);
          }
          .figure-crop-50 img {
            transform: translateY(-50.3%);
          }
          .figure-crop-small {
            width: 80%;
            margin-left: auto;
            margin-right: auto;
          }
          .reveal .slide img {
            max-height: 950px;
            width: auto;
            height: auto;
          }
        </style>
---

# Introduction and Definitions

## Overview of Simple Linear Regression

::: {.hidden}
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
:::

- The **simple linear regression model** is used to study the relationship between **two variables**.
- It has many limitations, but nevertheless there are examples in the literature where the simple linear regression is applied (e.g., stock returns predictability).
- It is also a good starting point for learning the regression technique.

## Key Definitions

**Terminology for Simple Regression**

| $y$ | $x$ |
|-----|-----|
| Dependent variable | Independent variable |
| Explained variable | Explanatory variable |
| Response variable | Control variable |
| Predicted variable | Predictor variable |
| Regressand | Regressor |

## Sample vs Population

- The econometrician observes random data:

  | observation | dependent variable | regressor |
  |------------|--------------------|-----------|
  | 1 | $Y_{1}$ | $X_{1}$ |
  | 2 | $Y_{2}$ | $X_{2}$ |
  | $\vdots$ | $\vdots$ | $\vdots$ |
  | $n$ | $Y_{n}$ | $X_{n}$ |

- A pair $X_{i}, Y_{i}$ is called an observation.
- **Sample**: $\left\{ \left( X_{i},Y_{i}\right) : i=1,\ldots,n\right\}$.
- The population is the **joint distribution** of the sample.

# Simple Linear Regression Model

## Conditional Expectation Function

- We model the relationship between $Y$ and $X$ using the **conditional expectation**:
$$ \E{Y_{i}|X_{i}} = \alpha + \beta X_{i}. $$

- **Intercept**: $\alpha = \E{Y_{i}|X_{i}=0}$.

- **Slope**: $\beta$ measures the effect of a unit change in $X$ on $Y$:
$$ \begin{aligned}
\beta &= \E{Y_{i}|X_{i}=x+1} - \E{Y_{i}|X_{i}=x} \\ &= \left[ \alpha + \beta (x+1)\right] - \left[ \alpha + \beta x\right].
\end{aligned} $$

- **Marginal effect** of $X$ on $Y$:
$$ \beta = \frac{d\E{Y_{i}|X_{i}}}{dX_{i}}. $$

- **The effect is the same for all $x$**.

## Regression Error Term

- $\alpha$ and $\beta$ in $\E{Y_{i}|X_{i}} = \alpha + \beta X_{i}$ are **unknown**.
- **Residual (error)**:
$$ U_{i} = Y_{i} - \E{Y_{i}|X_{i}} = Y_{i} - \left( \alpha + \beta X_{i}\right). $$
  $U_{i}$'s are **unobservable**.

- **The model**:
$$ \begin{aligned}
Y_{i} &= \alpha + \beta X_{i} + U_{i}, \\ \E{U_{i}|X_{i}} &= 0.
\end{aligned} $$

## Functional Form: Linearity

- We consider a model that is linear in the coefficients $\alpha,\beta$: $Y_{i} = \alpha + \beta X_{i} + U_{i}$.
- The dependent variable and the regressor can be nonlinear functions of some other variables.
- The most popular function is $\log$.

## Log-Linear Specification

- Consider the following model:
$$ \log Y_{i} = \alpha + \beta X_{i} + U_{i}. $$

- In this case,
$$ \begin{aligned}
\beta &= \frac{d\left( \log Y_{i}\right)}{dX_{i}} \\ &= \frac{dY_{i}/Y_{i}}{dX_{i}} = \frac{dY_{i}/dX_{i}}{Y_{i}}.
\end{aligned} $$

- $\beta$ measures **percentage** change in $Y$ as a response to a unit change in $X$.
- In this model, it is assumed that the percentage change in $Y$ is the same for all values of $X$ (constant).
- In $\log \left( \text{Wage}_{i}\right) = \alpha + \beta \times \text{Education}_{i} + U_{i}$, $\beta$ measures the **return** to education.

## Log-Log Specification

- Consider the following model:
$$ \log Y_{i} = \alpha + \beta \log X_{i} + U_{i}. $$

- In this model,
$$ \begin{aligned}
\beta &= \frac{d\log Y_{i}}{d\log X_{i}} \\ &= \frac{dY_{i}/Y_{i}}{dX_{i}/X_{i}} = \frac{dY_{i}}{dX_{i}}\frac{X_{i}}{Y_{i}}.
\end{aligned} $$

- $\beta$ measures **elasticity**: the percentage change in $Y$ as a response to 1% change in $X$.
- Here, the elasticity is assumed to be the same for all values of $X$.
- Example: Cobb-Douglas production function:
$$ Y=\alpha K^{\beta_{1}}L^{\beta_{2}} \Longrightarrow \log Y=\log \alpha + \beta_{1}\log K + \beta_{2}\log L $$ 
  (two regressors: log of capital and log of labour).

## Assumption of "Orthogonality"

The model:
$$ Y_{i} = \alpha + \beta X_{i} + U_{i}. $$

We assume that $\E{U_{i}|X_{i}} = 0$.

- $\E{U_{i}} = 0$.
$$ \E{U_{i}} \overset{\text{Law of Iterated Expectation}}{=} \E{\E{U_{i}|X_{i}}} = \E{0} = 0. $$

- $\Cov{X_{i},U_{i}} = \E{X_{i}U_{i}} = 0$.
$$ \begin{aligned}
\E{X_{i}U_{i}} &\overset{\text{Law of Iterated Expectation}}{=} \E{\E{X_{i}U_{i}|X_{i}}} \\ &= \E{X_{i} \E{U_{i}|X_{i}}} = \E{X_{i} \cdot 0} = 0.
\end{aligned} $$

## Decomposition into Predicted and Error Components

$$ Y_{i} = \underbrace{\alpha + \beta X_{i}}_{ \text{Predicted by } X} + \underbrace{U_{i}}_{ \text{"Orthogonal to" } X} $$

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4
#| fig-align: center

# Parameters
beta0 <- 1
beta1 <- 0.8
x_vals <- c(1.5, 3.5, 5.5)
sigma <- 0.6

par(mar = c(4, 4, 2, 1))
plot(NA, xlim = c(0, 7), ylim = c(0, 7),
     xlab = expression(italic(x)), ylab = expression(italic(y)),
     xaxt = "n", yaxt = "n",
     main = expression(E(italic(y)*"|"*italic(x))~"as a linear function of"~italic(x)))
axis(1, at = x_vals, labels = expression(italic(x)[1], italic(x)[2], italic(x)[3]))
axis(2, at = NULL, labels = FALSE)

# Regression line
abline(a = beta0, b = beta1, lwd = 2)

# Draw density curves at each x value
for (x in x_vals) {
  mu <- beta0 + beta1 * x
  y_seq <- seq(mu - 2.5 * sigma, mu + 2.5 * sigma, length.out = 100)
  dens <- dnorm(y_seq, mean = mu, sd = sigma)
  dens_scaled <- dens * 1.2
  lines(x + dens_scaled, y_seq, lwd = 1.5)
  points(x, mu, pch = 19, cex = 1.2)
}

# Label with line pointing to regression line (positioned to avoid density curves)
label_x <- 5.0
label_y <- 0.8
point_on_line_x <- 6.8
point_on_line_y <- beta0 + beta1 * point_on_line_x
text(label_x, label_y, expression(E(italic(y)*"|"*italic(x)) == alpha + beta*italic(x)), adj = 0, cex = 0.9)
segments(6.5, label_y + 0.3, point_on_line_x, point_on_line_y, lwd = 0.7)
```

# Estimation of Parameters

## Estimation Problem

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4
#| fig-align: center

# Simulated data for n observations
set.seed(42)
n <- 15
X <- runif(n, 1, 10)
alpha_true <- -0.5
beta_true <- 0.4
Y <- alpha_true + beta_true * X + rnorm(n, 0, 0.8)

par(mar = c(4, 4, 2, 1))
plot(X, Y, pch = 19, col = "gray40",
     xlab = expression(italic(X)), ylab = expression(italic(Y)),
     xlim = c(0, max(X) + 1), ylim = c(min(Y) - 0.5, max(Y) + 0.5),
     main = expression("Scatterplot of "*italic(Y)*" and "*italic(X)*", and the population regression"))

# Add regression line
abline(a = alpha_true, b = beta_true, lwd = 1.5)

# Add axis lines at 0
abline(h = 0, lty = 1, col = "black", lwd = 0.5)
abline(v = 0, lty = 1, col = "black", lwd = 0.5)

# Label for regression line with pointer
label_x <- 7.8
label_y <- 1.0
point_on_line_x <- 9.5
point_on_line_y <- alpha_true + beta_true * point_on_line_x
text(label_x, label_y, expression(E(italic(Y)*"|"*italic(X)) == alpha + beta*italic(X)), cex = 0.8)
segments(8.8, 1.35, point_on_line_x, point_on_line_y, lwd = 0.7)
```

**Problem:** estimate the unknown parameters $\alpha$ and $\beta$ using the data ($n$ observations) on $Y$ and $X$.

## Method of Moments Estimator

- Our assumptions imply:
$$ \begin{aligned}
\E{U_{i}} &= \E{Y_{i} - \alpha - \beta X_{i}} = 0. \\ \E{X_{i}U_{i}} &= \E{X_{i}\left( Y_{i} - \alpha - \beta X_{i}\right)} = 0.
\end{aligned} $$

- An **estimator** is a function of the observable data; it can depend only on observable $X$ and $Y$. Let $\hat{\alpha}$ and $\hat{\beta}$ denote the estimators of $\alpha$ and $\beta$.

- **Method of moments:** replace expectations with averages. **Normal equations**:
$$ \begin{aligned}
\frac{1}{n}\sum_{i=1}^{n}\left( Y_{i} - \hat{\alpha} - \hat{\beta} X_{i}\right) &= 0. \\ \frac{1}{n}\sum_{i=1}^{n}X_{i}\left( Y_{i} - \hat{\alpha} - \hat{\beta} X_{i}\right) &= 0.
\end{aligned} $$

## Normal Equations

::: {.nonincremental}
- Let $\bar{Y}=\frac{1}{n}\sum_{i=1}^{n}Y_{i}$ and $\bar{X}=\frac{1}{n}\sum_{i=1}^{n}X_{i}$ (averages).

  $$ \frac{1}{n}\sum_{i=1}^{n}\left( Y_{i} - \hat{\alpha} - \hat{\beta}X_{i}\right) = 0 $$
    implies
  $$ \begin{aligned}
\frac{1}{n}\sum_{i=1}^{n}Y_{i} - \frac{1}{n}\sum_{i=1}^{n}\hat{\alpha} - \hat{\beta}\frac{1}{n}\sum_{i=1}^{n}X_{i} &= 0 \text{ or} \\ \bar{Y} - \hat{\alpha} - \hat{\beta}\bar{X} &= 0.
\end{aligned} $$

  The fitted regression line goes through the averages.

- 
  $$ \hat{\alpha} = \bar{Y} - \hat{\beta}\bar{X}. $$
:::

## Slope Estimator Derivation

- $\hat{\alpha} = \bar{Y} - \hat{\beta}\bar{X}$ and therefore
$$ \begin{aligned}
0 &= \frac{1}{n}\sum_{i=1}^{n}X_{i}\left( Y_{i} - \hat{\alpha} - \hat{\beta} X_{i}\right) \\ &= \sum_{i=1}^{n}X_{i}\left( Y_{i} - \left( \bar{Y} - \hat{\beta}\bar{X}\right) - \hat{\beta}X_{i}\right) \\ &= \sum_{i=1}^{n}X_{i}\left[ \left( Y_{i} - \bar{Y}\right) - \hat{\beta}\left( X_{i} - \bar{X}\right) \right] \\ &= \sum_{i=1}^{n}X_{i}\left( Y_{i} - \bar{Y}\right) - \hat{\beta}\sum_{i=1}^{n}X_{i}\left( X_{i} - \bar{X}\right).
\end{aligned} $$

## Final Formulas for OLS Estimators

- 
$$ 0 = \sum_{i=1}^{n}X_{i}\left( Y_{i} - \bar{Y}\right) - \hat{\beta}\sum_{i=1}^{n}X_{i}\left( X_{i} - \bar{X}\right) \text{ or}
$$ 
$$ \hat{\beta} = \frac{\sum_{i=1}^{n}X_{i}\left( Y_{i} - \bar{Y}\right)}{\sum_{i=1}^{n}X_{i}\left( X_{i} - \bar{X}\right)}. $$

- Since
$$ \begin{aligned}
\sum_{i=1}^{n}X_{i}\left( Y_{i} - \bar{Y}\right) &= \sum_{i=1}^{n}\left( X_{i} - \bar{X}\right) \left( Y_{i} - \bar{Y}\right) = \sum_{i=1}^{n}\left( X_{i} - \bar{X}\right) Y_{i} 
\\ \sum_{i=1}^{n}X_{i}\left( X_{i} - \bar{X}\right) &= \sum_{i=1}^{n}\left( X_{i} - \bar{X}\right)\left( X_{i} - \bar{X}\right) = \sum_{i=1}^{n}\left( X_{i} - \bar{X}\right)^{2}
\end{aligned} $$
  we can also write
$$ \hat{\beta} = \frac{\sum_{i=1}^{n}\left( X_{i} - \bar{X}\right) Y_{i}}{\sum_{i=1}^{n}\left( X_{i} - \bar{X}\right)^{2}}. $$

## Fitted Values and Residuals

::: {.nonincremental}
- **Fitted values**: $\hat{Y}_{i} = \hat{\alpha} + \hat{\beta}X_{i}$.
- **Fitted residuals**: $\hat{U}_{i} = Y_{i} - \hat{\alpha} - \hat{\beta}X_{i}$.
:::

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4
#| fig-align: center

# Generate data
set.seed(42)
n <- 10
X <- runif(n, 1, 9)
alpha_hat <- 0.5
beta_hat <- 0.3
Y_hat_all <- alpha_hat + beta_hat * X
Y <- Y_hat_all + rnorm(n, 0, 0.5)

par(mar = c(4, 4, 2, 1))
plot(X, Y, pch = 19, col = "gray50",
     xlab = expression(italic(X)), ylab = expression(italic(Y)),
     xlim = c(0, 10), ylim = c(0, max(Y) + 0.8),
     main = "Fitted values and residuals")

# Fitted regression line
abline(a = alpha_hat, b = beta_hat, lwd = 1.5)

# Select one point with a clear positive residual
i <- which.max(Y - Y_hat_all)
Xi <- X[i]
Yi <- Y[i]
Y_hat_i <- alpha_hat + beta_hat * Xi

# Highlight the selected point
points(Xi, Yi, pch = 19, col = "black", cex = 1.3)

# Function to draw a right curly brace
draw_brace <- function(x, y1, y2, col, width = 0.15) {
  mid <- (y1 + y2) / 2
  # Draw curly brace shape using segments
  n <- 20
  # Top half
  t1 <- seq(0, pi/2, length.out = n)
  brace_y1 <- seq(y2, mid + (y2-y1)/8, length.out = n)
  brace_x1 <- x + width * (1 - cos(t1)) / 2
  # Bottom half
  brace_y2 <- seq(mid - (y2-y1)/8, y1, length.out = n)
  brace_x2 <- x + width * (1 - cos(rev(t1))) / 2
  # Middle point
  lines(brace_x1, brace_y1, col = col, lwd = 1.5)
  lines(brace_x2, brace_y2, col = col, lwd = 1.5)
  # Connect at middle with point
  lines(c(x + width/2, x + width), c(mid + (y2-y1)/8, mid), col = col, lwd = 1.5)
  lines(c(x + width, x + width/2), c(mid, mid - (y2-y1)/8), col = col, lwd = 1.5)
}

# Thin dashed vertical line from 0 to Yi
segments(Xi, 0, Xi, Yi, lty = 2, lwd = 0.5)

# Blue brace for Y_hat (fitted value): from 0 to Y_hat_i
draw_brace(Xi + 0.1, 0, Y_hat_i, col = "blue", width = 0.25)

# Red brace for U_hat (residual): from Y_hat_i to Yi
draw_brace(Xi + 0.1, Y_hat_i, Yi, col = "red", width = 0.25)

# Labels
text(Xi + 0.15, Yi + 0.1, expression(Y[i]), cex = 0.85, adj = 0)
text(Xi + 0.45, Y_hat_i / 2, expression(hat(Y)[i]), cex = 0.85, adj = 0, col = "blue")
text(Xi + 0.45, (Yi + Y_hat_i) / 2, expression(hat(U)[i]), cex = 0.85, adj = 0, col = "red")

# Label for X_i on x-axis
mtext(expression(X[i]), side = 1, at = Xi, line = 0.5, cex = 0.8)

# Label for regression line (moved to lower left to avoid clash)
text(1.5, 2.8, expression(hat(Y) == hat(alpha) + hat(beta)*X), cex = 0.85)
segments(2.5, 2.6, 3.5, alpha_hat + beta_hat * 3.5, lwd = 0.7)
```

## True vs Fitted Regression Lines

::: {.nonincremental}
- **True:** $Y_{i} = \alpha + \beta X_{i} + U_{i}$, $\E{U_{i}} = \E{X_{i}U_{i}} = 0$.
- **Fitted:** $Y_{i} = \hat{\alpha} + \hat{\beta} X_{i} + \hat{U}_{i}$, $\sum_{i=1}^{n}\hat{U}_{i} = \sum_{i=1}^{n}X_{i}\hat{U}_{i} = 0$.
:::

```{r}
#| echo: false
#| fig-width: 6
#| fig-height: 4
#| fig-align: center

par(mar = c(4, 4, 2, 1))

# True parameters
alpha <- 1.0
beta <- 0.2

# Fitted parameters (steeper slope for better separation)
alpha_hat <- 0.3
beta_hat <- 0.38

# Plot setup
plot(NA, xlim = c(0, 10), ylim = c(0, 4.5),
     xlab = expression(italic(X)), ylab = expression(italic(Y)),
     main = "True vs fitted regression lines")

# True regression line (solid)
abline(a = alpha, b = beta, lty = 1, lwd = 1.5)

# Fitted regression line (dashed)
abline(a = alpha_hat, b = beta_hat, lty = 2, lwd = 1.5)

# Label for fitted line (upper, steeper line)
text(5.5, 3.8, expression(hat(Y) == hat(alpha) + hat(beta)*X), cex = 0.85)
segments(6.2, 3.6, 7.5, alpha_hat + beta_hat * 7.5, lwd = 0.7)

# Label for true line (lower, flatter line)
text(5.5, 1.8, expression(E(Y~"|"~X) == alpha + beta*X), cex = 0.85)
segments(6.2, 1.95, 7.5, alpha + beta * 7.5, lwd = 0.7)
```

## Ordinary Least Squares (OLS) Minimization

- Minimize $Q\left( a,b\right) = \sum_{i=1}^{n}\left( Y_{i} - a - bX_{i}\right)^{2}$ with respect to $a$ and $b$.

- Derivatives:
$$ \begin{aligned}
\frac{dQ\left( a,b\right)}{da} &= -2\sum_{i=1}^{n}\left( Y_{i} - a - bX_{i}\right). \\ \frac{dQ\left( a,b\right)}{db} &= -2\sum_{i=1}^{n}\left( Y_{i} - a - bX_{i}\right) X_{i}.
\end{aligned} $$

- First-order conditions:
$$ \begin{aligned}
0 &= \sum_{i=1}^{n}\left( Y_{i} - \hat{\alpha} - \hat{\beta}X_{i}\right) = \sum_{i=1}^{n}\hat{U}_{i}. \\ 0 &= \sum_{i=1}^{n}\left( Y_{i} - \hat{\alpha} - \hat{\beta}X_{i}\right) X_{i} = \sum_{i=1}^{n}\hat{U}_{i}X_{i}.
\end{aligned} $$

- Method of moments = OLS:
$$ \hat{\beta} = \frac{\sum_{i=1}^{n}\left( X_{i} - \bar{X}\right) Y_{i}}{\sum_{i=1}^{n}\left( X_{i} - \bar{X}\right)^{2}} \quad \text{and} \quad \hat{\alpha} = \bar{Y} - \hat{\beta}\bar{X}.
$$
