---
title: "Lecture 14: Causal inference"
subtitle: "Economics 326 — Introduction to Econometrics II"
author:
  - name: "Vadim Marmer, UBC"
format:
  html:
    output-file: 326_14_treatment_did.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: katex
    embed-resources: true
  pdf:
    output-file: 326_14_treatment_did.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
  revealjs:
    output-file: 326_14_treatment_did_slides.html
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: katex
    scrollable: true
    chalkboard: false
    self-contained: true
    transition: none
---

## Motivation: causal questions

::: {.hidden}
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
\gdef\Vhat#1{\widehat{\mathrm{Var}}\left(#1\right)}
\gdef\se#1{\mathrm{se}\left(#1\right)}
:::

- Many important questions in economics are **causal**:
  - Does job training increase earnings?
  - Does a new drug improve health outcomes?

- The **fundamental challenge**: we can never observe the same individual both with and without treatment at the same time.
- This is modeled using the **potential outcomes framework**: each individual has two potential outcomes, one under treatment and one under control, but we only observe one of them.

## Potential outcomes

- **Treatment status**: $D_i = 1$ if individual $i$ receives treatment, $D_i = 0$ otherwise.
- For each individual $i$, define two **potential outcomes**:
  - $Y_i(1)$: outcome if individual $i$ receives treatment ($D_i = 1$),
  - $Y_i(0)$: outcome if individual $i$ does not receive treatment ($D_i = 0$).

- The **individual treatment effect** for person $i$ is:

  $$
  Y_i(1) - Y_i(0).
  $$

- Example: if $Y_i$ is earnings and $D_i$ indicates job training, then $Y_i(1) - Y_i(0)$ is the causal effect of training on earnings for person $i$.

## The fundamental problem

- The **observed outcome** is:

  $$
  Y_i = D_i \, Y_i(1) + (1 - D_i) \, Y_i(0).
  $$

- If $D_i = 1$, we observe $Y_i(1)$ but not $Y_i(0)$.
- If $D_i = 0$, we observe $Y_i(0)$ but not $Y_i(1)$.

- We can never observe both potential outcomes for the same individual. This is the **fundamental problem of causal inference**.

## Average treatment effects

- Since individual treatment effects $Y_i(1) - Y_i(0)$ are unobservable, we focus on averages.

- **Average Treatment Effect (ATE):**

  $$
  \text{ATE} = \E{Y_i(1) - Y_i(0)}.
  $$

- **Average Treatment Effect on the Treated (ATT):**

  $$
  \text{ATT} = \E{Y_i(1) - Y_i(0) \mid D_i = 1}.
  $$

- ATE averages over the entire population; ATT averages only over those who actually receive treatment.

## Selection bias

- A naive comparison of treated and untreated outcomes yields:

  \begin{align*}
  &\E{Y_i \mid D_i = 1} - \E{Y_i \mid D_i = 0} \\
  &= \E{Y_i(1) \mid D_i = 1} - \E{Y_i(0) \mid D_i = 0}.
  \end{align*}

- Add and subtract $\E{Y_i(0) \mid D_i = 1}$:

  \begin{align*}
  &= \underbrace{\E{Y_i(1) - Y_i(0) \mid D_i = 1}}_{\text{ATT}} \\
  &\quad + \underbrace{\E{Y_i(0) \mid D_i = 1} - \E{Y_i(0) \mid D_i = 0}}_{\text{Selection bias}}.
  \end{align*}

- **Selection bias** arises when the treatment and control groups differ in their baseline outcomes $Y_i(0)$.
- E.g., people who choose to enroll in job training may differ systematically from those who do not.

## Random assignment

- Under **random assignment**, treatment $D_i$ is independent of potential outcomes:

  $$
  \E{Y_i(0) \mid D_i = 1} = \E{Y_i(0) \mid D_i = 0} = \E{Y_i(0)}.
  $$

- The selection bias vanishes, and the simple difference in means equals the ATE:

  \begin{align*}
  &\E{Y_i \mid D_i = 1} - \E{Y_i \mid D_i = 0} \\
  &= \E{Y_i(1) - Y_i(0)} = \text{ATE} = \text{ATT}.
  \end{align*}

- Randomized experiments (like the National Supported Work program) achieve this by randomly assigning individuals to treatment and control groups.

## Regression with a treatment dummy

- With random assignment, the ATE can be estimated by regressing $Y_i$ on a treatment dummy:

  $$
  Y_i = \alpha + \tau D_i + U_i,
  $$

  where $\E{U_i \mid D_i} = 0$ under randomization.

- The OLS estimate of $\tau$ equals the difference in sample means:

  $$
  \hat{\tau} = \bar{Y}_1 - \bar{Y}_0,
  $$

  where $\bar{Y}_1$ and $\bar{Y}_0$ are the sample averages for the treated and control groups.

## Example: Lalonde data

- The **National Supported Work (NSW)** demonstration recruited disadvantaged workers (long-term unemployed, high-school dropouts, former drug users, ex-offenders).

- Among eligible applicants, some were **randomly assigned** to receive job training (treated group), and the rest formed the **experimental control group**. Both groups come from the same disadvantaged population.

- We use the `jtrain2` dataset from the `wooldridge` package, based on the Lalonde (1986) study:

  ```{r}
  #| echo: true
  #| message: false
  library(wooldridge)
  data(jtrain2)
  set.seed(326)
  jtrain2[sample(nrow(jtrain2), 10), c("train", "re78", "educ", "age", "black", "married")]
  ```

- `train`: 1 if randomly assigned to job training, 0 if assigned to control.
- `re78`: real earnings in 1978 (thousands of dollars).

## Estimating the ATE

- Since treatment was randomly assigned, the coefficient on `train` estimates the ATE:

  ```{r}
  #| echo: true
  summary(lm(re78 ~ train, data = jtrain2))$coefficients
  ```

- The estimated ATE is approximately \$1,800 (`re78` is in thousands): on average, participation in the job training program increased 1978 earnings by about \$1,800.

## Observational studies

- In many settings, treatment is **not randomly assigned**. Individuals self-select into treatment.

- E.g., workers who voluntarily enroll in job training may be more motivated or have lower baseline earnings than those who do not enroll.

- Similarly, workers at firms that offer 401(k) plans tend to have higher incomes, so comparing savings between eligible and ineligible workers confounds the plan effect with income differences.

- Self-selection generates **selection bias**, and the simple difference in means no longer estimates the ATE.

- To estimate treatment effects from observational data, we need to **control for covariates** that affect both treatment selection and outcomes.

## Potential outcomes with a covariate

- Suppose the potential outcomes depend linearly on a covariate $X_i$:

  \begin{align*}
  Y_i(0) &= \alpha_0 + \beta_0 X_i + U_i(0), \\
  Y_i(1) &= \alpha_1 + \beta_1 X_i + U_i(1),
  \end{align*}

  where $\E{U_i(0) \mid X_i, D_i} = 0$ and $\E{U_i(1) \mid X_i, D_i} = 0$.

- These assumptions mean that, after conditioning on $X_i$, treatment assignment $D_i$ is as good as random. This is called **conditional mean independence** (or "selection on observables").
- While $Y_i(0)$ and $Y_i(1)$ depend on $X_i$ (and thus on $D_i$), the residual terms after controlling for $X_i$, namely $U_i(0)$ and $U_i(1)$, are uncorrelated with both $X_i$ and $D_i$.
- Note that the effect of $X_i$ on the potential outcomes can differ between the treated and control groups: $\beta_1$ may not equal $\beta_0$.

## The ATE with a covariate

- Taking expectations of the potential outcomes:

  \begin{align*}
  \E{Y_i(1)} &= \alpha_1 + \beta_1 \E{X_i}, \\
  \E{Y_i(0)} &= \alpha_0 + \beta_0 \E{X_i}.
  \end{align*}

- The ATE is:

  \begin{align*}
  \text{ATE} &= \E{Y_i(1) - Y_i(0)} \\
  &= (\alpha_1 - \alpha_0) + (\beta_1 - \beta_0)\E{X_i}.
  \end{align*}

- The ATE depends on the difference in intercepts **and** the difference in slopes, weighted by the population mean of $X_i$.

## Two separate regressions

- One approach: estimate separate regressions for each group.

- **Control group** ($D_i = 0$): $Y_i = \alpha_0 + \beta_0 X_i + U_i(0).$

- **Treatment group** ($D_i = 1$): $Y_i = \alpha_1 + \beta_1 X_i + U_i(1).$

- The estimated ATE is:

  \begin{align*}
  \widehat{\text{ATE}} &= (\hat{\alpha}_1 - \hat{\alpha}_0) \\
  &\quad + (\hat{\beta}_1 - \hat{\beta}_0)\bar{X},
  \end{align*}

  where $\bar{X}$ is the overall sample mean of $X_i$.

## Combined regression with interactions

- The observed outcome $Y_i = D_i Y_i(1) + (1 - D_i) Y_i(0)$ can be written as a single regression. Expanding:

  \begin{align*}
  Y_i &= \alpha_0(1 - D_i) + \alpha_1 D_i \\
  &\quad + \beta_0 X_i(1 - D_i) + \beta_1 X_i D_i + U_i \\
  &= \alpha_0 + (\alpha_1 - \alpha_0)D_i \\
  &\quad + \beta_0 X_i + (\beta_1 - \beta_0)X_i D_i + U_i,
  \end{align*}

  where $U_i = (1 - D_i)U_i(0) + D_i U_i(1)$.

- This is a regression of $Y_i$ on $D_i$, $X_i$, and the interaction $X_i D_i$.

- The coefficient on $D_i$ is $\alpha_1 - \alpha_0$, which is **not** the ATE (unless $\beta_1 = \beta_0$).

## The demeaning trick

- Write $X_i = \E{X_i} + (X_i - \E{X_i})$. Then the interaction term becomes:

  \begin{align*}
  (\beta_1 - \beta_0)X_i D_i &= (\beta_1 - \beta_0)\E{X_i} D_i \\
  &\quad + (\beta_1 - \beta_0)(X_i - \E{X_i})D_i.
  \end{align*}

- Substituting into the regression:

  \begin{align*}
  Y_i &= \alpha_0 + \Big[(\alpha_1 - \alpha_0) + (\beta_1 - \beta_0)\E{X_i}\Big] D_i \\
  &\quad + \beta_0 X_i + (\beta_1 - \beta_0)(X_i - \E{X_i})D_i + U_i.
  \end{align*}

- Define $\tau = (\alpha_1 - \alpha_0) + (\beta_1 - \beta_0)\E{X_i}$ and $\delta = \beta_1 - \beta_0$:

  \begin{align*}
  Y_i &= \alpha_0 + \tau D_i + \beta_0 X_i \\
  &\quad + \delta(X_i - \E{X_i})D_i + U_i.
  \end{align*}

- The coefficient $\tau$ on $D_i$ is exactly the **ATE**.

## Estimating the ATE with covariates

- In practice, replace $\E{X_i}$ with the sample mean $\bar{X}$ and run the regression:

  \begin{align*}
  Y_i &= \hat{\alpha}_0 + \hat{\tau} D_i + \hat{\beta}_0 X_i \\
  &\quad + \hat{\delta}(X_i - \bar{X})D_i + \text{residual}.
  \end{align*}

- The coefficient $\hat{\tau}$ on $D_i$ directly estimates the ATE.

- This works because demeaning the interaction "absorbs" the $(\beta_1 - \beta_0)\E{X_i}$ part of the ATE into the coefficient on $D_i$.

## Why not regress $Y_i$ on $D_i$ and $X_i$?

- A simpler regression omits the interaction:

  $$
  Y_i = a + \tau D_i + b X_i + V_i.
  $$

- This is valid if $\beta_1 = \beta_0$ (the covariate has the same effect in both groups). Then $\delta = 0$, the interaction drops out, and the coefficient on $D_i$ is the ATE.

- If $\beta_1 \neq \beta_0$ (the effect of $X_i$ differs between groups), the omitted interaction creates bias: the interaction $X_iD_i$ affects $Y_i$ and is correlated with $D_i$, so the coefficient on $D_i$ does not equal the ATE.

## Example: separate regressions

- Estimate separate regressions of `re78` on `educ` for the treatment and control groups:

  ```{r}
  #| echo: true
  reg0 <- lm(re78 ~ educ, data = jtrain2, subset = (train == 0))
  reg1 <- lm(re78 ~ educ, data = jtrain2, subset = (train == 1))
  cbind(Control = coef(reg0), Treatment = coef(reg1))
  ```

- Compute the estimated ATE:

  ```{r}
  #| echo: true
  xbar <- mean(jtrain2$educ)
  a0 <- coef(reg0)[1]; b0 <- coef(reg0)[2]
  a1 <- coef(reg1)[1]; b1 <- coef(reg1)[2]
  ATE_manual <- (a1 - a0) + (b1 - b0) * xbar
  cat("Sample mean of educ:", round(xbar, 2), "\n")
  cat("Estimated ATE:", round(ATE_manual, 2), "\n")
  ```

## Example: regression lines

- The two regression lines, with a vertical line at $\bar{X}$ and the ATE marked as the gap:

  ```{r}
  #| echo: false
  #| fig-align: center
  #| fig-width: 8
  #| fig-height: 5.5

  educ_range <- seq(5, 18, length.out = 100)
  yhat0 <- coef(reg0)[1] + coef(reg0)[2] * educ_range
  yhat1 <- coef(reg1)[1] + coef(reg1)[2] * educ_range

  plot(educ_range, yhat0, type = "l", lwd = 2, col = "blue",
       xlab = "Education (years)", ylab = "Earnings in 1978 ($)",
       ylim = range(c(yhat0, yhat1)),
       main = "Separate regression lines by treatment group")
  lines(educ_range, yhat1, lwd = 2, col = "red", lty = 2)

  # Vertical line at xbar
  xbar <- mean(jtrain2$educ)
  abline(v = xbar, lty = 3, col = "gray50")

  # ATE arrow at xbar
  y0_at_xbar <- coef(reg0)[1] + coef(reg0)[2] * xbar
  y1_at_xbar <- coef(reg1)[1] + coef(reg1)[2] * xbar
  arrows(xbar + 0.3, y0_at_xbar, xbar + 0.3, y1_at_xbar,
         code = 3, length = 0.1, lwd = 1.5, col = "black")
  text(xbar + 0.7, (y0_at_xbar + y1_at_xbar) / 2, "ATE", cex = 1.1)
  text(xbar, min(yhat0, yhat1) - 300, expression(bar(X)), cex = 1.1)

  legend("topleft", legend = c("Control (train = 0)", "Treatment (train = 1)"),
         col = c("blue", "red"), lty = c(1, 2), lwd = 2)
  ```

## Example: demeaned regression

- Create the demeaned interaction and run the combined regression:

  ```{r}
  #| echo: true
  jtrain2$educ_dm <- jtrain2$educ - mean(jtrain2$educ)
  reg_dm <- lm(re78 ~ train + educ + I(train * educ_dm), data = jtrain2)
  summary(reg_dm)$coefficients
  ```

- The coefficient on `train` directly estimates the ATE, matching the result from the separate regressions approach.

## Example: adding more covariates

- So far, we used only `educ` as the control. The dataset contains more pre-treatment characteristics: `age`, `black`, `married`.

- With random assignment, slopes are approximately equal across groups, so we can add controls directly without demeaned interactions:

  ```{r}
  #| echo: true
  reg_X <- lm(re78 ~ train + educ + age + black + married, data = jtrain2)
  round(summary(reg_X)$coefficients, 1)
  ```

## Comparison: simple vs. controlled

- Compare the estimated ATE and its standard error across specifications, including one with demeaned interactions:

  ```{r}
  #| echo: true
  m1 <- lm(re78 ~ train, data = jtrain2)
  m2 <- lm(re78 ~ train + educ + age + black + married, data = jtrain2)
  covs <- c("educ", "age", "black", "married")
  for (v in covs) jtrain2[[paste0(v, "_dm")]] <- jtrain2[[v]] - mean(jtrain2[[v]])
  m3 <- lm(re78 ~ train + educ + age + black + married +
             I(train * educ_dm) + I(train * age_dm) +
             I(train * black_dm) + I(train * married_dm), data = jtrain2)
  make_row <- function(m) c(Estimate = coef(m)["train"],
                            SE = summary(m)$coef["train", "Std. Error"])
  tab <- rbind(
    "No controls"            = make_row(m1),
    "Controls, equal slopes"  = make_row(m2),
    "Demeaned interactions"   = make_row(m3)
  )
  round(tab, 2)
  ```

- All three estimates are nearly identical. The demeaned interactions barely matter because randomization ensures approximately equal slopes across groups.

## Why do controls change little here?

- Under **random assignment**, $D_i$ is independent of all covariates: the treatment and control groups are balanced in expectation.

- Because of balance, including $X_i$ does not change the coefficient on $D_i$: there is no selection bias to remove.

- The NSW program recruited a narrow, disadvantaged population: long-term unemployed, high-school dropouts, former drug users, ex-offenders.

- Within this homogeneous group, covariates like `educ`, `age`, `black`, and `married` vary little and have limited predictive power for earnings.

- Both the point estimate and the standard error are nearly unchanged because the controls add almost no information beyond the treatment dummy.

## Example: 401(k) eligibility and savings

- The `k401ksubs` dataset (from the `wooldridge` package) contains 9,275 households from the 1991 Survey of Income and Program Participation.

  ```{r}
  #| echo: true
  data(k401ksubs)
  cat("n =", nrow(k401ksubs), "\n")
  set.seed(326)
  k401ksubs[sample(nrow(k401ksubs), 8), c("e401k", "nettfa", "inc", "age", "marr")]
  ```

- **Treatment**: `e401k` = 1 if the worker's employer offers a 401(k) plan, 0 otherwise.
- **Outcome**: `nettfa` = net total financial assets (\$1000s).
- **Key covariate**: `inc` = family income (\$1000s).

- This is **observational**: eligibility depends on the employer, and higher-income workers tend to work at firms offering 401(k) plans.

## The selection problem

- Group means by eligibility status:

  ```{r}
  #| echo: true
  grp <- split(k401ksubs, k401ksubs$e401k)
  tab <- rbind(
    "Ineligible (e401k=0)" = colMeans(grp[["0"]][, c("nettfa", "inc", "age")]),
    "Eligible (e401k=1)"   = colMeans(grp[["1"]][, c("nettfa", "inc", "age")])
  )
  round(tab, 1)
  ```

- Eligible workers have much higher income and much higher savings. The raw gap in savings is **not** a treatment effect — it partly reflects income differences.

## Separate regressions

- Estimate separate regressions of `nettfa` on `inc` for each group:

  ```{r}
  #| echo: true
  reg0 <- lm(nettfa ~ inc, data = k401ksubs, subset = (e401k == 0))
  reg1 <- lm(nettfa ~ inc, data = k401ksubs, subset = (e401k == 1))
  cbind(Ineligible = coef(reg0), Eligible = coef(reg1))
  ```

- Compute the estimated ATE from the two-regression formula:

  ```{r}
  #| echo: true
  xbar <- mean(k401ksubs$inc)
  a0 <- coef(reg0)[1]; b0 <- coef(reg0)[2]
  a1 <- coef(reg1)[1]; b1 <- coef(reg1)[2]
  ATE_manual <- (a1 - a0) + (b1 - b0) * xbar
  cat("Sample mean of inc:", round(xbar, 2), "\n")
  cat("Estimated ATE:", round(ATE_manual, 2), "\n")
  ```

## Demeaned regression

- Create the demeaned interaction and run the combined regression:

  ```{r}
  #| echo: true
  k401ksubs$inc_dm <- k401ksubs$inc - mean(k401ksubs$inc)
  reg_dm <- lm(nettfa ~ e401k + inc + I(e401k * inc_dm), data = k401ksubs)
  summary(reg_dm)$coefficients
  ```

- The coefficient on `e401k` directly estimates the ATE, matching the result from the separate regressions.

## Comparison across specifications

- Four specifications for the effect of 401(k) eligibility on net financial assets:

  ```{r}
  #| echo: true
  m_none <- lm(nettfa ~ e401k, data = k401ksubs)
  m_inc  <- lm(nettfa ~ e401k + inc, data = k401ksubs)
  m_int  <- lm(nettfa ~ e401k + inc + I(e401k * inc), data = k401ksubs)

  make_row <- function(m) c(Estimate = coef(m)["e401k"],
                            SE = summary(m)$coef["e401k", "Std. Error"])
  tab <- rbind(
    "No controls"             = make_row(m_none),
    "Income, no interaction"  = make_row(m_inc),
    "Interaction (demeaned)"     = make_row(reg_dm),
    "Interaction (not demeaned)" = make_row(m_int)
  )
  round(tab, 2)
  ```

- **No controls**: selection bias inflates the estimate — eligible workers earn more and save more regardless of 401(k) access.
- **Income, no interaction**: removes much of the bias but constrains the model to equal slopes.
- **Interaction (demeaned)**: the coefficient on `e401k` is the ATE evaluated at mean income, showing 401(k) eligibility increases savings by about \$5,400.
- **Interaction (not demeaned)**: the coefficient on `e401k` is the **CATE** (conditional average treatment effect) evaluated at `inc = 0`:

  $$
  \text{CATE}(x) = \E{Y_i(1) - Y_i(0) \mid X_i = x} = (\alpha_1 - \alpha_0) + (\beta_1 - \beta_0)x.
  $$

  At $x = 0$ (zero income), this is an extrapolation far outside the data range — the **sign flips** to negative.
- Contrast with `jtrain2`: there, controls barely mattered because randomization already eliminated selection bias. Here, controls and the demeaning trick are essential.

