---
title: "Lecture 12: Hypothesis testing in multiple regression"
subtitle: "Economics 326 — Introduction to Econometrics II"
author:
  - name: "Vadim Marmer, UBC"
format:
  html:
    output-file: 326_12_mreg_testing.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: katex
  pdf:
    output-file: 326_12_mreg_testing.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
  revealjs:
    output-file: 326_12_mreg_testing_slides.html
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: katex
    scrollable: true
    chalkboard: false
    self-contained: true
    transition: none
---

## The model

::: {.hidden}
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
\gdef\Vhat#1{\widehat{\mathrm{Var}}\left(#1\right)}
\gdef\se#1{\mathrm{se}\left(#1\right)}
:::

- We consider the classical normal linear regression model:

  1. $Y_{i}=\beta_{0}+\beta_{1}X_{1,i}+\ldots+\beta_{k}X_{k,i}+U_{i}.$

  2. $\E{U_{i} \mid \mathbf{X}}=0$ for all $i$'s.

  3. $\E{U_{i}^{2} \mid \mathbf{X}}=\sigma^{2}$ for all $i$'s.

  4. $\E{U_{i}U_{j} \mid \mathbf{X}}=0$ for all $i\neq j$.

  5. Conditional on $\mathbf{X}$, $U_{i}$'s are jointly normally distributed.

- We also assume **no perfect multicollinearity**: there exist no constants $c_{1},\ldots,c_{k},c_{k+1}$ (not all zero) such that for **all $i$'s**:

  $$
  c_{1}X_{1,i}+\ldots+c_{k}X_{k,i}+c_{k+1}=0.
  $$

## Testing a single coefficient

- Take the $j$-th coefficient $\beta_{j}$, $j\in\left\{0,1,\ldots,k\right\}.$

- Under our assumptions, conditional on $\mathbf{X}$, the OLS estimator $\hat{\beta}_{j}$ satisfies

  $$
  \hat{\beta}_{j}\sim N\left(\beta_{j},\Var{\hat{\beta}_{j} \mid \mathbf{X}}\right),
  $$

  where $\Var{\hat{\beta}_{j} \mid \mathbf{X}}=\sigma^{2}/\sum_{i=1}^{n}\tilde{X}_{j,i}^{2}$ (see Lecture 11).

- Therefore, $\left(\hat{\beta}_{j}-\beta_{j}\right)/\sqrt{\Var{\hat{\beta}_{j} \mid \mathbf{X}}}\sim N\left(0,1\right).$

- Since $\sigma^{2}$ is unknown, so is $\Var{\hat{\beta}_{j} \mid \mathbf{X}}$. We estimate it by

  $$
  \Vhat{\hat{\beta}_{j}}=\frac{s^{2}}{\sum_{i=1}^{n}\tilde{X}_{j,i}^{2}},
  $$

  where $s^{2}=\sum_{i=1}^{n}\hat{U}_{i}^{2}/\left(n-k-1\right)$ (see Lecture 10).


- Conditional on $\mathbf{X}$,

  $$
  \frac{\hat{\beta}_{j}-\beta_{j}}{\sqrt{\Vhat{\hat{\beta}_{j}}}}\sim t_{n-k-1}.
  $$

- Standard error: $\se{\hat{\beta}_{j}}=\sqrt{\Vhat{\hat{\beta}_{j}}}=\sqrt{s^{2}/\sum_{i=1}^{n}\tilde{X}_{j,i}^{2}}$.

- Consider testing $H_{0}:\beta_{j}=\beta_{j,0}$ against $H_{1}:\beta_{j}\neq\beta_{j,0}$.

- Under $H_{0},$ conditional on $\mathbf{X}$,

  $$
  T=\frac{\hat{\beta}_{j}-\beta_{j,0}}{\sqrt{\Vhat{\hat{\beta}_{j}}}}\sim t_{n-k-1}.
  $$

- Let $t_{df,\tau}$ be the $\tau$-th quantile of the $t_{df}$ distribution.

- **Test:** Reject $H_{0}$ when $\left\vert T\right\vert>t_{n-k-1,1-\alpha/2}.$

- **P-value:** $\text{p-value}=2\left(1-F_{t_{n-k-1}}\!\left(\left\vert T\right\vert\right)\right),$ where $F_{t_{n-k-1}}$ is the CDF of the $t_{n-k-1}$ distribution.

## Testing a linear combination of coefficients

- Let $c_{0},c_{1},\ldots,c_{k},r$ be given constants. Consider testing

  \begin{align*}
  H_{0} &: c_{0}\beta_{0}+c_{1}\beta_{1}+\ldots+c_{k}\beta_{k}=r \text{ against} \\
  H_{1} &: c_{0}\beta_{0}+c_{1}\beta_{1}+\ldots+c_{k}\beta_{k}\neq r.
  \end{align*}

- **Example 1:** Consider $\ln Y_{i}=\beta_{0}+\beta_{1}\ln L_{i}+\beta_{2}\ln K_{i}+U_{i}.$ To test for constant returns to scale, $H_{0}:\beta_{1}+\beta_{2}=1,$ set $c_{0}=0, c_{1}=1, c_{2}=1, r=1.$

- **Example 2:** Consider $\ln\left(\text{Wage}_{i}\right) = \beta_{0}+\beta_{1}\text{Experience}_{i}+\beta_{2}\text{PrevExperience}_{i}+\ldots+U_{i}.$ To test whether the two experience variables have the same effect on wage, $H_{0}:\beta_{1}-\beta_{2}=0,$ set $c_{0}=0, c_{1}=1, c_{2}=-1, c_{3}=\ldots=c_{k}=0, r=0.$

- **Example 3:** Consider $\ln\left(\text{Wage}_{i}\right)=\beta_{0}+\beta_{1}\text{Exper}_{i}+\beta_{2}\text{Exper}_{i}^{2}+\ldots+U_{i}.$ The marginal effect of experience is $\beta_{1}+2\beta_{2}\text{Exper}_{i}.$ For a concave profile ($\beta_{2}<0$), the marginal effect is smallest at the highest experience level. To test whether the marginal effect equals zero at $\text{Exper}=20$: $H_{0}:\beta_{1}+40\beta_{2}=0,$ with $c_{1}=1, c_{2}=40, r=0.$

## Testing a linear combination of coefficients

- Conditional on $\mathbf{X}$, under $H_{0}: c_{0}\beta_{0}+c_{1}\beta_{1}+\ldots+c_{k}\beta_{k}=r,$

  \begin{align*}
  & \frac{c_{0}\hat{\beta}_{0}+c_{1}\hat{\beta}_{1}+\ldots+c_{k}\hat{\beta}_{k}-r}{\sqrt{\Var{c_{0}\hat{\beta}_{0}+c_{1}\hat{\beta}_{1}+\ldots+c_{k}\hat{\beta}_{k} \mid \mathbf{X}}}} \\
  &= \frac{c_{0}\hat{\beta}_{0}+c_{1}\hat{\beta}_{1}+\ldots+c_{k}\hat{\beta}_{k}-\left(c_{0}\beta_{0}+c_{1}\beta_{1}+\ldots+c_{k}\beta_{k}\right)}{\sqrt{\Var{c_{0}\hat{\beta}_{0}+c_{1}\hat{\beta}_{1}+\ldots+c_{k}\hat{\beta}_{k} \mid \mathbf{X}}}} \\
  & \sim N\left(0,1\right).
  \end{align*}

- The variance of the linear combination is

  \begin{align*}
  &\Var{c_{0}\hat{\beta}_{0}+c_{1}\hat{\beta}_{1}+\ldots+c_{k}\hat{\beta}_{k} \mid \mathbf{X}} \\
  &= \sum_{j=0}^{k}c_{j}^{2}\Var{\hat{\beta}_{j} \mid \mathbf{X}} + \sum_{j=0}^{k}\sum_{l\neq j}c_{j}c_{l}\Cov{\hat{\beta}_{j},\hat{\beta}_{l} \mid \mathbf{X}}.
  \end{align*}

## Testing a linear combination of coefficients

- Consider

  $$
  T=\frac{c_{0}\hat{\beta}_{0}+c_{1}\hat{\beta}_{1}+\ldots+c_{k}\hat{\beta}_{k}-r}{\sqrt{\Vhat{c_{0}\hat{\beta}_{0}+c_{1}\hat{\beta}_{1}+\ldots+c_{k}\hat{\beta}_{k}}}}.
  $$

- Under $H_{0}: c_{0}\beta_{0}+c_{1}\beta_{1}+\ldots+c_{k}\beta_{k}=r,$ conditional on $\mathbf{X}$,

  $$
  T\sim t_{n-k-1}.
  $$

- **Two-sided test:** Reject $H_{0}$ when $\left\vert T\right\vert>t_{n-k-1,1-\alpha/2}.$

## CRS test: details

- Consider the model $\ln Y_{i}=\beta_{0}+\beta_{1}\ln L_{i}+\beta_{2}\ln K_{i}+U_{i}.$

- We test for constant returns to scale: $H_{0}:\beta_{1}+\beta_{2}=1.$

- The test statistic: $T=\dfrac{\hat{\beta}_{1}+\hat{\beta}_{2}-1}{\sqrt{\Vhat{\hat{\beta}_{1}+\hat{\beta}_{2}}}}.$

- The estimated variance:

  \begin{align*}
  \Vhat{\hat{\beta}_{1}+\hat{\beta}_{2}} &= \Vhat{\hat{\beta}_{1}}+\Vhat{\hat{\beta}_{2}} \\
  &\quad +2\widehat{\mathrm{Cov}}\left(\hat{\beta}_{1},\hat{\beta}_{2}\right).
  \end{align*}

  - $\Vhat{\hat{\beta}_{1}}$ and $\Vhat{\hat{\beta}_{2}}$ are obtained from the standard errors reported by R.

  - In R, `vcov(fit)` returns the full variance-covariance matrix of the coefficient estimates, including $\widehat{\mathrm{Cov}}\left(\hat{\beta}_{1},\hat{\beta}_{2}\right)$.

- Reject $H_{0}:\beta_{1}+\beta_{2}=1$ if $\left\vert T\right\vert>t_{n-3,1-\alpha/2}.$

## Example

- We generate 1000 observations from the model:

  \begin{align*}
  &\left.\begin{array}{l}
  L_{i}=e^{l_{i}} \\
  K_{i}=e^{k_{i}}
  \end{array}\right\} \text{ where } l_{i},k_{i} \text{ are iid } N\left(0,1\right), \Cov{l_{i},k_{i}}=0.5, \\
  &U_{i}\sim \text{iid } N\left(0,1\right) \text{ is independent of } l_{i},k_{i}, \\
  &Y_{i}=L_{i}^{0.35}K_{i}^{0.52}e^{U_{i}}.
  \end{align*}

- We estimate the equation:

  $$
  \ln Y_{i}=\beta_{0}+\beta_{1}\ln L_{i}+\beta_{2}\ln K_{i}+U_{i}.
  $$

- We test $H_{0}:\beta_{1}+\beta_{2}=1$ against $H_{1}:\beta_{1}+\beta_{2}\neq 1$ at the 5% significance level.

  ```{r}
  #| echo: true
  set.seed(123)
  n <- 1000
  lnL <- rnorm(n)
  lnK <- 0.5 * lnL + sqrt(1 - 0.5^2) * rnorm(n)
  U <- rnorm(n)
  lnY <- 0.35 * lnL + 0.52 * lnK + U
  ```

## Example: regression output

- Regression output:

  ```{r}
  #| echo: true
  fit <- lm(lnY ~ lnL + lnK)
  summary(fit)
  ```

- The variance-covariance matrix of the coefficient estimates:

  ```{r}
  #| echo: true
  vcov(fit)
  ```

- The critical value $t_{n-3,0.975}$:

  ```{r}
  #| echo: true
  qt(1 - 0.025, df = fit$df.residual)
  ```

## Example: manual calculation

- From the regression output:

  ```{r}
  #| echo: true
  b1 <- coef(fit)["lnL"]
  b2 <- coef(fit)["lnK"]
  V <- vcov(fit)

  cat("b1 =", b1, "\n")
  cat("b2 =", b2, "\n")
  cat("Var(b1) =", V["lnL", "lnL"], "\n")
  cat("Var(b2) =", V["lnK", "lnK"], "\n")
  cat("Cov(b1, b2) =", V["lnL", "lnK"], "\n")
  ```

- The standard error of $\hat{\beta}_{1}+\hat{\beta}_{2}$:

  ```{r}
  #| echo: true
  se_sum <- sqrt(V["lnL", "lnL"] + V["lnK", "lnK"] + 2 * V["lnL", "lnK"])
  cat("se(b1 + b2) =", se_sum, "\n")
  ```

- The test statistic:

  ```{r}
  #| echo: true
  T_stat <- (b1 + b2 - 1) / se_sum
  cat("T =", T_stat, "\n")
  ```

- The critical value:

  ```{r}
  #| echo: true
  cv <- qt(1 - 0.025, df = fit$df.residual)
  cat("|T| =", abs(T_stat), ", critical value =", cv, "\n")
  ```

- Since $\left\vert T\right\vert > t_{997,0.975}$, we reject $H_{0}.$

- Ignoring the covariance leads to an incorrect result:

  ```{r}
  #| echo: true
  se_wrong <- sqrt(V["lnL", "lnL"] + V["lnK", "lnK"])
  T_wrong <- (b1 + b2 - 1) / se_wrong
  cat("T (ignoring covariance) =", T_wrong, "\n")
  ```

## Re-parametrization approach

- We test $\beta_{1}+\beta_{2}=1$ in $\ln Y_{i}=\beta_{0}+\beta_{1}\ln L_{i}+\beta_{2}\ln K_{i}+U_{i}.$

- Define $\delta=\beta_{1}+\beta_{2}$, or $\beta_{2}=\delta-\beta_{1},$ so that

  \begin{align*}
  \ln Y_{i} &= \beta_{0}+\beta_{1}\ln L_{i}+\beta_{2}\ln K_{i}+U_{i} \\
  &= \beta_{0}+\beta_{1}\ln L_{i}+\left(\delta-\beta_{1}\right)\ln K_{i}+U_{i} \\
  &= \beta_{0}+\beta_{1}\left(\ln L_{i}-\ln K_{i}\right)+\delta\ln K_{i}+U_{i}.
  \end{align*}

- Generate a new variable $D_{i}=\ln L_{i}-\ln K_{i}.$

- Estimate $\ln Y_{i}=\beta_{0}+\beta_{1}D_{i}+\delta\ln K_{i}+U_{i}$.

- Test $H_{0}:\delta=1$ against $H_{1}:\delta\neq 1.$

## Example: reparameterization

- Reparameterized regression output:

  ```{r}
  #| echo: true
  D <- lnL - lnK
  fit2 <- lm(lnY ~ D + lnK)
  summary(fit2)
  ```

- The 95% confidence interval for the coefficient on $\ln K$:

  ```{r}
  #| echo: true
  confint(fit2, "lnK")
  ```

- The interval does not include 1, so we reject $H_{0}.$

- In the original equation, $\hat{\beta}_{1}+\hat{\beta}_{2}$ and $\se{\hat{\beta}_{1}+\hat{\beta}_{2}}$ equal the coefficient and standard error of $\ln K$ in the reparameterized regression.

## Testing with `linearHypothesis()` in R

- The `car` package provides `linearHypothesis()` for testing linear restrictions on regression coefficients.

- Testing for constant returns to scale ($\beta_{1}+\beta_{2}=1$):

  ```{r}
  #| echo: true
  #| message: false
  library(car)
  linearHypothesis(fit, "lnL + lnK = 1")
  ```

- `linearHypothesis()` reports an F-statistic. If $T \sim t_{n-k-1}$, then $F = T^{2} \sim F_{1, n-k-1}$.

- For a single restriction, the F-test and the two-sided t-test are equivalent: $F = T^{2}$, with identical p-values.

- Testing for equal effects ($\beta_{1}=\beta_{2}$):

  ```{r}
  #| echo: true
  linearHypothesis(fit, "lnL = lnK")
  ```
