---
title: "Lecture 2: Review of Probability"
subtitle: "Economics 326 — Introduction to Econometrics II"
author:
  - name: "Vadim Marmer, UBC"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    output-file: 326_02_review.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: katex
  pdf:
    output-file: 326_02_review.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
  revealjs:
    output-file: 326_02_review_slides.html
    date: ""
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: katex
    scrollable: true
    chalkboard: false
    self-contained: true
    transition: none
---

## Randomness

::: {.hidden}
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
\gdef\nfrac#1#2{{}^{#1}\!/\!{}_{#2}}
:::

- **Random experiment**: an experiment the outcome of which cannot be predicted with certainty, even if the experiment is repeated under the same conditions.

- **Event**: a collection of outcomes of a random experiment.

- **Probability**: a function from events to $[0,1]$ interval.
  - If $\Omega$ is a collection of all possible outcomes, $P(\Omega) = 1$.
  - If $A$ is an event, $P(A) \geq 0$.
  - If $A_1, A_2, \ldots$ is a sequence of *disjoint* events, $P(A_1 \text{ or } A_2 \text{ or } \ldots) = P(A_1) + P(A_2) + \ldots$.

## Random variable

- **Random variable**: a numerical representation of a random experiment.

- **Coin-flipping** example:

  | Outcome | $X$ | $Y$ | $Z$ |
  |---------|-----|-----|-----|
  | Heads   | 0   | 1   | -1  |
  | Tails   | 1   | 0   | 1   |

- **Rolling a dice**

  | Outcome | $X$ | $Y$ |
  |---------|-----|-----|
  | 1       | 1   | 0   |
  | 2       | 2   | 1   |
  | 3       | 3   | 0   |
  | 4       | 4   | 1   |
  | 5       | 5   | 0   |
  | 6       | 6   | 1   |

## Summation operator

- Let $\{x_i: i = 1, \ldots, n\}$ be a sequence of numbers.
$$
\sum_{i=1}^{n} x_i = x_1 + x_2 + \ldots + x_n.
$$

- For a constant $c$:
$$
\sum_{i=1}^{n} c = nc.
$$
$$
\sum_{i=1}^{n} cx_i = c(x_1 + x_2 + \ldots + x_n) = c\sum_{i=1}^{n} x_i.
$$

## Summation operator (continued)

- Let $\{y_i: i = 1, \ldots, n\}$ be another sequence of numbers, and $a, b$ be two constants:
$$
\sum_{i=1}^{n}(ax_i + by_i) = a\sum_{i=1}^{n} x_i + b\sum_{i=1}^{n} y_i.
$$

- But:
$$
\sum_{i=1}^{n} x_i y_i \neq \sum_{i=1}^{n} x_i \sum_{i=1}^{n} y_i.
$$
$$
\sum_{i=1}^{n} \frac{x_i}{y_i} \neq \frac{\sum_{i=1}^{n} x_i}{\sum_{i=1}^{n} y_i}.
$$
$$
\sum_{i=1}^{n} x_i^2 \neq \left(\sum_{i=1}^{n} x_i\right)^2.
$$

## Discrete random variables

We often distinguish between **discrete** and **continuous** random variables.

- A **discrete** random variable takes on only a **finite or countably infinite** number of values.

- The **distribution** of a discrete random variable is a list of all possible values and the probability that each value would occur:

  | **Value** | $x_1$ | $x_2$ | $\ldots$ | $x_n$ |
  |-----------|-------|-------|----------|-------|
  | **Probability** | $p_1$ | $p_2$ | $\ldots$ | $p_n$ |

  Here $p_i$ denotes the probability of a random variable $X$ taking on value $x_i$:
  $$
  p_i = P(X = x_i) \text{ (Probability Mass Function (PMF)).}
  $$
  Each $p_i$ is between 0 and 1, and $\sum_{i=1}^{n} p_i = 1$.

## Example: Bernoulli distribution

- Consider a single trial with two outcomes: "success" (with probability $p$) or "failure" (with probability $1-p$).

- Define the random variable:
  $$
  X = \begin{cases} 1 & \text{if success} \\ 0 & \text{if failure} \end{cases}
  $$

- Then $X$ follows a **Bernoulli distribution**: $X \sim Bernoulli(p)$.

- PMF:
  $$
  P(X = x) = p^x (1-p)^{1-x}, \quad x \in \{0, 1\}.
  $$

## Discrete random variables (continued)

- Indicator function:
$$
\mathbf{1}(x_i \leq x) = \begin{cases} 1 & \text{if } x_i \leq x \\ 0 & \text{if } x_i > x \end{cases}
$$

- **Cumulative Distribution Function (CDF)**:
$$
F(x) = P(X \leq x) = \sum_i p_i \mathbf{1}(x_i \leq x).
$$
- $F(x)$ is non-decreasing.
- For discrete random variables, the CDF is a step function.

## Example: CDF of Bernoulli(0.3)

```{r}
#| echo: false
#| fig-width: 4
#| fig-height: 2.5
#| fig-align: center

p <- 0.3
par(mar = c(4, 4, 1, 1))

# Plot the step function
plot(NA, xlim = c(-0.5, 1.5), ylim = c(0, 1.1),
     xlab = "x", ylab = "F(x)", xaxt = "n")
axis(1, at = c(0, 1))

# F(x) = 0 for x < 0
segments(-0.5, 0, 0, 0, lwd = 2, col = "blue")
# F(x) = 1-p for 0 <= x < 1
segments(0, 1-p, 1, 1-p, lwd = 2, col = "blue")
# F(x) = 1 for x >= 1
segments(1, 1, 1.5, 1, lwd = 2, col = "blue")

# Closed circles at jump points (right-continuous)
points(c(0, 1), c(1-p, 1), pch = 19, col = "blue")
# Open circles at jump points
points(c(0, 1), c(0, 1-p), pch = 1, col = "blue")
```

$$
F(x) = \begin{cases} 0 & \text{if } x < 0 \\ 1-p & \text{if } 0 \leq x < 1 \\ 1 & \text{if } x \geq 1 \end{cases}
$$

## Continuous random variable

- A random variable is continuously distributed if the range of possible values it can take is uncountably infinite (for example, a real line).

- A continuous random variable takes on any real value with **zero** probability.

- For continuous random variables, the CDF is continuous and differentiable.

- The derivative of the CDF is called the **Probability Density Function (PDF)**:
$$
f(x) = \frac{dF(x)}{dx} \text{ and } F(x) = \int_{-\infty}^{x} f(u) du;
$$
$$
\int_{-\infty}^{\infty} f(x) dx = 1.
$$
- Since $F(x)$ is non-decreasing, $f(x) \geq 0$ for all $x$.

## Example: Uniform distribution

A random variable $X$ follows a **Uniform distribution** on $[0, 1]$, written $X \sim Uniform(0, 1)$, if it is equally likely to take any value in $[0, 1]$.

:::: {.columns}
::: {.column width="50%"}
**PDF:**
$$
f(x) = \begin{cases} 0 & \text{if } x < 0 \\ 1 & \text{if } 0 \leq x \leq 1 \\ 0 & \text{if } x > 1 \end{cases}
$$
:::

::: {.column width="50%"}
**CDF:**
$$
F(x) = \begin{cases} 0 & \text{if } x < 0 \\ x & \text{if } 0 \leq x \leq 1 \\ 1 & \text{if } x > 1 \end{cases}
$$
:::
::::

:::: {.columns}
::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 4
#| fig-height: 2.5
#| fig-align: center

par(mar = c(4, 4, 1, 1))
x <- seq(-0.5, 1.5, length.out = 1000)
pdf_vals <- dunif(x, min = 0, max = 1)
plot(x, pdf_vals, type = "l", lwd = 2, col = "blue",
     xlab = "x", ylab = "f(x)",
     ylim = c(0, 1.2), xaxt = "n")
axis(1, at = c(0, 0.5, 1))
segments(-0.5, 0, 0, 0, lwd = 2, col = "blue")
segments(1, 0, 1.5, 0, lwd = 2, col = "blue")
points(c(0, 1), c(1, 1), pch = 19, col = "blue")
points(c(0, 1), c(0, 0), pch = 1, col = "blue")
```
:::

::: {.column width="50%"}
```{r}
#| echo: false
#| fig-width: 4
#| fig-height: 2.5
#| fig-align: center

par(mar = c(4, 4, 1, 1))
x <- seq(-0.5, 1.5, length.out = 1000)
cdf_vals <- punif(x, min = 0, max = 1)
plot(x, cdf_vals, type = "l", lwd = 2, col = "red",
     xlab = "x", ylab = "F(x)",
     ylim = c(0, 1.1), xaxt = "n")
axis(1, at = c(0, 0.5, 1))
```
:::
::::

## Joint distribution (discrete)

- Two random variables $X, Y$

  |       | $y_1$    | $y_2$    | $\cdots$ | $y_m$    | Marginal                  |
  |-------|----------|----------|----------|----------|---------------------------|
  | $x_1$ | $p_{11}$ | $p_{12}$ | $\cdots$ | $p_{1m}$ | $p_1^X=\sum_{j=1}^mp_{1j}$ |
  | $x_2$ | $p_{21}$ | $p_{22}$ | $\cdots$ | $p_{2m}$ | $p_2^X=\sum_{j=1}^mp_{2j}$ |
  | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$ | $\vdots$               |
  | $x_n$ | $p_{n1}$ | $p_{n2}$ | $\cdots$ | $p_{nm}$ | $p_n^X=\sum_{j=1}^mp_{nj}$ |

  **Joint PMF**: $p_{ij} = P(X = x_i, Y = y_j)$.

  **Marginal PMF**: $p_i^X = P(X = x_i) = \sum_{j=1}^{m} p_{ij}$.

- **Conditional Distribution**: If $P(X = x_1) \neq 0$,
$$
p_j^{Y|X=x_1} = P(Y = y_j | X = x_1) = \frac{P(Y = y_j, X = x_1)}{P(X = x_1)} = \frac{p_{1,j}}{p_1^X}
$$

## Joint distribution (continuous)

- Joint PDF: $f_{X,Y}(x, y)$ and $\int_{-\infty}^{\infty} \int_{-\infty}^{\infty} f_{X,Y}(x, y) dx dy = 1$.

- Marginal PDF: $f_X(x) = \int_{-\infty}^{\infty} f_{X,Y}(x, y) dy$.

- Conditional PDF: $f_{Y|X=x}(y|x) = f_{X,Y}(x, y) / f_X(x)$.

## Independence

- Two (discrete) random variables are independent if **for all** $x, y$:
$$
P(X = x, Y = y) = P(X = x) P(Y = y).
$$

- If independent:
$$
P(Y = y | X = x) = \frac{P(X = x, Y = y)}{P(X = x)} = P(Y = y).
$$

- Two continuous random variables are independent if **for all** $x, y$:
$$
f_{X,Y}(x, y) = f_X(x) f_Y(y).
$$

- If independent, $f_{Y|X}(y|x) = f_Y(y)$ for all $x$.

## Expected value

- Let $g$ be some function:
  $$
  \E{g(X)} = \sum_i g(x_i) p_i \text{ (discrete).}
  $$
  $$
  \E{g(X)} = \int g(x) f(x) dx \text{ (continuous).}
  $$
  Expectation is a transformation of a distribution (PMF or PDF) and is a **constant**!

- **Mean** (center of a distribution):
  $$
  \E{X} = \sum_i x_i p_i \text{ or } \E{X} = \int x f(x) dx.
  $$

- **Variance** (spread of a distribution): $\Var{X} = \E{(X - \E{X})^2}$
  $$
  \Var{X} = \sum_i (x_i - \E{X})^2 p_i \text{ or } \Var{X} = \int (x - \E{X})^2 f(x) dx.
  $$

- **Standard deviation**: $\sqrt{\Var{X}}$.



## Properties

- Linearity:
  - Discrete:
\begin{align*}
\E{a + bX} &= \sum_i (a + bx_i) p_i = a \sum_i p_i + b \sum_i x_i p_i \\
&= a + b\E{X}.
\end{align*}
  - Continuous:
\begin{align*}
\E{a + bX} &= \int (a + bx) f(x) dx = a\int f(x) dx + b\int x f(x) dx \\
&= a + b\E{X}.
\end{align*}


- Re-centering: a random variable $X - \E{X}$ has mean zero:
$$
\E{X - \E{X}} = \E{X} - \E{\E{X}} = \E{X} - \E{X} = 0.
$$



## Properties (continued)

- Variance formula: $\Var{X} = \E{X^2} - (\E{X})^2$
  \begin{align*}
  \Var{X} &= \E{(X - \E{X})^2} \\
  &= \E{(X - \E{X})(X - \E{X})} \\
  &= \E{(X - \E{X})X - (X - \E{X}) \cdot \E{X}} \\
  &= \E{(X - \E{X})X} - \E{(X - \E{X}) \cdot \E{X}} \\
  &= \E{X^2 - X \cdot \E{X}} - \E{X} \cdot \E{X - \E{X}} \\
  &= \E{X^2} - \E{X} \cdot \E{X} - \E{X} \cdot 0\\
  & = \E{X^2} - (\E{X})^2
  \end{align*}

- If $\E{X} = 0$ then $\Var{X} = \E{X^2}$.

## Properties (continued)

- If $c$ is a constant, $\E{c} = c$, and
$$
\Var{c} = \E{(c - \E{c})^2} = (c - c)^2 = 0.
$$


- $\Var{a + bX} = b^2 \Var{X}$
  \begin{align*}
  \Var{a + bX} &= \E{((a + bX) - \E{a + bX})^2}\\
  & = \E{(a + bX - a - b\E{X})^2} \\
  &= \E{(bX - b\E{X})^2} = \E{b^2(X - \E{X})^2} \\
  &= b^2 \E{(X - \E{X})^2} \\
  &= b^2 \Var{X}.
  \end{align*}

- Re-scaling: Let $\Var{X} = \sigma^2$, so the standard deviation is $\sigma$:
$$
\Var{\frac{X}{\sigma}} = \frac{1}{\sigma^2} \Var{X} = 1.
$$

## Example: Bernoulli distribution (continued)

- Recall: $X \sim Bernoulli(p)$ takes values $\{0, 1\}$ with $P(X=1) = p$ and $P(X=0) = 1-p$.

- Mean:
  $$
  \E{X} = 0 \cdot (1-p) + 1 \cdot p = p.
  $$

- Variance:
  $$
  \E{X^2} = 0^2 \cdot (1-p) + 1^2 \cdot p = p.
  $$
  $$
  \Var{X} = \E{X^2} - (\E{X})^2 = p - p^2 = p(1-p).
  $$

## Example: Uniform distribution (continued)

- Recall: $X \sim Uniform(0, 1)$ has PDF $f(x) = 1$ for $x \in [0, 1]$.

- Mean:
  $$
  \E{X} = \int_0^1 x \cdot 1 \, dx = \left. \frac{x^2}{2} \right|_0^1 = \frac{1}{2}.
  $$

- Variance:
  $$
  \E{X^2} = \int_0^1 x^2 \cdot 1 \, dx = \left. \frac{x^3}{3} \right|_0^1 = \frac{1}{3}.
  $$
  $$
  \Var{X} = \E{X^2} - (\E{X})^2 = \frac{1}{3} - \left(\frac{1}{2}\right)^2 = \frac{1}{3} - \frac{1}{4} = \frac{1}{12}.
  $$

## Covariance

- **Covariance**: Let $X, Y$ be two random variables.
$$
\Cov{X, Y} = \E{(X - \E{X})(Y - \E{Y})}.
$$
$$
\Cov{X, Y} = \sum_i \sum_j (x_i - \E{X})(y_j - \E{Y}) \cdot P(X = x_i, Y = y_j).
$$
$$
\Cov{X, Y} = \int \int (x - \E{X})(y - \E{Y}) f_{X,Y}(x, y) dx dy.
$$

- $\Cov{X, Y} = \E{XY} - \E{X} \E{Y}$.
  $$
  \Cov{X, Y} = \E{(X - \E{X})(Y - \E{Y})} = \E{XY} - \E{X} \cdot \E{Y}.
  $$

## Properties of covariance

- $\Cov{X, c} = 0$.

- $\Cov{X, X} = \Var{X}$.

- $\Cov{X, Y} = \Cov{Y, X}$.

- $\Cov{X, Y + Z} = \Cov{X, Y} + \Cov{X, Z}$.

- $\Cov{a_1 + b_1 X, a_2 + b_2 Y} = b_1 b_2 \Cov{X, Y}$.

- If $X$ and $Y$ are independent then $\Cov{X, Y} = 0$.

- $\Var{X \pm Y} = \Var{X} + \Var{Y} \pm 2\Cov{X, Y}$.

## Correlation

- Correlation coefficient:
$$
\mathrm{Corr}(X, Y) = \frac{\Cov{X, Y}}{\sqrt{\Var{X} \Var{Y}}}.
$$

- Cauchy-Schwartz inequality: $|\Cov{X, Y}| \leq \sqrt{\Var{X} \Var{Y}}$ and therefore
$$
-1 \leq \mathrm{Corr}(X, Y) \leq 1.
$$

- $\mathrm{Corr}(X, Y) = \pm 1$ $\Leftrightarrow$ $Y = a + bX$.

## Conditional expectation

- Suppose you know that $X = x$. You can update your expectation of $Y$ by **conditional expectation**:
$$
\E{Y | X = x} = \sum_i y_i P(Y = y_i | X = x) \text{ (discrete)}
$$
$$
\E{Y | X = x} = \int y f_{Y|X}(y|x) dy \text{ (continuous).}
$$

- $\E{Y | X = x}$ is a constant.

- $\E{Y | X}$ is a function of $X$ and is a **random variable** and a function of $X$ (Uncertainty about $X$ has not been realized yet):
$$
\E{Y | X} = \sum_i y_i P(Y = y_i | X) = g(X)
$$
$$
\E{Y | X} = \int y f_{Y|X}(y|X) dy = g(X),
$$
for some function $g$ that depends on PMF (PDF).

## Properties of conditional expectation

- Conditional expectation satisfies all properties of unconditional expectation.

- Once you condition on $X$, you can treat any function of $X$ as a constant:
$$
\E{h_1(X) + h_2(X) Y | X} = h_1(X) + h_2(X) \E{Y | X},
$$
for any functions $h_1$ and $h_2$.

- **Law of Iterated Expectation (LIE)**:
$$
\E{\E{Y | X}} = \E{Y},
$$
$$
\E{\E{Y | X, Z} | X} = \E{Y | X}.
$$

- Conditional variance:
$$
\Var{Y | X} = \E{(Y - \E{Y | X})^2 | X}.
$$

- **Mean independence**:
$$
\E{Y | X} = \E{Y} = \text{constant.}
$$

## Example: conditional expectation and LIE

::: {.nonincremental}
- Joint PMF of $X \in \{0, 1\}$ and $Y \in \{1, 2, 3\}$:
:::

```{=latex}
\renewcommand{\arraystretch}{1.4}
```

|  | $Y=1$ | $Y=2$ | $Y=3$ | $P(X=x)$ |
|---|---|---|---|---|
| $X=0$ | $\nfrac{1}{10}$ | $\nfrac{2}{10}$ | $\nfrac{2}{10}$ | $\nfrac{1}{2}$ |
| $X=1$ | $\nfrac{2}{10}$ | $\nfrac{2}{10}$ | $\nfrac{1}{10}$ | $\nfrac{1}{2}$ |
| $P(Y=y)$ | $\nfrac{3}{10}$ | $\nfrac{4}{10}$ | $\nfrac{3}{10}$ |  |

```{=latex}
\renewcommand{\arraystretch}{1.0}
```

- Conditional PMFs (divide joint probabilities by marginal of $X$):
$$
P(Y=y | X=0): \quad \tfrac{1/10}{1/2},\ \tfrac{2/10}{1/2},\ \tfrac{2/10}{1/2} = \tfrac{1}{5},\ \tfrac{2}{5},\ \tfrac{2}{5}
$$
$$
P(Y=y | X=1): \quad \tfrac{2/10}{1/2},\ \tfrac{2/10}{1/2},\ \tfrac{1/10}{1/2} = \tfrac{2}{5},\ \tfrac{2}{5},\ \tfrac{1}{5}
$$

- $X$ and $Y$ are not independent since the conditional PMFs of $Y$ given $X=x$ are different for $x=0$ and $x=1$.

- Conditional expectations: $\E{Y \mid X=x}=\sum_{j=1}^{3} y_{j} P\left(Y=y_{j} \mid X=x\right)$
$$
\E{Y | X=0} = 1 \cdot \tfrac{1}{5} + 2 \cdot \tfrac{2}{5} + 3 \cdot \tfrac{2}{5} = \tfrac{1+4+6}{5} = \tfrac{11}{5}
$$
$$
\E{Y | X=1} = 1 \cdot \tfrac{2}{5} + 2 \cdot \tfrac{2}{5} + 3 \cdot \tfrac{1}{5} = \tfrac{2+4+3}{5} = \tfrac{9}{5}
$$

- LIE check: $\E{\E{Y|X}} = \frac{11}{5} \cdot \frac{1}{2} + \frac{9}{5} \cdot \frac{1}{2} = \frac{11}{10} + \frac{9}{10} = 2$
- Direct: $\E{Y} = 1 \cdot \frac{3}{10} + 2 \cdot \frac{4}{10} + 3 \cdot \frac{3}{10} = \frac{3+8+9}{10} = 2$ ✓

## Proof of the Law of Iterated Expectations

- Let $X$ take values $x_1, \ldots, x_n$ and $Y$ take values $y_1, \ldots, y_m$.

::: {.fragment}
$\displaystyle \E{\E{Y | X}} = \sum_{i=1}^{n} \E{Y | X = x_i}\, P(X = x_i)$
:::

::: {.fragment}
$\displaystyle = \sum_{i=1}^{n} \left(\sum_{j=1}^{m} y_j\, P(Y = y_j | X = x_i)\right) P(X = x_i)$
:::

::: {.fragment}
$\displaystyle = \sum_{i=1}^{n} \sum_{j=1}^{m} y_j\, P(Y = y_j | X = x_i)\, P(X = x_i)$
:::

::: {.fragment}
$\displaystyle = \sum_{j=1}^{m} y_j \sum_{i=1}^{n} P(Y = y_j,\, X = x_i)$
:::

::: {.fragment}
$\displaystyle = \sum_{j=1}^{m} y_j\, P(Y = y_j) = \E{Y}$
:::

- Key steps: 
  - $P(Y = y_j | X = x_i) \cdot P(X = x_i) = P(Y = y_j,\, X = x_i)$.
  - Changing the order of summation.
  - Summing over $i$ gives marginal $P(Y = y_j)$.
  
## Relationship between different concepts of independence

$$
\begin{array}{c}
X \text{ and } Y \text{ are independent} \\
\Downarrow \\
\E{Y | X} = \text{constant (mean independence)} \\
\Downarrow \\
\Cov{X, Y} = 0 \text{ (uncorrelatedness)}
\end{array}
$$

## Normal distribution

- A normal rv is a continuous rv that can take on any value. The PDF of a normal rv $X$ is
$$
f(x) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{(x - \mu)^2}{2\sigma^2}\right), \text{ where}
$$
$$
\mu = \E{X} \text{ and } \sigma^2 = \Var{X}.
$$
We usually write $X \sim N(\mu, \sigma^2)$.

- If $X \sim N(\mu, \sigma^2)$, then $a + bX \sim N(a + b\mu, b^2\sigma^2)$.

## Standard Normal distribution

- Standard Normal rv has $\mu = 0$ and $\sigma^2 = 1$. Its PDF is $\phi(z) = \frac{1}{\sqrt{2\pi}} \exp\left(-\frac{z^2}{2}\right)$.

- Symmetric around zero (mean): if $Z \sim N(0, 1)$, $P(Z > z) = P(Z < -z)$.

- Thin tails: $P(-1.96 \leq Z \leq 1.96) = 0.95$.

- If $X \sim N(\mu, \sigma^2)$, then $(X - \mu)/\sigma \sim N(0, 1)$.

## Bivariate Normal distribution

- $X$ and $Y$ have a bivariate normal distribution if their joint PDF is given by:
  $$
  f(x, y) = \frac{1}{2\pi\sqrt{(1-\rho^2) \sigma_X^2 \sigma_Y^2}} \exp\left[-\frac{Q}{2(1-\rho^2)}\right],
  $$
  where $Q = \frac{(x-\mu_X)^2}{\sigma_X^2} + \frac{(y-\mu_Y)^2}{\sigma_Y^2} - 2\rho\frac{(x-\mu_X)(y-\mu_Y)}{\sigma_X\sigma_Y}$,

  $\mu_X = \E{X}, \mu_Y = \E{Y}, \sigma_X^2 = \Var{X}, \sigma_Y^2 = \Var{Y}$, and $\rho = \mathrm{Corr}(X, Y)$.

## Properties of Bivariate Normal distribution

If $X$ and $Y$ have a bivariate normal distribution:

- $a + bX + cY \sim N(\mu^*, (\sigma^*)^2)$, where
  $$
  \mu^* = a + b\mu_X + c\mu_Y, \quad (\sigma^*)^2 = b^2\sigma_X^2 + c^2\sigma_Y^2 + 2bc\rho\sigma_X\sigma_Y.
  $$


- $\Cov{X, Y} = 0 \Longrightarrow X$ and $Y$ are independent.

- $\E{Y | X} = \mu_Y + \frac{\Cov{X, Y}}{\sigma_X^2}(X - \mu_X)$.

- Can be generalized to more than 2 variables (multivariate normal).

## Appendix: The Cauchy-Schwartz Inequality

- Claim: $|\Cov{X, Y}| \leq \sqrt{\Var{X} \Var{Y}}$.

- **Proof**: Define
  $$
  U = Y - \beta X,
  $$
  where
  $$
  \beta = \frac{\Cov{X, Y}}{\Var{X}},
  $$
  - Note that $\beta$ is a constant!
  - Also note the connection to regression and OLS in the definition of $\beta$.

::: {.content-hidden when-format="revealjs"}
- Since variances are always non-negative:

  \begin{alignat*}{2}
  0 & \leq \Var{U}  &&\\
    & = \Var{Y - \beta X}  &&\quad (\text{def. of } U)\\
    & = \Var{Y} + \Var{\beta X} - 2\Cov{Y, \beta X}  &&\quad (\text{prop. of var.})\\
    & = \Var{Y} + \beta^2 \Var{X} - 2\beta \Cov{X, Y}  &&\quad (\text{prop. of var., cov.})\\
    & = \Var{Y} + \underbrace{\left(\frac{\Cov{X, Y}}{\Var{X}}\right)^{\!2}}_{=\beta^2} \!\Var{X} &&\\
    & \phantom{=\Var{Y}} - 2 \underbrace{\left(\frac{\Cov{X, Y}}{\Var{X}} \right)}_{=\beta}\Cov{X, Y}  &&\quad (\text{def. of } \beta)\\
    & = \Var{Y} + \frac{\Cov{X, Y}^2}{\Var{X}} - 2 \frac{\Cov{X, Y}^2}{\Var{X}}   &&\\
    & = \Var{Y} - \frac{\Cov{X, Y}^2}{\Var{X}}. &&
  \end{alignat*}
:::

::: {.content-visible when-format="revealjs"}
- Since variances are always non-negative:

  $0 \leq \Var{U}$

  [$\quad= \Var{Y - \beta X} \quad (\text{def. of } U)$]{.fragment}

  [$\quad= \Var{Y} + \Var{\beta X} - 2\Cov{Y, \beta X} \quad (\text{prop. of var.})$]{.fragment}

  [$\quad= \Var{Y} + \beta^2 \Var{X} - 2\beta \Cov{X, Y} \quad (\text{prop. of var., cov.})$]{.fragment}

  [$\quad= \Var{Y} + \underbrace{\left(\frac{\Cov{X, Y}}{\Var{X}}\right)^{\!2}}_{=\beta^2} \!\Var{X}$]{.fragment}

  [$\quad\phantom{=\Var{Y}} - 2 \underbrace{\left(\frac{\Cov{X, Y}}{\Var{X}} \right)}_{=\beta}\Cov{X, Y} \quad (\text{def. of } \beta)$]{.fragment}

  [$\quad= \Var{Y} + \frac{\Cov{X, Y}^2}{\Var{X}} - 2 \frac{\Cov{X, Y}^2}{\Var{X}}$]{.fragment}

  [$\quad= \Var{Y} - \frac{\Cov{X, Y}^2}{\Var{X}}.$]{.fragment}
:::
- Rearranging:
  $$
  \Cov{X, Y}^2 \leq \Var{X} \Var{Y}
  $$
- or
  $$
  |\Cov{X, Y}| \leq \sqrt{\Var{X} \Var{Y}}.
  $$
