---
title: "Lecture 20: Two-Stage Least Squares"
subtitle: "Economics 326 — Introduction to Econometrics II"
author:
  - name: "Vadim Marmer, UBC"
date: today
date-format: "MMMM D, YYYY"
format:
  html:
    output-file: 326_20_2SLS.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: katex
    embed-resources: true
  pdf:
    output-file: 326_20_2SLS.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
  revealjs:
    output-file: 326_20_2SLS_slides.html
    date: ""
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: katex
    scrollable: true
    chalkboard: false
    self-contained: true
    transition: none
---

## Beyond the simple IV model

::: {.hidden}
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
\gdef\Vhat#1{\widehat{\mathrm{Var}}\left(#1\right)}
\gdef\se#1{\mathrm{se}\left(#1\right)}
:::

- Previously: simple IV model with one endogenous regressor, one instrument, no controls.

- The simple IV formula does not extend to models with:

  1. Exogenous control variables.

  2. Multiple instruments ($l > 1$).

- **Two-stage least squares (2SLS)** handles both extensions.

- **Notation:** $y$ = dependent variable, $Y$ = endogenous regressor, $X$'s = exogenous controls, $Z$'s = instruments.

## IV model with exogenous controls

- The model:
  $$
  y_i = \gamma_0 + {\color{red}\beta_1 Y_i} + {\color{blue}\gamma_1 X_{1,i} + \ldots + \gamma_k X_{k,i}} + U_i,
  $$
  where

  - $\gamma_0$ is the intercept: $\E{U_i} = 0$.

  - ${\color{red}Y_i}$ is the **endogenous** regressor: $\Cov{Y_i, U_i} \neq 0$.

  - ${\color{blue}X_{1,i}, \ldots, X_{k,i}}$ are $k$ **exogenous** regressors:
    $$
    \Cov{X_{1,i}, U_i} = \ldots = \Cov{X_{k,i}, U_i} = 0.
    $$

- $l$ instruments ${\color{blue}Z_{1,i}, \ldots, Z_{l,i}}$, excluded from the regression.

## Identification with instruments

- There are ${\color{red}k + 2}$ unknown coefficients:
  $$
  y_i = {\color{red}\gamma_0} + {\color{red}\beta_1} Y_i + {\color{red}\gamma_1} X_{1,i} + \ldots + {\color{red}\gamma_k} X_{k,i} + U_i.
  $$

- Exogeneity gives only ${\color{blue}k + 1}$ equations:
  $$
  \E{U_i} = 0, \quad \Cov{X_{j,i}, U_i} = 0, \quad j = 1, \ldots, k.
  $$

- $l$ instruments provide $l$ additional moment conditions:
  $$
  \Cov{Z_{j,i}, U_i} = 0, \quad j = 1, \ldots, l,
  $$
  for a total of ${\color{blue}k + 1 + l}$ equations.

- **Order condition:** $l \geq 1$.

  - $l = 1$: exactly identified.
  - $l > 1$: overidentified.

## The first-stage equation

- Consider a system of two equations. The original regression becomes the **second stage**; a new **first-stage** equation describes how $Y_i$ depends on $Z$'s and $X$'s:
  $$
  \begin{aligned}
  \text{(first stage)} \quad Y_i &= \pi_0 + \pi_1 Z_{1,i} + \ldots + \pi_l Z_{l,i} \\
  &\quad + \pi_{l+1} X_{1,i} + \ldots + \pi_{l+k} X_{k,i} + V_i \\[4pt]
  \text{(second stage)} \quad y_i &= \gamma_0 + \beta_1 Y_i + \gamma_1 X_{1,i} + \ldots + \gamma_k X_{k,i} + U_i
  \end{aligned}
  $$

- All RHS variables in the first-stage equation are exogenous → the $\pi$'s can be estimated consistently by OLS.

- $Y_i$ is endogenous because $\Cov{U_i, V_i} \neq 0$.

- **IV relevance condition:** at least one $\pi_j \neq 0$ for $j = 1, \ldots, l$.

- $Y_i$ contains both exogenous variation (driven by $Z$'s and $X$'s) and endogenous variation (correlated with $U_i$).

- OLS uses all variation in $Y_i$ → **inconsistent**.

- **Idea:** estimate $\beta_1$ using only the **exogenous variation** in $Y_i$.

- The first stage extracts this variation: $\hat{Y}_i$ captures the part of $Y_i$ explained by $Z$'s and $X$'s.

- $X$'s appear in the first stage because they affect $Y_i$ directly and can be correlated with $Z$'s.

- $\hat{Y}_i$ depends only on exogenous variables → uncorrelated with $U_i$.

## 2SLS: two stages

- **Stage 1:** Regress $Y_i$ on $Z_{1,i}, \ldots, Z_{l,i}, X_{1,i}, \ldots, X_{k,i}$ by OLS. Obtain fitted values:
  $$
  \begin{aligned}
  \hat{Y}_i &= \hat{\pi}_0 + \hat{\pi}_1 Z_{1,i} + \ldots + \hat{\pi}_l Z_{l,i} \\
  &\quad + \hat{\pi}_{l+1} X_{1,i} + \ldots + \hat{\pi}_{l+k} X_{k,i}.
  \end{aligned}
  $$

- **Stage 2:** Regress $y_i$ on $\hat{Y}_i$ and $X_{1,i}, \ldots, X_{k,i}$ by OLS:
  $$
  \begin{aligned}
  y_i &= \hat{\gamma}_0^{2SLS} + \hat{\beta}_1^{2SLS} \hat{Y}_i \\
  &\quad + \hat{\gamma}_1^{2SLS} X_{1,i} + \ldots + \hat{\gamma}_k^{2SLS} X_{k,i} + \hat{U}_i.
  \end{aligned}
  $$

- The 2SLS estimators are **consistent** and **asymptotically normal**.

- Standard errors from naïve second-stage OLS are **incorrect**; statistical packages report corrected standard errors.

## 2SLS with a single instrument

- Special case: $l = 1$ (one instrument $Z_i$), $k = 0$ (no exogenous controls).

- Stage 1: regress $Y_i$ on $Z_i$ → $\hat{Y}_i = \hat{\alpha}_0 + \hat{\alpha}_1 Z_i$.

- Stage 2: regress $y_i$ on $\hat{Y}_i$.

- The 2SLS estimator coincides with the simple IV estimator from the previous lecture.

- With exogenous controls or multiple instruments, the simple IV formula no longer works; 2SLS is the standard approach.

## Example: returns to education (setup)

- Estimate the returns to education using the MROZ dataset (Wooldridge, 2006):
  $$
  \begin{aligned}
  \ln \text{Wage}_i &= \gamma_0 + {\color{red}\beta_1 \text{Educ}_i} \\
  &\quad + \gamma_1 \text{Exper}_i + \gamma_2 \text{Exper}_i^2 + U_i.
  \end{aligned}
  $$

  - **Endogenous**: ${\color{red}\text{Educ}}$ (education).
  - **Instruments**: $\text{MotherEduc}$ and $\text{FatherEduc}$ (parents' education).
  - **Exogenous**: $\text{Exper}$ and $\text{Exper}^2$ (experience).

- The model is **overidentified** ($l = 2 > 1$).

- In R, use `ivreg()` from the `AER` package with `vcovHC(..., type = "HC1")` for heteroskedasticity-robust standard errors.

```{r}
#| echo: false
#| message: false
options(scipen = 999)
library(wooldridge)
library(AER)
library(lmtest)
library(sandwich)
data(mroz)
d <- subset(mroz, inlf == 1)
d$expersq <- d$exper^2
```

## Example: first stage

**First-stage regression** ($\text{Educ}$ on instruments and exogenous regressors):

```{r}
first_stage <- lm(educ ~ exper + expersq + motheduc + fatheduc, data = d)
coeftest(first_stage, vcov = vcovHC(first_stage, type = "HC1"))
```

Both instruments (`motheduc`, `fatheduc`) are statistically significant. First-stage $R^2 =$ `r round(summary(first_stage)$r.squared, 4)`.

## Example: 2SLS vs OLS

- **2SLS second stage:**

  ```{r}
  iv_fit <- ivreg(lwage ~ educ + exper + expersq |
                    exper + expersq + motheduc + fatheduc, data = d)
  coeftest(iv_fit, vcov = vcovHC(iv_fit, type = "HC1"))
  ```

- **OLS for comparison:**

  ```{r}
  ols_fit <- lm(lwage ~ educ + exper + expersq, data = d)
  coeftest(ols_fit, vcov = vcovHC(ols_fit, type = "HC1"))
  ```

- The 2SLS estimate ($\hat{\beta}_1^{2SLS} = 0.061$) is smaller than OLS ($\hat{\beta}_1^{OLS} = 0.107$), consistent with upward ability bias.

- The 2SLS standard error ($0.033$) is larger than OLS ($0.013$): 2SLS uses only the exogenous variation in $\text{Educ}$, so estimates are noisier.

## Multiple endogenous regressors

- So far: one endogenous variable with exogenous controls and multiple instruments.

- In practice, models may have several endogenous regressors.

- Example:
  $$
  \begin{aligned}
  \ln \text{Wage}_i &= \gamma_0 + {\color{red}\beta_1 \text{Educ}_i + \beta_2 \text{Children}_i} \\
  &\quad + {\color{blue}\gamma_1 \text{Age}_i + \gamma_2 \text{Sex}_i} + U_i.
  \end{aligned}
  $$

  - ${\color{red}\text{Endogenous}}$ regressors: education and children (family size).

  - ${\color{blue}\text{Exogenous}}$ regressors: age, sex, and a constant.

## General model

- General model with $m$ endogenous regressors:
  $$
  \begin{aligned}
  y_i &= \gamma_0 + {\color{red}\beta_1 Y_{1,i} + \ldots + \beta_m Y_{m,i}} \\
  &\quad + {\color{blue}\gamma_1 X_{1,i} + \ldots + \gamma_k X_{k,i}} + U_i,
  \end{aligned}
  $$
  where

  - ${\color{red}Y_{1,i}, \ldots, Y_{m,i}}$ are $m$ **endogenous** regressors:
    $$
    \Cov{Y_{1,i}, U_i} \neq 0, \ldots, \Cov{Y_{m,i}, U_i} \neq 0.
    $$

  - ${\color{blue}X_{1,i}, \ldots, X_{k,i}}$ are $k$ **exogenous** regressors:
    $$
    \Cov{X_{1,i}, U_i} = \ldots = \Cov{X_{k,i}, U_i} = 0.
    $$

## Identification and the order condition

- There are ${\color{red}k + 1 + m}$ unknown coefficients, but exogeneity of $X$'s gives only ${\color{blue}k + 1}$ equations. We need $m$ more.

- $l$ additional exogenous IVs ${\color{blue}Z_{1,i}, \ldots, Z_{l,i}}$, excluded from the second-stage equation, provide $l$ moment conditions:
  $$
  \Cov{Z_{j,i}, U_i} = 0, \quad j = 1, \ldots, l.
  $$

- **Order condition:** ${\color{red}l \geq m}$.

  - $l = m$: exactly identified.
  - $l > m$: overidentified.
  - $l < m$: underidentified (coefficients cannot be estimated).

## First-stage equations

- The system has $m$ **first-stage** equations, one per endogenous regressor:
  $$
  \begin{aligned}
  Y_{1,i} &= \pi_{0,1} + \pi_{1,1} Z_{1,i} + \ldots + \pi_{l,1} Z_{l,i} \\
  &\quad + \pi_{l+1,1} X_{1,i} + \ldots + \pi_{l+k,1} X_{k,i} + V_{1,i}, \\
  &\;\;\vdots \\
  Y_{m,i} &= \pi_{0,m} + \pi_{1,m} Z_{1,i} + \ldots + \pi_{l,m} Z_{l,i} \\
  &\quad + \pi_{l+1,m} X_{1,i} + \ldots + \pi_{l+k,m} X_{k,i} + V_{m,i}.
  \end{aligned}
  $$

- The exogenous regressors $X$'s appear because they can be correlated with $Y$'s.

- The $X$'s and $Z$'s are **uncorrelated** with all errors $U$ and $V$'s.

## 2SLS: the first stage

- Estimate each first-stage equation by OLS. The fitted values are:
  $$
  \begin{aligned}
  \hat{Y}_{1,i} &= \hat{\pi}_{0,1} + \hat{\pi}_{1,1} Z_{1,i} + \ldots + \hat{\pi}_{l,1} Z_{l,i} \\
  &\quad + \hat{\pi}_{l+1,1} X_{1,i} + \ldots + \hat{\pi}_{l+k,1} X_{k,i}, \\
  &\;\;\vdots \\
  \hat{Y}_{m,i} &= \hat{\pi}_{0,m} + \hat{\pi}_{1,m} Z_{1,i} + \ldots + \hat{\pi}_{l,m} Z_{l,i} \\
  &\quad + \hat{\pi}_{l+1,m} X_{1,i} + \ldots + \hat{\pi}_{l+k,m} X_{k,i}.
  \end{aligned}
  $$

- The $\hat{Y}$'s are functions of $Z$'s and $X$'s (all exogenous), so they are asymptotically uncorrelated with the errors.

## 2SLS: the second stage

- In the second stage, regress (OLS) $y$ on a constant, $\hat{Y}$'s, and $X$'s:
  $$
  \begin{aligned}
  y_i &= \hat{\gamma}_0^{2SLS} + \hat{\beta}_1^{2SLS} \hat{Y}_{1,i} + \ldots + \hat{\beta}_m^{2SLS} \hat{Y}_{m,i} \\
  &\quad + \hat{\gamma}_1^{2SLS} X_{1,i} + \ldots + \hat{\gamma}_k^{2SLS} X_{k,i} + \hat{U}_i.
  \end{aligned}
  $$

- The 2SLS estimators $\hat{\beta}_1^{2SLS}, \ldots, \hat{\beta}_m^{2SLS}, \hat{\gamma}_0^{2SLS}, \ldots, \hat{\gamma}_k^{2SLS}$ are **consistent** and **asymptotically normal**.

- Standard errors from naïve second-stage OLS are **incorrect**: they do not account for the estimation error in $\hat{\pi}$'s from the first stage.

- Statistical packages report the corrected standard errors.
