---
title: "Lecture 7: Confidence intervals"
subtitle: "Economics 326 — Introduction to Econometrics II"
author:
  - name: "Vadim Marmer, UBC"
format:
  html:
    output-file: 326_07_cis.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: mathjax
  pdf:
    output-file: 326_07_cis.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
    include-in-header:
      text: |
        \newcommand{\fragment}[1]{#1}
  revealjs:
    output-file: 326_07_cis_slides.html
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: mathjax
    mathjax: 3
    scrollable: true
    chalkboard: false
    self-contained: true
    transition: none
    include-after-body: reveal_mathjax_fragments.html
---

## Point vs interval estimators

::: {.hidden}
$
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Vhat#1{\widehat{\mathrm{Var}}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
\gdef\se#1{\mathrm{se}\left(#1\right)}
$
:::

```{=html}
<span style="display:none">\(\newcommand{\fragment}[1]{\class{mjxfrag}{#1}}\)</span>
```

- Recall our model:

  1. $Y_{i}=\beta _{0}+\beta _{1}X_{i}+U_{i},\qquad i=1,\ldots ,n.$
  2. $\E{U_{i}\mid \mathbf{X}} =0$ for all $i$'s.
  3. $\E{U_{i}^{2}\mid \mathbf{X}} =\sigma ^{2}$ for all $i$'s.
  4. $\E{U_{i}U_{j}\mid \mathbf{X}} =0$ for all $i\neq j$.

- So far we have established that conditionally on $\mathbf{X}$:
  - $\E{\hat{\beta}_{1} \mid \mathbf{X}} = \beta_{1}$ (unbiasedness),
  - $\Var{\hat{\beta}_{1} \mid \mathbf{X}} = \dfrac{\sigma^{2}}{\sum_{i=1}^{n}(X_{i}-\bar{X})^{2}}$.

- If the $U_i$'s are continuously distributed, then with probability **one**, $\hat{\beta}_{1}\neq \beta _{1}$:
  $$
  \hat{\beta}_{1} = \beta_{1} + \frac{\sum_{i=1}^{n}(X_{i}-\bar{X})U_{i}}{\sum_{i=1}^{n}(X_{i}-\bar{X})^{2}}.
  $$

- An **interval estimator** is a random interval $[LB, UB]$ that contains the true parameter value with a pre-specified probability.

- To construct an interval estimator for $\beta_{1}$, we need to know the **distribution** of $\hat{\beta}_{1}$.

- This requires an additional assumption about the distribution of $U_i$'s. Let's first review the normal distribution.

## Normal distribution

- A normal rv is a continuous rv that can take on any value. The PDF of a normal rv $X$ is
  $$
  f(x) = \frac{1}{\sqrt{2\pi\sigma^2}} \exp\left(-\frac{(x - \mu)^2}{2\sigma^2}\right), \text{ where}
  $$
  $$
  \mu = \E{X} \text{ and } \sigma^2 = \Var{X}.
  $$
  We usually write $X \sim N(\mu, \sigma^2)$.

- If $X \sim N(\mu, \sigma^2)$, then $a + bX \sim N(a + b\mu, b^2\sigma^2)$.

## Standard normal distribution

- A standard normal rv has $\mu = 0$ and $\sigma^2 = 1$. Its PDF is $\phi(z) = \frac{1}{\sqrt{2\pi}} \exp\left(-\frac{z^2}{2}\right)$.

- Symmetric around zero (mean): if $Z \sim N(0, 1)$, 
  \begin{align*}
  P(Z > 0) &= P(Z < 0)=0.5,\\
  P(Z > z) &= P(Z < -z) \text{ for any } z.
  \end{align*}

- Thin tails: $P(-1.96 \leq Z \leq 1.96) = 0.95$.

- If $X \sim N(\mu, \sigma^2)$, then $(X - \mu)/\sigma \sim N(0, 1)$.

## Bivariate normal distribution

- $X$ and $Y$ have a bivariate normal distribution if their joint PDF is given by:
  $$
  f(x, y) = \frac{1}{2\pi\sqrt{(1-\rho^2) \sigma_X^2 \sigma_Y^2}} \exp\left[-\frac{Q}{2(1-\rho^2)}\right],
  $$
  where
  \begin{align*}
  Q &= \frac{(x-\mu_X)^2}{\sigma_X^2} + \frac{(y-\mu_Y)^2}{\sigma_Y^2} \\
  &\quad - 2\rho\frac{(x-\mu_X)(y-\mu_Y)}{\sigma_X\sigma_Y},
  \end{align*}
  $\mu_X = \E{X}$, $\mu_Y = \E{Y}$, $\sigma_X^2 = \Var{X}$, $\sigma_Y^2 = \Var{Y}$, and $\rho = \mathrm{Corr}(X, Y)$.

## Properties of bivariate normal

- If $X$ and $Y$ have a bivariate normal distribution, then $a + bX + cY \sim N(\mu^*, (\sigma^*)^2)$, where
  \begin{align*}
  \mu^* &= a + b\mu_X + c\mu_Y, \\
  (\sigma^*)^2 &= b^2\sigma_X^2 + c^2\sigma_Y^2 + 2bc\rho\sigma_X\sigma_Y.
  \end{align*}

- $\Cov{X, Y} = 0 \Longrightarrow X$ and $Y$ are independent.

- This can be generalized to more than 2 variables (multivariate normal).

## Normality of the OLS estimator

- **Assumption 5:** $U$'s are jointly normally distributed conditional on $\mathbf{X}$.

- Then $Y_{i}=\beta _{0}+\beta _{1}X_{i}+U_{i}$ are also jointly normally distributed conditional on $\mathbf{X}$.

- Since $\hat{\beta}_{1}=\sum_{i=1}^{n}w_{i}Y_{i}$, where $w_{i}=\frac{X_{i}-\bar{X}}{\sum_{l=1}^{n}\left( X_{l}-\bar{X}\right) ^{2}}$ depend only on $\mathbf{X}$, $\hat{\beta}_{1}$ is also normally distributed conditional on $\mathbf{X}$.

- Conditionally on $\mathbf{X}$:
  \begin{align*}
  &\hat{\beta}_{1} \mid \mathbf{X} \sim N\left( \beta _{1},\Var{\hat{\beta}_{1} \mid \mathbf{X}} \right), \\
  &\Var{\hat{\beta}_{1} \mid \mathbf{X}} =\frac{\sigma ^{2}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}}.
  \end{align*}

## Interval estimation problem

- We want to construct an **interval estimator** for $\beta _{1}$:

  - The interval estimator is called a **confidence interval** (CI).

  - A CI contains the **true** value $\beta _{1}$ **with some pre-specified probability** $1-\alpha$, where $\alpha$ is a small probability of error.

  - For example, if $\alpha =0.05$, then the random CI will contain $\beta _{1}$ with probability 0.95.

- $1-\alpha$ is called the **coverage probability**.

- Confidence interval: $CI_{1-\alpha }=[LB_{1-\alpha },UB_{1-\alpha }].$ The lower bound (LB) and upper bound (UB) should depend on the coverage probability $1-\alpha.$

- The formal definition of CI: It is a **random interval** $CI_{1-\alpha}$ such that conditionally on $\mathbf{X}$,
  $$
  P\left( \beta _{1}\in CI_{1-\alpha } \mid \mathbf{X}\right) =1-\alpha .
  $$
  Note that the random element is $CI_{1-\alpha}$.

- Sometimes, a CI is defined as $P\left( \beta _{1}\in CI_{1-\alpha}\right) \geq 1-\alpha .$

## Symmetric CIs

- One approach to constructing CIs is to consider a **symmetric** interval around the estimator $\hat{\beta}_{1}$:
  $$
  CI_{1-\alpha }=\left[ \hat{\beta}_{1}-c_{1-\alpha },\hat{\beta}_{1}+c_{1-\alpha }\right]
  $$

- The problem is choosing $c_{1-\alpha }$ such that $P\left( \beta_{1}\in CI_{1-\alpha } \mid \mathbf{X}\right) =1-\alpha .$

- In choosing $c_{1-\alpha }$, we will be relying on the fact that, given our assumptions and conditionally on $\mathbf{X}$:
  \begin{align*}
  &\hat{\beta}_{1} \mid \mathbf{X} \sim N\left( \beta _{1},\Var{\hat{\beta}_{1} \mid \mathbf{X}}\right), \\
  &\Var{\hat{\beta}_{1} \mid \mathbf{X}} =\frac{\sigma ^{2}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}}.
  \end{align*}

- Note that conditionally on $\mathbf{X}$:
  $$
  \frac{\hat{\beta}_{1}-\beta _{1}}{\sqrt{\Var{\hat{\beta}_{1} \mid \mathbf{X}} }}\sim N\left( 0,1\right) .
  $$

## Standard normal quantiles

- Let $Z\sim N\left( 0,1\right) .$ The $\tau$-th **quantile** (percentile) of the standard normal distribution is $z_{\tau }$ such that
  $$
  P\left( Z\leq z_{\tau }\right) =\tau .
  $$

- **Median**: $\tau =0.5$ and $z_{0.5}=0.$ ($P\left( Z\leq 0\right) =0.5$).

- If $\tau =0.975$ then $z_{0.975}=1.96$. Due to symmetry, if $\tau =0.025$ then $z_{0.025}=-1.96.$

## $\sigma^2$ is known (infeasible CIs)

- **Suppose** (for a moment) that $\sigma ^{2}$ is known, and we can compute exactly the variance of $\hat{\beta}_{1}$:
  $$
  \Var{\hat{\beta}_{1} \mid \mathbf{X}} =\frac{\sigma ^{2}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}}.
  $$

- Consider the following CI:
  \begin{align*}
  CI_{1-\alpha } = \Big[ &\hat{\beta}_{1}-z_{1-\alpha /2}\sqrt{\Var{\hat{\beta}_{1} \mid \mathbf{X}} }, \\
  &\hat{\beta}_{1}+z_{1-\alpha /2}\sqrt{\Var{\hat{\beta}_{1} \mid \mathbf{X}} }\Big] .
  \end{align*}

- For example, if $1-\alpha =0.95 \Longleftrightarrow \alpha =0.05 \Longleftrightarrow z_{1-\alpha/2}=z_{0.975}=1.96$, and $CI_{0.95}$ is
  \begin{align*}
  \hat{\beta}_{1} \pm 1.96\sqrt{\Var{\hat{\beta}_{1} \mid \mathbf{X}} } .
  \end{align*}

## Infeasible CI validity ($\sigma^2$ known)

::: {.nonincremental}

- **Goal:** show that $P\left( \beta _{1}\in CI_{1-\alpha} \mid \mathbf{X}\right) =1-\alpha$.

- **Notation:** $\sigma_{\hat{\beta}_{1}} = \sqrt{\Var{\hat{\beta}_{1} \mid \mathbf{X}}}$.

- **Key fact:** $Z=\dfrac{\hat{\beta}_{1}-\beta _{1}}{\sigma_{\hat{\beta}_{1}}}\sim N(0,1)$ conditionally on $\mathbf{X}$.

:::

$$
\begin{aligned}
&P\left(\beta _{1} \in CI_{1-\alpha} \mid \mathbf{X}\right) \\
&\fragment{{}= P\left(\hat{\beta}_{1} - z_{1-\alpha /2}\,\sigma_{\hat{\beta}_{1}} \leq \beta _{1} \leq \hat{\beta}_{1} + z_{1-\alpha /2}\,\sigma_{\hat{\beta}_{1}} \mid \mathbf{X}\right)} \\
&\fragment{{}= P\left(-z_{1-\alpha /2}\,\sigma_{\hat{\beta}_{1}} \leq \beta _{1} - \hat{\beta}_{1} \leq z_{1-\alpha /2}\,\sigma_{\hat{\beta}_{1}} \mid \mathbf{X}\right)} \\
&\fragment{{}= P\left(-z_{1-\alpha /2}\,\sigma_{\hat{\beta}_{1}} \leq \hat{\beta}_{1} - \beta _{1} \leq z_{1-\alpha /2}\,\sigma_{\hat{\beta}_{1}} \mid \mathbf{X}\right)} \\
&\fragment{{}= P\left(-z_{1-\alpha /2} \leq \frac{\hat{\beta}_{1} - \beta _{1}}{\sigma_{\hat{\beta}_{1}}} \leq z_{1-\alpha /2} \mid \mathbf{X}\right)} \\
&\fragment{{}= P\left(-z_{1-\alpha /2} \leq Z \leq z_{1-\alpha /2} \mid \mathbf{X}\right)} \\
&\fragment{{}= P\left(Z \leq z_{1-\alpha /2} \mid \mathbf{X}\right) - P\left(Z \leq -z_{1-\alpha /2} \mid \mathbf{X}\right)} \\
&\fragment{{}= \left(1-\alpha /2\right) - \alpha /2} \\
&\fragment{{}= 1-\alpha .}
\end{aligned}
$$

## Feasible CIs ($\sigma^2$ unknown)

- Since $\sigma ^{2}$ is unknown, we must estimate it from the data:
  \begin{align*}
  s^{2} &= \frac{1}{n-2}\sum_{i=1}^{n}\hat{U}_{i}^{2} \\
  &= \frac{1}{n-2}\sum_{i=1}^{n}\left( Y_{i}-\hat{\beta}_{0}-\hat{\beta}_{1}X_{i}\right) ^{2}.
  \end{align*}

- The **standard error** of $\hat{\beta}_{1}$ is defined as
  \begin{align*}
  \se{\hat{\beta}_{1}} &= \sqrt{\Vhat{\hat{\beta}_{1}}} \\
  &= \sqrt{\frac{s^{2}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}}}.
  \end{align*}

- Replacing $\sigma$ by its estimate does not give a normal distribution anymore:
  $$
  \frac{\hat{\beta}_{1}-\beta _{1}}{\se{\hat{\beta}_{1}}}\mid \mathbf{X}\sim t_{n-2}.
  $$
  Here $t_{n-2}$ denotes the $t$-distribution with $n-2$ degrees of freedom.

- The degrees of freedom depend on

  - the sample size ($n$),

  - and the number of parameters one has to estimate to compute $s^{2}$ (two in this case, $\beta _{0}$ and $\beta _{1}$).

## Feasible CIs ($\sigma^2$ unknown)

- Let $t_{df,\tau }$ be the $\tau$-th quantile of the $t$-distribution with the number of degrees of freedom $df$: If $T\sim t_{df}$ then
  $$
  P\left( T\leq t_{df,\tau }\right) =\tau .
  $$

- Similarly to the normal distribution, the $t$-distribution is centered at zero and is symmetric around zero: $t_{n-2,1-\alpha /2}=-t_{n-2,\alpha/2}.$

- We can now construct a feasible $CI_{1-\alpha }$ as
  $$
  \hat{\beta}_{1}\pm t_{n-2,1-\alpha /2} \times \se{\hat{\beta}_{1}}.
  $$

## Example

- Data: `rental` from the `wooldridge` R package. 64 US cities in 1990.
  - `rent`: average monthly rent (\$)
  - `avginc`: per capita income (\$)

- Model: Rent$_{i}=\beta _{0}+\beta _{1}$AvgInc$_{i}+U_{i}.$
- R implementation:
  ```{r}
  # Load data and run OLS regression
  library(wooldridge)
  data("rental")
  rental90 <- subset(rental, y90 == 1)
  reg <- lm(rent ~ avginc, data = rental90)
  summary(reg)
  ```



- 95% CI for the slope coefficient:
  ```{r}
  confint(reg, "avginc", level = 0.95)
  ```


- 90% CI for the slope coefficient:
  ```{r}
  confint(reg, "avginc", level = 0.90)
  ```


## The effect of estimating $\sigma^2$

- The $t$-distribution has heavier tails than the normal.

- $t_{df,1-\alpha /2}>z_{1-\alpha /2}$, but as $df$ increases $t_{df,1-\alpha /2}\rightarrow z_{1-\alpha /2}.$

- When the sample size $n$ is large, $t_{n-2,1-\alpha /2}$ can be replaced with $z_{1-\alpha /2}.$

- In R, use `qt()` for $t$-quantiles and `qnorm()` for $z$-quantiles:
  ```{r}
  # z critical value for 95% CI
  qnorm(0.975)

  # t critical values for 95% CI with different df
  qt(0.975, df = 30)
  qt(0.975, df = 100)
  qt(0.975, df = 1000)
  qt(0.975, df = 10000)
  ```

## Interpretation of confidence intervals

- The confidence interval $CI_{1-\alpha }$ is a function of the **sample** $\left\{ \left( Y_{i},X_{i}\right) :i=1,\ldots ,n\right\}$, and therefore is **random**. This allows us to talk about the probability of $CI_{1-\alpha }$ containing the true value of $\beta _{1}.$

- Once the confidence interval is computed given the data, we have its **one realization**. The realization of $CI_{1-\alpha }$ (the computed confidence interval) is not random, and it does not make sense anymore to talk about the probability that it includes the true $\beta _{1}.$

- **Once the confidence interval is computed, it either contains the true value $\beta _{1}$ or it does not**.
