---
title: "Lecture 6: Estimating the variance of errors"
subtitle: "Economics 326 — Introduction to Econometrics II"
author:
  - name: "Vadim Marmer, UBC"
format:
  html:
    output-file: 326_06_errors_variance.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: katex
  pdf:
    output-file: 326_06_errors_variance.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
  revealjs:
    output-file: 326_06_errors_variance_slides.html
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: katex
    scrollable: true
    chalkboard: false
    self-contained: true
    transition: none
---

## The importance of $\sigma^2$

::: {.hidden}
$
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
\gdef\Vhat#1{\widehat{\mathrm{Var}}\left(#1\right)}
\gdef\se#1{\mathrm{se}\left(#1\right)}
$
:::

- The variance of $\hat{\beta}$ depends on the unknown $\sigma^{2} = \E{U_i^{2} \mid \mathbf{X}}$:
  $$
  \Var{\hat{\beta} \mid \mathbf{X}} = \frac{\sigma^{2}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right)^{2}}.
  $$
- If $U$'s were observable, we could estimate $\sigma^{2}$ by $\frac{1}{n}\sum_{i=1}^{n}U_{i}^{2}$, which is unbiased. This is not possible as $U$'s are unobservable.
- Using sample residuals instead, $\hat{U}_{i}=Y_{i}-\hat{\alpha}-\hat{\beta}X_{i}$, gives a feasible estimator:
  $$
  \hat{\sigma}^{2}=\frac{1}{n}\sum_{i=1}^{n}\hat{U}_{i}^{2},
  $$
- $\hat{\sigma}^{2}$ is biased.

## An unbiased estimator of $\sigma^2$

- An unbiased estimator of $\sigma^{2}$ is
  $$
  s^{2}=\frac{1}{n-2}\sum_{i=1}^{n}\hat{U}_{i}^{2}.
  $$
- Assumptions:
  1. $Y_{i}=\alpha +\beta X_{i}+U_{i}$.
  2. $\E{U_{i}\mid \mathbf{X}} =0$ for all $i$.
  3. $\E{U_{i}^{2}\mid \mathbf{X}} =\sigma ^{2}$ for all $i$.
  4. $\E{U_{i}U_{j}\mid \mathbf{X}} =0$ for all $i\neq j$.
- Since $\hat{U}_{i}=Y_{i}-\hat{\alpha}-\hat{\beta}X_{i}$, dividing by $n-2$ adjusts for estimating two parameters: $\alpha$ and $\beta$.

## Expressing $\hat{U}_i$ in terms of $U_i$

- $\hat{U}_{i}=Y_{i}-\hat{\alpha}-\hat{\beta}X_{i}$
- $\hat{\alpha}=\bar{Y}-\hat{\beta}\bar{X}$, so
  $$
  \hat{U}_{i} = \left( Y_{i}-\bar{Y}\right) -\hat{\beta}\left( X_{i}-\bar{X}\right)
  $$

- Also:
  $$
  Y_{i}-\bar{Y} = \beta \left( X_{i}-\bar{X}\right) +U_{i}-\bar{U},
  $$
- We have the following relationship between $\hat{U}_i$ and $U_i$:
  $$
  \hat{U}_{i}=U_{i}-\bar{U} -\left( \hat{\beta}-\beta \right)\left( X_{i}-\bar{X}\right).
  $$
- $\hat{U}_i$ is related to $U_i$, but it is contaminated by the estimation errors.

## Expanding $\sum \hat{U}_i^2$

- We have:
  $$
  \hat{U}_{i} = \left( Y_{i}-\bar{Y}\right) -\hat{\beta}\left( X_{i}-\bar{X}\right)
  $$
- The squared residual is:
  \begin{align*}
  \hat{U}_{i}^{2} &= \left( U_{i}-\bar{U}\right) ^{2}+\left( \hat{\beta}-\beta \right)^{2}\left( X_{i}-\bar{X}\right)^{2} \\
  &\quad -2\left( \hat{\beta}-\beta \right) \left( X_{i}-\bar{X}\right) \left( U_{i}-\bar{U}\right).
  \end{align*}

- Thus,
  \begin{align*}
  \sum_{i=1}^{n}\hat{U}_{i}^{2} &=\sum_{i=1}^{n}\left( U_{i}-\bar{U}\right) ^{2}\\
  &\quad+\left( \hat{\beta}-\beta \right) ^{2}\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2} \\
  &\quad-2\left( \hat{\beta}-\beta \right) \sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) \left( U_{i}-\bar{U}\right).
  \end{align*}

- To show $\E{\sum_{i=1}^{n}\hat{U}_{i}^{2}\mid \mathbf{X}}=\left( n-2\right) \sigma ^{2}$, we verify the three terms on the RHS:

- **Claim 1:**
  $$
  \E{\sum_{i=1}^{n}\left( U_{i}-\bar{U}\right) ^{2}\mid \mathbf{X}}=\left( n-1\right) \sigma ^{2}.
  $$
- **Claim 2:**
  \begin{align*}
  &\E{\left( \hat{\beta}-\beta \right) ^{2}\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}\mid \mathbf{X}} \\
  &\quad =\sigma ^{2}.
  \end{align*}
- **Claim 3:**
  \begin{align*}
  &-2\cdot \E{\left( \hat{\beta}-\beta \right) \sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) \left( U_{i}-\bar{U}\right)\mid \mathbf{X}} \\
  &\quad =-2\sigma ^{2}.
  \end{align*}

## Estimating the variance of $\hat{\beta}$

- Variance of $\hat{\beta}$ conditional on $\mathbf{X}$:
  $$
  \Var{\hat{\beta}\mid \mathbf{X}} =\frac{\sigma ^{2}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}}.
  $$
- Estimator of $\sigma ^{2}$:
  \begin{align*}
  s^{2} &= \frac{1}{n-2}\sum_{i=1}^{n}\hat{U}_{i}^{2} \\
  &= \frac{1}{n-2}\sum_{i=1}^{n}\left( Y_{i}-\hat{\alpha}-\hat{\beta}X_{i}\right) ^{2}.
  \end{align*}
- Estimator of the variance of $\hat{\beta}$:
  $$
  \Vhat{\hat{\beta}} =\frac{s^{2}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}}.
  $$
- Standard error of $\hat{\beta}$:
  \begin{align*}
  \se{\hat{\beta}} &= \sqrt{\Vhat{\hat{\beta}}} \\
  &= \sqrt{\frac{s^{2}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}}}.
  \end{align*}

## Example in R

- Regress hourly wage on years of education using the `wage1` dataset from Wooldridge:

  ```{r}
  library(wooldridge)
  data("wage1")
  fit <- lm(wage ~ educ, data = wage1)
  summary(fit)
  ```

- `Std. Error` column: standard errors $\se{\hat{\alpha}}$ and $\se{\hat{\beta}}$
- `Residual standard error`: $s = \sqrt{s^2}$; so $s^2 = 3.378^2 \approx 11.41$
- The estimate $s^2$ can also be extracted directly:

  ```{r}
  summary(fit)$sigma^2
  ```

## Proof of Claim 1

\begin{align*}
\sum_{i=1}^{n}\left( U_{i}-\bar{U}\right) ^{2}
&= \sum_{i=1}^{n}U_{i}^{2}-\frac{1}{n}\left( \sum_{i=1}^{n}U_{i}\right) ^{2} \\
&= \sum_{i=1}^{n}U_{i}^{2} \\
&\quad -\frac{1}{n}\left( \sum_{i=1}^{n}U_{i}^{2}+\sum_{i=1}^{n}\sum_{j\neq i}U_{i}U_{j}\right).
\end{align*}
Taking conditional expectations and using the assumptions,
\begin{align*}
\E{\sum_{i=1}^{n}\left( U_{i}-\bar{U}\right) ^{2}\mid \mathbf{X}}
&= n\sigma ^{2}-\frac{1}{n}\cdot n\sigma ^{2} \\
&= \left( n-1\right) \sigma ^{2}.
\end{align*}

## Proof of Claim 2

Because $\E{\hat{\beta}\mid \mathbf{X}}=\beta$,
\begin{align*}
\E{\left( \hat{\beta}-\beta \right) ^{2}\mid \mathbf{X}}
&= \Var{\hat{\beta}\mid \mathbf{X}} \\
&= \frac{\sigma ^{2}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}}.
\end{align*}
Hence,
\begin{align*}
&\E{\left( \hat{\beta}-\beta \right) ^{2}\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}\mid \mathbf{X}} \\
&\quad =\sigma^{2}.
\end{align*}

## Proof of Claim 3

Note that
$$
\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) \left( U_{i}-\bar{U}\right) =\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) U_{i},
$$
and
$$
\hat{\beta}-\beta =\frac{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) U_{i}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}}.
$$
Therefore,
\begin{align*}
&\left( \hat{\beta}-\beta \right) \sum_{i=1}^{n}\left( X_{i}-\bar{X}\right)\left( U_{i}-\bar{U}\right) \\
&\quad =\frac{\left( \sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) U_{i}\right) ^{2}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}}.
\end{align*}
Conditionally on $\mathbf{X}$,
\begin{align*}
&\E{\left( \hat{\beta}-\beta \right) \sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) \left( U_{i}-\bar{U}\right)\mid \mathbf{X}} \\
&\quad = \frac{\sigma ^{2}\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}} \\
&\quad = \sigma ^{2}.
\end{align*}

## Putting it together

Using the three expectations above,
\begin{align*}
\E{\sum_{i=1}^{n}\hat{U}_{i}^{2}\mid \mathbf{X}}
&= \left( n-1\right) \sigma ^{2}+\sigma^{2}-2\sigma ^{2} \\
&= \left( n-2\right) \sigma ^{2},
\end{align*}
so $s^{2}$ is unbiased for $\sigma^{2}$.
