---
title: "Lecture 17: Asymptotics"
subtitle: "Economics 326 — Introduction to Econometrics II"
author:
  - name: "Vadim Marmer, UBC"
format:
  html:
    output-file: 326_17_asy.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: katex
    embed-resources: true
  pdf:
    output-file: 326_17_asy.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
  revealjs:
    output-file: 326_17_asy_slides.html
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: katex
    scrollable: true
    chalkboard: false
    self-contained: true
    transition: none
---

# Part I: Consistency

## Why we need large-sample theory

::: {.hidden}
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
\gdef\Vhat#1{\widehat{\mathrm{Var}}\left(#1\right)}
\gdef\se#1{\mathrm{se}\left(#1\right)}
:::

- The OLS estimator $\hat{\beta}$ has desirable properties:

  - $\hat{\beta}$ is unbiased if the errors are **strongly exogenous**: $\E{U_i \mid \mathbf{X}} =0.$

  - If in addition the errors are **homoskedastic**, then $\Vhat{\hat{\beta}}=s^{2}/\sum_{i=1}^{n}\left( X_{i}-\bar{X}\right) ^{2}$ is an unbiased estimator of the conditional variance of $\hat{\beta}$.

  - If in addition the errors are **normally** distributed (given $\mathbf{X}$), then $T=\left( \hat{\beta}-\beta \right) /\sqrt{\Vhat{\hat{\beta}}}$ has a $t$ distribution which can be used for hypothesis testing.

## Limitations of finite-sample theory

- If the errors are only **weakly exogenous**:
  $$
  \E{X_{i}U_{i}} =0,
  $$
  the OLS estimator is in general biased.

- If the errors are **heteroskedastic**:
  $$
  \E{U_{i}^{2} \mid X_{i}} =h\left( X_{i}\right),
  $$
  the "usual" variance formula is invalid; we also do not have an unbiased estimator for the variance in this case.

- If the errors are **not normally distributed** conditional on $\mathbf{X}$, then $T$- and $F$-statistics do not have $t$ and $F$ distributions under the null hypothesis.

- Asymptotic (large-sample) theory allows us to derive **approximate** properties and distributions of estimators and test statistics by assuming that the sample size $n$ is very large.

## Convergence in probability and LLN

- Let $\theta _{n}$ be a sequence of random variables indexed by the sample size $n.$ We say that $\theta _{n}$ **converges in probability** if
  $$
  \lim_{n\rightarrow \infty }P\left( \left\vert \theta _{n}-\theta \right\vert \geq \varepsilon \right) =0\text{ for all }\varepsilon >0.
  $$

- We denote this as $\theta _{n}\rightarrow _{p}\theta$ or $p\lim \theta _{n}=\theta.$

- An example of convergence in probability is a Law of Large Numbers (LLN):

  Let $X_{1},X_{2},\ldots ,X_{n}$ be a random sample such that $\E{X_{i}} =\mu$ for all $i=1,\ldots ,n,$ and define $\bar{X}_{n}=\frac{1}{n}\sum_{i=1}^{n}X_{i}.$ Then, under certain conditions,
  $$
  \bar{X}_{n}\rightarrow _{p}\mu.
  $$

## LLN

- Let $X_{1},\ldots ,X_{n}$ be a sample of **independent identically distributed** (**iid**) random variables. Let $\E{X_{i}}=\mu$. If $\Var{X_{i}}=\sigma ^{2}<\infty$, then
  $$
  \bar{X}_{n}\rightarrow _{p}\mu.
  $$

- In fact when the data are **iid**, the LLN holds if
  $$
  \E{\left\vert X_{i}\right\vert} <\infty,
  $$
  but we prove the result under a stronger assumption that $\Var{X_{i}}<\infty.$

## Markov's inequality

- **Markov's inequality.** Let $W$ be a random variable. For $\varepsilon >0$ and $r>0$,
  $$
  P\left( \left\vert W\right\vert \geq \varepsilon \right) \leq \frac{\E{\left\vert W\right\vert ^{r}}}{\varepsilon ^{r}}.
  $$

- With $r=2,$ we have **Chebyshev's inequality**. Suppose that $\E{X}=\mu.$ Take $W\equiv X-\mu$ and apply Markov's inequality with $r=2$. For $\varepsilon >0,$

  $$
  \begin{aligned}
  P\left( \left\vert X-\mu \right\vert \geq \varepsilon \right) &\leq \frac{\E{\left\vert X-\mu \right\vert ^{2}}}{\varepsilon ^{2}} \\
  &= \frac{\Var{X}}{\varepsilon ^{2}}.
  \end{aligned}
  $$

- The probability of observing an outlier (a large deviation of $X$ from its mean $\mu$) can be bounded by the variance.

## Proof of the LLN

Fix $\varepsilon >0$ and apply Markov's inequality with $r=2:$

$$
\begin{aligned}
P\left( \left\vert \bar{X}_{n}-\mu \right\vert \geq \varepsilon \right)
&= P\left( \left\vert \frac{1}{n}\sum_{i=1}^{n}X_{i}-\mu \right\vert \geq \varepsilon \right)  \\
&= P\left( \left\vert \frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\mu \right) \right\vert \geq \varepsilon \right)  \\
&\leq \frac{\E{\left( \frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\mu \right) \right) ^{2}}}{\varepsilon ^{2}} \\
&= \frac{1}{n^{2}\varepsilon ^{2}}\left( \sum_{i=1}^{n}\E{\left( X_{i}-\mu \right) ^{2}}+\sum_{i=1}^{n}\sum_{j\neq i}\E{\left( X_{i}-\mu \right) \left( X_{j}-\mu \right)} \right)  \\
&= \frac{1}{n^{2}\varepsilon ^{2}}\left( \sum_{i=1}^{n}\Var{X_{i}}+\sum_{i=1}^{n}\sum_{j\neq i}\Cov{X_{i},X_{j}}\right)  \\
&= \frac{n\sigma ^{2}}{n^{2}\varepsilon ^{2}} = \frac{\sigma ^{2}}{n\varepsilon ^{2}} \rightarrow 0 \text{ as }n\rightarrow \infty \text{ for all }\varepsilon >0.
\end{aligned}
$$

## Averaging and variance reduction

- Let $X_{1},\ldots ,X_{n}$ be a sample and suppose that

  $$
  \begin{aligned}
  \E{X_{i}} &= \mu \text{ for all }i=1,\ldots ,n, \\
  \Var{X_{i}} &= \sigma ^{2}\text{ for all }i=1,\ldots ,n, \\
  \Cov{X_{i},X_{j}} &= 0\text{ for all }j\neq i.
  \end{aligned}
  $$

- The mean of the sample average:

  $$
  \begin{aligned}
  \E{\bar{X}_{n}} &= \E{\frac{1}{n}\sum_{i=1}^{n}X_{i}} \\
  &= \frac{1}{n}\sum_{i=1}^{n}\E{X_{i}} \\
  &= \frac{1}{n}\sum_{i=1}^{n}\mu = \frac{1}{n}n\mu =\mu.
  \end{aligned}
  $$

## Variance of the sample average

- The variance of the sample average:

  $$
  \begin{aligned}
  \Var{\bar{X}_{n}} &= \Var{\frac{1}{n}\sum_{i=1}^{n}X_{i}} \\
  &= \frac{1}{n^{2}}\Var{\sum_{i=1}^{n}X_{i}} \\
  &= \frac{1}{n^{2}}\left( \sum_{i=1}^{n}\Var{X_{i}}+\sum_{i=1}^{n}\sum_{j\neq i}\Cov{X_{i},X_{j}}\right) \\
  &= \frac{1}{n^{2}}\left( \sum_{i=1}^{n}\sigma ^{2}+\sum_{i=1}^{n}\sum_{j\neq i}0\right) \\
  &= \frac{1}{n^{2}}n\sigma ^{2}=\frac{\sigma ^{2}}{n}.
  \end{aligned}
  $$

- The variance of the average approaches zero as $n\rightarrow \infty$ if the observations are **uncorrelated**.

## Convergence in probability: properties

- **Slutsky's Lemma.** Suppose that $\theta _{n}\rightarrow _{p}\theta,$ and let $g$ be a function continuous at $\theta.$ Then,
  $$
  g\left( \theta _{n}\right) \rightarrow _{p}g\left( \theta \right).
  $$

  - If $\theta _{n}\rightarrow _{p}\theta,$ then $\theta _{n}^{2}\rightarrow _{p}\theta ^{2}.$

  - If $\theta _{n}\rightarrow _{p}\theta$ and $\theta \neq 0,$ then $1/\theta _{n}\rightarrow _{p}1/\theta.$

- Suppose that $\theta _{n}\rightarrow _{p}\theta$ and $\lambda _{n}\rightarrow _{p}\lambda.$ Then,

  - $\theta _{n}+\lambda _{n}\rightarrow _{p}\theta +\lambda.$

  - $\theta _{n}\lambda _{n}\rightarrow _{p}\theta \lambda.$

  - $\theta _{n}/\lambda _{n}\rightarrow _{p}\theta /\lambda$ provided that $\lambda \neq 0.$

## Consistency

- Let $\hat{\beta}_{n}$ be an estimator of $\beta$ based on a sample of size $n.$

- We say that $\hat{\beta}_{n}$ is a **consistent** estimator of $\beta$ if as $n\rightarrow \infty,$
  $$
  \hat{\beta}_{n}\rightarrow _{p}\beta.
  $$

- Consistency means that the **probability** of the event that the distance between $\hat{\beta}_{n}$ and $\beta$ exceeds $\varepsilon >0$ can be made arbitrarily small by increasing the sample size.

## Consistency of OLS

- Suppose that:

  1. The data $\left\{ \left( Y_{i},X_{i}\right) :i=1,\ldots ,n\right\}$ are iid.

  2. $Y_{i}=\beta _{0}+\beta _{1}X_{i}+U_{i},$ where $\E{U_{i}} =0.$

  3. $\E{X_{i}U_{i}} =0.$

  4. $0<\Var{X_{i}}<\infty.$

- Let $\hat{\beta}_{0,n}$ and $\hat{\beta}_{1,n}$ be the OLS estimators of $\beta _{0}$ and $\beta _{1}$ based on a sample of size $n$. Under Assumptions 1--4,
  $$
  \begin{aligned}
  \hat{\beta}_{0,n} &\rightarrow _{p}\beta _{0}, \\
  \hat{\beta}_{1,n} &\rightarrow _{p}\beta _{1}.
  \end{aligned}
  $$

- The key identifying assumption is Assumption 3: $\Cov{X_{i},U_{i}}=0.$

## Proof of consistency

- Write

  $$
  \begin{aligned}
  \hat{\beta}_{1,n} = \frac{\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) Y_{i}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2}}
  &= \beta _{1}+\frac{\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2}} \\
  &= \beta _{1}+\frac{\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i}}{\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2}}.
  \end{aligned}
  $$

- We will show that
  $$
  \begin{aligned}
  \frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i} &\rightarrow _{p}0, \\
  \frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2} &\rightarrow _{p}\Var{X_{i}},
  \end{aligned}
  $$

- Since $\Var{X_{i}}\neq 0,$
  $$
  \hat{\beta}_{1,n} = \beta _{1}+\frac{\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i}}{\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2}} \rightarrow _{p} \beta _{1}+\frac{0}{\Var{X_{i}}}= \beta _{1}.
  $$

## Numerator converges to zero

$$
\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i} = \frac{1}{n}\sum_{i=1}^{n}X_{i}U_{i}-\bar{X}_{n}\left( \frac{1}{n}\sum_{i=1}^{n}U_{i}\right).
$$

By the LLN,

$$
\begin{aligned}
\frac{1}{n}\sum_{i=1}^{n}X_{i}U_{i} &\rightarrow _{p}\E{X_{i}U_{i}} =0, \\
\bar{X}_{n} &\rightarrow _{p}\E{X_{i}}, \\
\frac{1}{n}\sum_{i=1}^{n}U_{i} &\rightarrow _{p}\E{U_{i}} =0.
\end{aligned}
$$

Hence,

$$
\begin{aligned}
\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i} &= \frac{1}{n}\sum_{i=1}^{n}X_{i}U_{i}-\bar{X}_{n}\left( \frac{1}{n}\sum_{i=1}^{n}U_{i}\right) \\
&\rightarrow _{p}0-\E{X_{i}} \cdot 0 = 0.
\end{aligned}
$$

## Denominator converges to $\Var{X_i}$

- The sample variance can be written as
  $$
  \frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2} = \frac{1}{n}\sum_{i=1}^{n}X_{i}^{2}-\bar{X}_{n}^{2}.
  $$

- By the LLN, $\frac{1}{n}\sum_{i=1}^{n}X_{i}^{2}\rightarrow _{p}\E{X_{i}^{2}}$ and $\bar{X}_{n}\rightarrow _{p}\E{X_{i}}.$

- By Slutsky's Lemma, $\bar{X}_{n}^{2}\rightarrow _{p}\left( \E{X_{i}}\right) ^{2}.$

- Thus,
  $$
  \frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2}=\frac{1}{n}\sum_{i=1}^{n}X_{i}^{2}-\bar{X}_{n}^{2}\rightarrow _{p}\E{X_{i}^{2}} -\left( \E{X_{i}}\right) ^{2}=\Var{X_{i}}.
  $$

## Multiple regression

- Under similar conditions to 1--4, one can establish consistency of OLS for the multiple linear regression model:
  $$
  Y_{i}=\beta _{0}+\beta _{1}X_{1,i}+\ldots +\beta _{k}X_{k,i}+U_{i},
  $$
  where $\E{U_{i}}=0.$

- The key assumption is that the errors and regressors are uncorrelated:
  $$
  \E{X_{1,i}U_{i}} =\ldots =\E{X_{k,i}U_{i}} =0.
  $$

## Omitted variables and OLS inconsistency

- Suppose that the true model has two regressors:
  $$
  \begin{aligned}
  &Y_{i}=\beta _{0}+\beta _{1}X_{1,i}+\beta _{2}X_{2,i}+U_{i}, \\
  &\E{X_{1,i}U_{i}} =\E{X_{2,i}U_{i}} =0.
  \end{aligned}
  $$

- Suppose that the econometrician includes **only** $X_{1}$ in the regression when estimating $\beta _{1}$:

  $$
  \begin{aligned}
  \tilde{\beta}_{1,n} &= \frac{\sum_{i=1}^{n}\left( X_{1,i}-\bar{X}_{1,n}\right) Y_{i}}{\sum_{i=1}^{n}\left( X_{1,i}-\bar{X}_{1,n}\right) ^{2}} \\
  &= \frac{\sum_{i=1}^{n}\left( X_{1,i}-\bar{X}_{1,n}\right) \left( \beta _{0}+\beta _{1}X_{1,i}+\beta _{2}X_{2,i}+U_{i}\right) }{\sum_{i=1}^{n}\left( X_{1,i}-\bar{X}_{1,n}\right) ^{2}} \\
  &= \beta _{1}+\beta _{2}\frac{\sum_{i=1}^{n}\left( X_{1,i}-\bar{X}_{1,n}\right) X_{2,i}}{\sum_{i=1}^{n}\left( X_{1,i}-\bar{X}_{1,n}\right) ^{2}}+\frac{\sum_{i=1}^{n}\left( X_{1,i}-\bar{X}_{1,n}\right) U_{i}}{\sum_{i=1}^{n}\left( X_{1,i}-\bar{X}_{1,n}\right) ^{2}}.
  \end{aligned}
  $$

- Dividing numerator and denominator by $n$ and applying the LLN as before:

  - The noise term vanishes:
    $$
    \frac{\frac{1}{n}\sum_{i=1}^{n}\left( X_{1,i}-\bar{X}_{1,n}\right) U_{i}}{\frac{1}{n}\sum_{i=1}^{n}\left( X_{1,i}-\bar{X}_{1,n}\right) ^{2}} \rightarrow _{p} \frac{\Cov{X_{1,i},U_{i}}}{\Var{X_{1,i}}} = 0.
    $$

  - The bias term converges:
    $$
    \frac{\frac{1}{n}\sum_{i=1}^{n}\left( X_{1,i}-\bar{X}_{1,n}\right) X_{2,i}}{\frac{1}{n}\sum_{i=1}^{n}\left( X_{1,i}-\bar{X}_{1,n}\right) ^{2}} \rightarrow _{p} \frac{\Cov{X_{1,i},X_{2,i}}}{\Var{X_{1,i}}}.
    $$

- Therefore,
  $$
  \tilde{\beta}_{1,n} \rightarrow _{p} \beta _{1}+\beta _{2}\frac{\Cov{X_{1,i},X_{2,i}}}{\Var{X_{1,i}}}.
  $$

- $\tilde{\beta}_{1,n}$ is **inconsistent** unless:

  1. $\beta _{2}=0$ (the model is correctly specified).

  2. $\Cov{X_{1,i},X_{2,i}}=0$ (the omitted variable is **uncorrelated** with the included regressor).

## OVB through the composite error

- In this example, the model contains two regressors:
  $$
  \begin{aligned}
  &Y_{i}=\beta _{0}+\beta _{1}X_{1,i}+\beta _{2}X_{2,i}+U_{i}, \\
  &\E{X_{1,i}U_{i}} =\E{X_{2,i}U_{i}} =0.
  \end{aligned}
  $$

- However, since $X_{2}$ is not controlled for, it goes into the error term:
  $$
  \begin{aligned}
  Y_{i} &= \beta _{0}+\beta _{1}X_{1,i}+V_{i},\text{ where} \\
  V_{i} &= \beta _{2}X_{2,i}+U_{i}.
  \end{aligned}
  $$

- For consistency of $\tilde{\beta}_{1,n}$ we need $\Cov{X_{1,i},V_{i}} = 0$; however,

  $$
  \begin{aligned}
  \Cov{X_{1,i},V_{i}} &= \Cov{X_{1,i},\beta _{2}X_{2,i}+U_{i}} \\
  &= \Cov{X_{1,i},\beta _{2}X_{2,i}}+\Cov{X_{1,i},U_{i}} \\
  &= \beta _{2}\Cov{X_{1,i},X_{2,i}}+0 \\
  &\neq 0\text{, unless }\beta _{2}=0\text{ or }\Cov{X_{1,i},X_{2,i}}=0.
  \end{aligned}
  $$

# Part II: Asymptotic Normality

## Why do we need asymptotic normality?

- In the previous lectures, we showed that the OLS estimator has an **exact** normal distribution when the **errors are normally distributed**.

  - The same assumption is needed to show that the $T$ statistic has a $t$-distribution and the $F$ statistic has an $F$-distribution.

- In this lecture, we argue that even when the errors are **not** normally distributed, the OLS estimator has an **approximately normal distribution** in large samples, provided that some additional conditions hold.

  - This property is used for hypothesis testing: in large samples, the $T$ statistic has a standard normal distribution and the $F$ statistic has a $\chi^{2}$ distribution (approximately).

## Asymptotic normality

- Let $W_{n}$ be a sequence of random variables indexed by the sample size $n.$

  - Typically, $W_{n}$ will be a function of some estimator, such as $W_{n}=\sqrt{n}\left( \hat{\beta}_{n}-\beta \right)$.

- We say that $W_{n}$ has an **asymptotically** normal distribution if its **CDF** converges to a **normal CDF**.

- Let $W$ be any random variable with a normal $N\left( 0,\sigma^{2}\right)$ distribution and let $F$ denote its CDF. We say that $W_{n}$ has an asymptotically normal distribution if for all $x\in \mathbb{R}$:

  $$
  F_{n}\left( x\right) =P\left( W_{n}\leq x\right) \rightarrow P\left( W\leq x\right) =F\left( x\right) \text{ as }n\rightarrow \infty .
  $$

  - We denote this as $W_{n}\rightarrow _{d}W$ or $W_{n}\rightarrow _{d}N\left( 0,\sigma ^{2}\right).$

## Convergence in distribution

- Asymptotic normality is an example of convergence in distribution.

- We say that a sequence of random variables $W_{n}$ converges in distribution to $W$ (denoted as $W_{n}\rightarrow _{d}W$) if the CDF of $W_{n}$ converges to the CDF of $W$ at all points where the CDF of $W$ is continuous.

- Convergence in distribution is convergence of the **CDFs**.

## Central Limit Theorem (CLT)

- An example of convergence in distribution is a CLT.

- Let $X_{1},\ldots ,X_{n}$ be a sample of **iid** random variables such that $\E{X_{i}} =0$ and $\Var{X_{i}} =\sigma ^{2}>0$ (finite). Then, as $n\rightarrow \infty,$

  $$
  \frac{1}{\sqrt{n}}\sum_{i=1}^{n}X_{i}\rightarrow _{d}N\left( 0,\sigma^{2}\right) .
  $$

## CLT with non-zero mean

- For the CLT we impose 3 assumptions: **(1)** iid; **(2)** Mean zero; **(3)** Finite variance different from zero.

- If $X_{1},\ldots ,X_{n}$ are iid but $\E{X_{i}} =\mu \neq 0,$ then consider $X_{i}-\mu.$ Since $\E{X_{i}-\mu} =0,$ we have

  $$
  \frac{1}{\sqrt{n}}\sum_{i=1}^{n}\left( X_{i}-\mu \right) \rightarrow_{d}N\left( 0,\Var{X_{i}} \right) .
  $$

  Then

  $$
  \begin{aligned}
  \frac{1}{\sqrt{n}}\sum_{i=1}^{n}\left( X_{i}-\mu \right) &= \sqrt{n}\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\mu \right) \\
  &= \sqrt{n}\left( \frac{1}{n}\sum_{i=1}^{n}X_{i}-\frac{1}{n}\sum_{i=1}^{n}\mu \right) \\
  &= \sqrt{n}\left( \bar{X}_{n}-\mu \right) .
  \end{aligned}
  $$

## CLT for the sample average

- From the previous slide:
  $$
  \frac{1}{\sqrt{n}}\sum_{i=1}^{n}\left( X_{i}-\mu \right) = \sqrt{n}\left( \bar{X}_{n}-\mu \right) .
  $$

- Thus, the CLT can be stated as

  $$
  \sqrt{n}\left( \bar{X}_{n}-\mu \right) \rightarrow _{d}N\left( 0,\Var{X_{i}} \right) .
  $$

- By the LLN,

  $$
  \bar{X}_{n}-\mu \rightarrow _{p}0,
  $$

  and

  $$
  \Var{\sqrt{n}\left( \bar{X}_{n}-\mu \right)} = n\Var{\bar{X}_{n}} = n\frac{\Var{X_{i}}}{n} = \Var{X_{i}}.
  $$

## Properties

- Suppose that $W_{n}\rightarrow _{d}N\left( 0,\sigma ^{2}\right)$ and $\theta _{n}\rightarrow _{p}\theta.$ Then,

  $$
  \theta _{n}W_{n}\rightarrow _{d}\theta N\left( 0,\sigma ^{2}\right) \stackrel{d}{=} N\left( 0,\theta ^{2}\sigma ^{2}\right) ,
  $$

  and

  $$
  \theta _{n}+W_{n}\rightarrow _{d}\theta +N\left( 0,\sigma ^{2}\right) \stackrel{d}{=} N\left( \theta ,\sigma ^{2}\right) .
  $$

- Suppose that $Z_{n}\rightarrow _{d}Z\sim N\left( 0,1\right).$ Then,

  $$
  Z_{n}^{2}\rightarrow _{d}Z^{2}\equiv \chi _{1}^{2}.
  $$

- If $W_{n}\rightarrow _{d}c=$ constant, then $W_{n}\rightarrow _{p}c.$

## Asymptotic normality of OLS

- Suppose that:

  1. The data $\left\{ \left( Y_{i},X_{i}\right) :i=1,\ldots ,n\right\}$ are iid.
  2. $Y_{i}=\beta _{0}+\beta _{1}X_{i}+U_{i},$ where $\E{U_{i}} =0.$
  3. $\E{X_{i}U_{i}} =0.$
  4. $0<\Var{X_{i}} <\infty.$
  5. $0<\E{\left( X_{i}-\E{X_{i}}\right) ^{2}U_{i}^{2}} <\infty$ and $0<\E{U_{i}^{2}} <\infty.$

- Let $\hat{\beta}_{1,n}$ be the OLS estimator of $\beta _{1}.$ Then,

  $$
  \sqrt{n}\left( \hat{\beta}_{1,n}-\beta _{1}\right) \rightarrow _{d}N\left( 0,\frac{\E{\left( X_{i}-\E{X_{i}}\right) ^{2}U_{i}^{2}}}{\left(\Var{X_{i}} \right) ^{2}}\right).
  $$

- $V=\dfrac{\E{\left( X_{i}-\E{X_{i}}\right) ^{2}U_{i}^{2}}}{\left(\Var{X_{i}} \right) ^{2}}$ is called the **asymptotic variance** of $\hat{\beta}_{1,n}.$

## Large-sample approximation for OLS

- Let $\overset{a}{\sim}$ denote "approximately in large samples."

- The asymptotic normality

  $$
  \sqrt{n}\left( \hat{\beta}_{1,n}-\beta _{1}\right) \rightarrow _{d}N\left(0,V\right)
  $$

  can be viewed as the following large-sample **approximation**:

  $$
  \sqrt{n}\left( \hat{\beta}_{1,n}-\beta _{1}\right) \overset{a}{\sim} N\left(0,V\right) ,
  $$

  or

  $$
  \hat{\beta}_{1,n}\overset{a}{\sim} N\left( \beta _{1},V/n\right) .
  $$

## Proof: decomposition

Write

$$
\hat{\beta}_{1,n}=\beta _{1}+\frac{\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i}}{\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2}}.
$$

Now

$$
\hat{\beta}_{1,n}-\beta _{1}=\frac{\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i}}{\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2}},
$$

and

$$
\sqrt{n}\left( \hat{\beta}_{1,n}-\beta _{1}\right) =\frac{\frac{1}{\sqrt{n}}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i}}{\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2}}.
$$

## Proof: combining the limits

$$
\sqrt{n}\left( \hat{\beta}_{1,n}-\beta _{1}\right) =\frac{\frac{1}{\sqrt{n}}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i}}{\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2}}.
$$

In Part I, we established

$$
\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2}\rightarrow_{p}\Var{X_{i}}.
$$

We will show that

$$
\frac{1}{\sqrt{n}}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i}\rightarrow _{d}N\left( 0,\E{\left( X_{i}-\E{X_{i}}\right)^{2}U_{i}^{2}} \right),
$$

so that

\begin{align*}
\sqrt{n}\left( \hat{\beta}_{1,n}-\beta _{1}\right) &= \frac{\frac{1}{\sqrt{n}}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i}}{\frac{1}{n}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) ^{2}} \\
&\rightarrow _{d}\frac{N\left( 0,\E{\left( X_{i}-\E{X_{i}}\right) ^{2}U_{i}^{2}} \right)}{\Var{X_{i}}} \\
&\stackrel{d}{=} N\left( 0,\frac{\E{\left( X_{i}-\E{X_{i}}\right)^{2}U_{i}^{2}}}{\left(\Var{X_{i}} \right) ^{2}}\right).
\end{align*}

## Proof: numerator CLT

$$
\begin{aligned}
\frac{1}{\sqrt{n}}\sum_{i=1}^{n}\left( X_{i}-\bar{X}_{n}\right) U_{i} &= \frac{1}{\sqrt{n}}\sum_{i=1}^{n}\left( X_{i}-\E{X_{i}}+\E{X_{i}}-\bar{X}_{n}\right) U_{i} \\
&= \frac{1}{\sqrt{n}}\sum_{i=1}^{n}\left( X_{i}-\E{X_{i}}\right) U_{i}+\left( \E{X_{i}}-\bar{X}_{n}\right) \frac{1}{\sqrt{n}}\sum_{i=1}^{n}U_{i}.
\end{aligned}
$$

We have

$$
\E{\left( X_{i}-\E{X_{i}}\right) U_{i}} = \E{X_{i}U_{i}}-\E{X_{i}}\E{U_{i}}=0,
$$

and $0<\E{\left( X_{i}-\E{X_{i}}\right) ^{2}U_{i}^{2}} <\infty,$ so by the CLT,

$$
\frac{1}{\sqrt{n}}\sum_{i=1}^{n}\left( X_{i}-\E{X_{i}}\right) U_{i}\rightarrow_{d}N\left( 0,\E{\left( X_{i}-\E{X_{i}}\right) ^{2}U_{i}^{2}}\right).
$$

## Proof: second term vanishes

It is left to show that

$$
\left( \E{X_{i}}-\bar{X}_{n}\right) \frac{1}{\sqrt{n}}\sum_{i=1}^{n}U_{i}\rightarrow _{p}0.
$$

We have $\E{U_{i}}=0$ and $0<\E{U_{i}^{2}}<\infty.$ By the CLT,

$$
\frac{1}{\sqrt{n}}\sum_{i=1}^{n}U_{i}\rightarrow _{d}N\left(0,\E{U_{i}^{2}}\right).
$$

By the LLN,

$$
\E{X_{i}}-\bar{X}_{n}\rightarrow _{p}0.
$$

Hence, the result follows.

# Part III: Asymptotic Variance

## Asymptotic variance

- In Part II, we showed that when the data are iid and the regressors are exogenous,
  $$
  \begin{aligned}
  Y_{i} &= \beta_{0} + \beta_{1}X_{i} + U_{i}, \\
  \E{U_{i}} &= \E{X_{i}U_{i}} = 0,
  \end{aligned}
  $$
  the OLS estimator of $\beta_{1}$ is asymptotically normal:
  $$
  \begin{aligned}
  \sqrt{n}\left(\hat{\beta}_{1,n} - \beta_{1}\right) &\rightarrow_{d} N\left(0, V\right), \\
  V &= \frac{\E{\left(X_{i} - \E{X_{i}}\right)^{2}U_{i}^{2}}}{\left(\Var{X_{i}}\right)^{2}}.
  \end{aligned}
  $$

- For hypothesis testing, we need a **consistent** estimator of the asymptotic variance $V$:
  $$
  \hat{V}_{n} \rightarrow_{p} V.
  $$

## Simplifying $V$ under homoskedasticity

- Assume that the errors are **homoskedastic**:
  $$
  \E{U_{i}^{2} \mid X_{i}} = \sigma^{2} \text{ for all } X_{i}\text{'s.}
  $$

- In this case, the asymptotic variance can be simplified using the Law of Iterated Expectation:
  $$
  \begin{aligned}
  \E{\left(X_{i} - \E{X_{i}}\right)^{2}U_{i}^{2}} &= \E{\E{\left(X_{i} - \E{X_{i}}\right)^{2}U_{i}^{2} \mid X_{i}}} \\
  &= \E{\left(X_{i} - \E{X_{i}}\right)^{2} \E{U_{i}^{2} \mid X_{i}}} \\
  &= \E{\left(X_{i} - \E{X_{i}}\right)^{2} \sigma^{2}} \\
  &= \sigma^{2}\,\E{\left(X_{i} - \E{X_{i}}\right)^{2}} = \sigma^{2}\Var{X_{i}}.
  \end{aligned}
  $$

## Estimating $V$: method of moments

- Thus, when the errors are homoskedastic with $\E{U_{i}^{2}} = \sigma^{2},$
  $$
  V = \frac{\E{\left(X_{i} - \E{X_{i}}\right)^{2}U_{i}^{2}}}{\left(\Var{X_{i}}\right)^{2}} = \frac{\sigma^{2}\Var{X_{i}}}{\left(\Var{X_{i}}\right)^{2}} = \frac{\sigma^{2}}{\Var{X_{i}}}.
  $$

- Let $\hat{U}_{i} = Y_{i} - \hat{\beta}_{0,n} - \hat{\beta}_{1,n}X_{i}$, where $\hat{\beta}_{0,n}$ and $\hat{\beta}_{1,n}$ are the OLS estimators of $\beta_{0}$ and $\beta_{1}.$

- A consistent estimator for the asymptotic variance can be constructed using the **Method of Moments**:
  $$
  \begin{aligned}
  \hat{\sigma}_{n}^{2} &= \frac{1}{n}\sum_{i=1}^{n}\hat{U}_{i}^{2}, \\
  \Vhat{X_{i}} &= \frac{1}{n}\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}, \text{ and} \\
  \hat{V}_{n} &= \frac{\hat{\sigma}_{n}^{2}}{\frac{1}{n}\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}}.
  \end{aligned}
  $$

## Why the LLN does not apply directly

- From the previous slide:
  $$
  \hat{V}_{n} = \frac{\hat{\sigma}_{n}^{2}}{\frac{1}{n}\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}},\quad \hat{\sigma}_{n}^{2} = \frac{1}{n}\sum_{i=1}^{n}\hat{U}_{i}^{2},\quad \hat{U}_{i} = Y_{i} - \hat{\beta}_{0,n} - \hat{\beta}_{1,n}X_{i}.
  $$

- When proving the consistency of OLS (Part I), we showed that
  $$
  \frac{1}{n}\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2} \rightarrow_{p} \Var{X_{i}},
  $$
  and to establish $\hat{V}_{n} \rightarrow_{p} V,$ we need to show that $\hat{\sigma}_{n}^{2} \rightarrow_{p} \sigma^{2}.$

- The LLN **cannot** be applied directly to
  $$
  \frac{1}{n}\sum_{i=1}^{n}\hat{U}_{i}^{2}
  $$
  because the $\hat{U}_{i}$'s are not iid: they are **dependent** through $\hat{\beta}_{0,n}$ and $\hat{\beta}_{1,n}.$

## Proof of $\hat{\sigma}^{2}_{n} \rightarrow_{p} \sigma^{2}$

- First, write
  $$
  \begin{aligned}
  \hat{U}_{i} &= Y_{i} - \hat{\beta}_{0,n} - \hat{\beta}_{1,n}X_{i} \\
  &= \left(\beta_{0} + \beta_{1}X_{i} + U_{i}\right) - \hat{\beta}_{0,n} - \hat{\beta}_{1,n}X_{i} \\
  &= U_{i} - \left(\hat{\beta}_{0,n} - \beta_{0}\right) - \left(\hat{\beta}_{1,n} - \beta_{1}\right)X_{i}.
  \end{aligned}
  $$

- Now,
  $$
  \hat{\sigma}_{n}^{2} = \frac{1}{n}\sum_{i=1}^{n}\hat{U}_{i}^{2} = \frac{1}{n}\sum_{i=1}^{n}\left(U_{i} - \left(\hat{\beta}_{0,n} - \beta_{0}\right) - \left(\hat{\beta}_{1,n} - \beta_{1}\right)X_{i}\right)^{2}.
  $$

## Completing the consistency proof

- We have
  $$
  \begin{aligned}
  \hat{\sigma}_{n}^{2} &= \frac{1}{n}\sum_{i=1}^{n}\left(U_{i} - \left(\hat{\beta}_{0,n} - \beta_{0}\right) - \left(\hat{\beta}_{1,n} - \beta_{1}\right)X_{i}\right)^{2} \\
  &= \frac{1}{n}\sum_{i=1}^{n}U_{i}^{2} + \left(\hat{\beta}_{0,n} - \beta_{0}\right)^{2} + \left(\hat{\beta}_{1,n} - \beta_{1}\right)^{2}\frac{1}{n}\sum_{i=1}^{n}X_{i}^{2} \\
  &\quad -2\left(\hat{\beta}_{0,n} - \beta_{0}\right)\frac{1}{n}\sum_{i=1}^{n}U_{i} - 2\left(\hat{\beta}_{1,n} - \beta_{1}\right)\frac{1}{n}\sum_{i=1}^{n}U_{i}X_{i} \\
  &\quad +2\left(\hat{\beta}_{0,n} - \beta_{0}\right)\left(\hat{\beta}_{1,n} - \beta_{1}\right)\frac{1}{n}\sum_{i=1}^{n}X_{i}.
  \end{aligned}
  $$

- By the LLN,
  $$
  \frac{1}{n}\sum_{i=1}^{n}U_{i}^{2} \rightarrow_{p} \E{U_{i}^{2}} = \sigma^{2}.
  $$

- Because $\hat{\beta}_{0,n}$ and $\hat{\beta}_{1,n}$ are consistent,
  $$
  \hat{\beta}_{0,n} - \beta_{0} \rightarrow_{p} 0 \text{ and } \hat{\beta}_{1,n} - \beta_{1} \rightarrow_{p} 0.
  $$

## Using $s^2$ instead of $\hat{\sigma}^2_n$

- Thus, when the errors are homoskedastic,
  $$
  \hat{V}_{n} = \frac{\hat{\sigma}_{n}^{2}}{\frac{1}{n}\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}}, \text{ with } \hat{\sigma}_{n}^{2} = \frac{1}{n}\sum_{i=1}^{n}\hat{U}_{i}^{2},
  $$
  is a consistent estimator of $V = \frac{\sigma^{2}}{\Var{X_{i}}}.$

- Similarly,
  $$
  s^{2} = \frac{1}{n-2}\sum_{i=1}^{n}\hat{U}_{i}^{2} \rightarrow_{p} \sigma^{2},
  $$
  and therefore
  $$
  \hat{V}_{n} = \frac{s^{2}}{\frac{1}{n}\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}}
  $$
  is also a consistent estimator of $V = \frac{\sigma^{2}}{\Var{X_{i}}}.$

- This version has an advantage over the one with $\hat{\sigma}_{n}^{2}$: in addition to being consistent, $s^{2}$ is also an unbiased estimator of $\sigma^{2}$ if the regressors are strongly exogenous.

## Asymptotic approximation

- The result $\sqrt{n}\left(\hat{\beta}_{1,n} - \beta_{1}\right) \rightarrow_{d} N\left(0, V\right)$ is used as the following **approximation**:
  $$
  \hat{\beta}_{1,n} \overset{a}{\sim} N\left(\beta_{1}, \frac{V}{n}\right),
  $$
  where $\overset{a}{\sim}$ denotes approximately in large samples. Thus, the variance of $\hat{\beta}_{1,n}$ can be taken as approximately $V/n.$

- With $\hat{V}_{n} = \frac{s^{2}}{\frac{1}{n}\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}}$ we have
  $$
  \frac{\hat{V}_{n}}{n} = \frac{s^{2}}{\frac{1}{n}\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}} \cdot \frac{1}{n} = \frac{s^{2}}{\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}}.
  $$

## Connection to the exact result

- From the previous slide:
  $$
  \frac{\hat{V}_{n}}{n} = \frac{s^{2}}{\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}}
  $$

- Thus, in the case of homoskedastic errors we have the following asymptotic approximation:
  $$
  \hat{\beta}_{1,n} \overset{a}{\sim} N\left(\beta_{1}, \frac{s^{2}}{\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}}\right).
  $$

- In finite samples, we have the same result **exactly**, when the regressors are **strongly exogenous** and the errors are **normal**.

## Asymptotic $T$-test

- Consider testing $H_{0}: \beta_{1} = \beta_{1,0}$ vs $H_{1}: \beta_{1} \neq \beta_{1,0}.$

- Consider the behavior of the $T$ statistic **under $H_{0}: \beta_{1} = \beta_{1,0}$**. Since
  $$
  \sqrt{n}\left(\hat{\beta}_{1,n} - \beta_{1}\right) \rightarrow_{d} N\left(0, V\right) \text{ and } \hat{V}_{n} \rightarrow_{p} V,
  $$
  we have
  $$
  \begin{aligned}
  T = \frac{\hat{\beta}_{1,n} - \beta_{1,0}}{\sqrt{\hat{V}_{n}/n}} &= \frac{\sqrt{n}\left(\hat{\beta}_{1,n} - \beta_{1,0}\right)}{\sqrt{\hat{V}_{n}}} \\
  &\overset{H_{0}}{=} \frac{\sqrt{n}\left(\hat{\beta}_{1,n} - \beta_{1}\right)}{\sqrt{\hat{V}_{n}}} \\
  &\rightarrow_{d} \frac{N\left(0, V\right)}{\sqrt{V}} \stackrel{d}{=} N\left(0, 1\right).
  \end{aligned}
  $$

## Asymptotic $T$-test: rejection rule

- Under $H_{0}: \beta_{1} = \beta_{1,0},$
  $$
  T = \frac{\hat{\beta}_{1,n} - \beta_{1,0}}{\sqrt{\hat{V}_{n}/n}} \rightarrow_{d} N\left(0, 1\right),
  $$
  provided that $\hat{V}_{n} \rightarrow_{p} V$ (the asymptotic variance of $\hat{\beta}_{1,n}$).

- An asymptotic size $\alpha$ test rejects $H_{0}: \beta_{1} = \beta_{1,0}$ against $H_{1}: \beta_{1} \neq \beta_{1,0}$ when
  $$
  \left|T\right| > z_{1-\alpha/2},
  $$
  where $z_{1-\alpha/2}$ is a **standard normal** critical value.

- Asymptotically, the variance of the OLS estimator is known; we behave as if the variance were known.

## Heteroskedastic errors

- In general, the **errors are heteroskedastic**: $\E{U_{i}^{2} \mid X_{i}}$ is not constant and changes with $X_{i}.$

- In this case, $\hat{V}_{n} = \frac{s^{2}}{\frac{1}{n}\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}}$ is **not** a consistent estimator of the asymptotic variance $V = \frac{\E{\left(X_{i} - \E{X_{i}}\right)^{2}U_{i}^{2}}}{\left(\Var{X_{i}}\right)^{2}}$:
  $$
  \begin{aligned}
  \frac{s^{2}}{\frac{1}{n}\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}} &\rightarrow_{p} \frac{\E{U_{i}^{2}}}{\Var{X_{i}}} \\
  &= \frac{\Var{X_{i}}\cdot\E{U_{i}^{2}}}{\left(\Var{X_{i}}\right)^{2}} \\
  &\neq \frac{\E{\left(X_{i} - \E{X_{i}}\right)^{2}U_{i}^{2}}}{\left(\Var{X_{i}}\right)^{2}}.
  \end{aligned}
  $$

## HC estimator of asymptotic variance

- In the case of heteroskedastic errors, a consistent estimator of $V = \frac{\E{\left(X_{i} - \E{X_{i}}\right)^{2}U_{i}^{2}}}{\left(\Var{X_{i}}\right)^{2}}$ can be constructed as follows:
  $$
  \hat{V}_{n}^{HC} = \frac{\frac{1}{n}\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}\hat{U}_{i}^{2}}{\left(\frac{1}{n}\sum_{i=1}^{n}\left(X_{i} - \bar{X}_{n}\right)^{2}\right)^{2}}.
  $$

- One can show that $\hat{V}_{n}^{HC} \rightarrow_{p} V$ whether the errors are heteroskedastic or homoskedastic.

- We have the following asymptotic approximation:
  $$
  \hat{\beta}_{1,n} \overset{a}{\sim} N\left(\beta_{1}, \frac{\hat{V}_{n}^{HC}}{n}\right),
  $$
  and the standard errors can be computed as $\se{\hat{\beta}_{1,n}} = \sqrt{\hat{V}_{n}^{HC}/n}.$

## HC standard errors in R

- In R, the HC estimator of standard errors can be obtained using the `sandwich` package:

  ```{r}
  #| echo: true
  library(wooldridge)
  library(lmtest)
  library(sandwich)
  data("wage1")
  reg <- lm(wage ~ educ + exper + tenure, data = wage1)
  ```

- **Standard (homoskedastic) standard errors:**

  ```{r}
  #| echo: true
  coeftest(reg)
  ```

- **HC (robust) standard errors:**

  ```{r}
  #| echo: true
  coeftest(reg, vcov = vcovHC(reg, type = "HC1"))
  ```
