---
title: "Lecture 1: Regression, identification, and the OLS estimator"
subtitle: "Economics 527: Econometric Methods"
author:
  - name: "Vadim Marmer, UBC"
format:
  html:
    output-file: 527_01_regression_ols.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: katex
    embed-resources: true
  pdf:
    output-file: 527_01_regression_ols.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
  revealjs:
    output-file: 527_01_regression_ols_slides.html
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: katex
    navigation-mode: linear
    scrollable: true
    chalkboard: false
    self-contained: true
    transition: none
---

# Motivation and data

## A motivating example: Return to education

::: {.hidden}
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
:::

- Suppose wages are generated according to:
$$
\begin{aligned}
\ln \text{Wage} = \beta_1 &+ {\color{blue}\beta_2}\,\text{Education} + \beta_3\,\text{Gender} + \beta_4\,\text{Race} \\
&+ \beta_5\,\text{Urban} + \beta_6\,\text{Experience} + U.
\end{aligned}
$$

- We have a sample of workers and, for each worker, we observe Wage, Education, etc.

- $U$ collects everything besides Education, Gender, etc. that affects wages. Ability is first on that list, and it is unobserved (latent).

- The unknown coefficient ${\color{blue}\beta_2}$ measures the return to education, holding everything else fixed.

- Questions/Issues:
  - Why in this equation $\beta_2$ measures the return to education;
  - How to "learn" $\beta_2$ from random data;
  - What data randomness means;
  - What has to be assumed about the unobserved $U$ to be able to learn $\beta_2$.

## Levels and logs 

- **Level-level**, $y = \beta x$:
$$
\beta = \frac{dy}{dx},
$$
and the proportional change in $y$ per unit of $x$ is $\beta / y$, which depends on where $y$ is.

- **Log-level**, $\ln y = \beta x$:
$$
\beta = \frac{d \ln y}{d x} = \frac{dy/dx}{y},
$$
so $\beta$ is the proportional rate of change of $y$ with respect to $x$. Multiply by 100 for a percentage.

- **Log-log**, $\ln y = \beta \ln x$:
$$
\beta = \frac{d \ln y}{d \ln x} = \frac{dy/y}{dx/x},
$$
the elasticity of $y$ with respect to $x$.



## Logs often "work better"

- Wages are bounded below by zero and have a long right tail. Their log is close to symmetric.

  ```{r}
  #| fig-align: center
  #| fig-width: 9
  #| fig-height: 3.6
  op <- par(mfrow = c(1, 2), mar = c(4, 4, 2.5, 1))
  x <- seq(0.01, 120, length.out = 600)
  plot(x, dlnorm(x, meanlog = log(25), sdlog = 0.6), type = "l", lwd = 2,
       xlab = "Wage", ylab = "Density", main = "Wage", yaxt = "n")
  z <- seq(log(25) - 2.2, log(25) + 2.2, length.out = 600)
  plot(z, dnorm(z, mean = log(25), sd = 0.6), type = "l", lwd = 2,
       xlab = "ln Wage", ylab = "Density", main = "ln Wage", yaxt = "n")
  par(op)
  ```

- A symmetric dependent variable is not required by anything that follows, but specifications with logs often "work better" in practice. 

## The data

- We observe a **sample** of $n$ workers:
$$
\{(Y_i, X_i) : i = 1, \ldots, n\}.
$$

- For worker $i$, $Y_i = \ln \text{Wage}_i$ is a scalar ($1\times 1$) and $X_i$ stacks the $k$ regressors, with a $1$ in the first position for the intercept:
$$
X_i = \begin{pmatrix} 1 \\ \text{Educ}_i \\ \text{Gender}_i \\ \vdots \end{pmatrix} \quad (k \times 1),
\qquad
\beta = \begin{pmatrix} \beta_1 \\ \beta_2 \\ \vdots \\ \beta_k \end{pmatrix} \quad (k \times 1).
$$

- The wage equation for worker $i$ is
$$
Y_i = X_i^\top \beta + U_i = \sum_{j=1}^{k} X_{ij}\,\beta_j + U_i,
$$
where $X_i^\top \beta = \beta^\top X_i$ is a scalar.

## What is random and what is fixed

- $\beta$ is **fixed** and unknown.

- $X_i$, $U_i$, and therefore $Y_i$ are **random**. 



- The **population** is the distribution that generates $\{(Y_i, X_i) : i = 1, \ldots, n\}$, a probability distribution rather than a group of people.

- In a cross-section, each observation is a different worker, so we take the observations to be drawn **independently and identically distributed**:
$$
(Y_i, X_i) \text{ are iid, with joint distribution } F_{Y, X_1, \ldots, X_k}.
$$

- Note that independence is between $(Y_i,X_i)$ and $(Y_j,X_j)$ for $i \neq j$, not between $Y_i$ and $X_i$.
- Under the iid assumption, we can take the population to be the joint distribution of one observation $(Y_i, X_i)$.


## Modelling the randomness

- A random experiment has a sample space $\Omega$ of outcomes $\omega$. These are possible states of the world.


- Let $X$ denote education and $V$ ability.
- The pair $(X, V)$ is a function from $\Omega$ to $\mathbb{R}^2$: for the state of the world $\omega$, $X(\omega)$ is the education and $V(\omega)$ is the ability.

  ```{r}
  #| fig-align: center
  #| fig-width: 8
  #| fig-height: 3.2
  op <- par(mar = c(0.5, 0.5, 0.5, 0.5))
  plot(NA, xlim = c(0, 10), ylim = c(0, 4), axes = FALSE, xlab = "", ylab = "", asp = 1)
  th <- seq(0, 2 * pi, length.out = 200)
  lines(2 + 1.6 * cos(th), 2 + 1.1 * sin(th), lwd = 2)
  text(0.5, 3.4, expression(Omega), cex = 1.4)
  points(c(1.5, 2.6), c(2.4, 1.6), pch = 19)
  text(1.5, 2.75, expression(omega[1])); text(2.6, 1.25, expression(omega[2]))
  arrows(6, 0.6, 9.6, 0.6, lwd = 2, length = 0.1); arrows(6, 0.6, 6, 3.6, lwd = 2, length = 0.1)
  text(9.3, 0.3, "Education"); text(6.9, 3.6, "Ability")
  points(c(8.2, 7.1), c(2.6, 1.4), pch = 19)
  arrows(1.7, 2.5, 8.05, 2.7, lwd = 1.5, length = 0.1, col = "gray30")
  arrows(2.8, 1.65, 6.95, 1.45, lwd = 1.5, length = 0.1, col = "gray30")
  text(5, 3.35, expression(bgroup("(", atop(X, V), ")") * (omega)), cex = 1.2)
  par(op)
  ```


- A probability $\mathrm{P}$ assigns to each event $A$ a number $\mathrm{P}(A) \in [0,1]$. It induces a distribution for $(X, V)$.
- The joint CDF of $(X,V)$ is a statement about sets of outcomes: for all $x, v \in \mathbb{R}$,
$$
\begin{aligned}
F_{X,V}(x,v) &= \mathrm{P}(X \le x,\, V \le v) \\
&= \mathrm{P}\big(\{\omega \in \Omega : X(\omega) \le x,\ V(\omega) \le v\}\big).
\end{aligned}
$$

# Probability

## Random variables and the CDF

- A random variable is a map $Y : \Omega \to \mathbb{R}$. For a set $A \subset \mathbb{R}$,
$$
\mathrm{P}(Y \in A) = \mathrm{P}\big(\{\omega : Y(\omega) \in A\}\big).
$$

- Taking $A = (-\infty, y]$ gives the **cumulative distribution function**:
$$
F_Y(y) = \mathrm{P}(Y \le y), \qquad y \in \mathbb{R}.
$$

- Properties:
  - $F_Y$ is nondecreasing and right-continuous, $F_Y(y) \to 0$ as $y \to -\infty$ and $F_Y(y) \to 1$ as $y \to \infty$;
  - $\mathrm{P}(a < Y \le b) = F_Y(b) - F_Y(a)$;
  - a jump of $F_Y$ at $y_1$ has size $\mathrm{P}(Y = y_1)$; where $F_Y$ is continuous, $\mathrm{P}(Y = y_1) = 0$.

## Continuous random variables and the PDF

- When $F_Y$ is differentiable, the **probability density function** is
$$
f_Y(y) = \frac{d F_Y(y)}{dy}, \qquad\text{so}\qquad F_Y(y) = \int_{-\infty}^{y} f_Y(v)\, dv.
$$

- $f_Y(y) \ge 0$, $\displaystyle\int_{-\infty}^{\infty} f_Y(y)\, dy = 1$, and for $A \subset \mathbb{R}$,
$$
\mathrm{P}(Y \in A) = \int_A f_Y(y)\, dy.
$$

- A density is not a probability: $\mathrm{P}(Y = y) = 0$ for every $y$, and $f_Y(y)$ can exceed one.

- Everything in this course is written for the continuous case. The discrete case replaces integrals with sums.

## Expectation

- For a function $h$,
$$
\E{h(Y)} = \int_{-\infty}^{\infty} h(y)\, f_Y(y)\, dy.
$$

- Throughout, every expectation we write down is assumed to exist and to be finite.

- $\E{Y}$ is the first moment, $\E{Y^2}$ the second, and
$$
\Var{Y} = \E{(Y - \E{Y})^2} = \E{Y^2} - (\E{Y})^2.
$$

- The mean is the best constant predictor of $Y$ under squared loss:
$$
\E{(Y - c)^2} = \E{(Y - \E{Y})^2} + (\E{Y} - c)^2 \ge \Var{Y},
$$
with equality only at $c = \E{Y}$. (The cross term $2(\E{Y} - c)\,\E{Y - \E{Y}}$ is zero.)

- Under absolute loss the minimiser of $\E{|Y - c|}$ is the median instead. Which loss we use decides which feature of the distribution we end up estimating.

## Joint distributions

- For two random variables $Y$ and $X$ on the same $\Omega$:
$$
\begin{aligned}
F_{Y,X}(y, x) &= \mathrm{P}(Y \le y,\, X \le x),\\
f_{Y,X}(y, x) &= \frac{\partial^2 F_{Y,X}(y,x)}{\partial y\, \partial x}.
\end{aligned}
$$

- $f_{Y,X} \ge 0$, $\displaystyle\int\!\!\int f_{Y,X}(y,x)\, dy\, dx = 1$, and for $A \subset \mathbb{R}^2$,
$$
\mathrm{P}\big((Y,X) \in A\big) = \int\!\!\int_A f_{Y,X}(y,x)\, dy\, dx.
$$

- With $k$ regressors the joint CDF of one observation is
$$
\begin{aligned}
F_{Y, X_1, \ldots, X_k}(y, x_1, \ldots, x_k)
&= \mathrm{P}(Y \le y,\ X_1 \le x_1,\ \ldots,\ X_k \le x_k).
\end{aligned}
$$

## From the joint density to a marginal

- **Claim.** The marginal density of $X$ is obtained by integrating $Y$ out:
$$
f_X(x) = \int_{-\infty}^{\infty} f_{Y,X}(y, x)\, dy.
$$

- **Proof.** Start from the marginal CDF and differentiate:
$$
\begin{aligned}
F_X(x) &= \mathrm{P}(X \le x) = \mathrm{P}(X \le x,\ Y < \infty) \\
&= \int_{-\infty}^{x} \int_{-\infty}^{\infty} f_{Y,X}(y, u)\, dy\, du, \\
f_X(x) &= \frac{d F_X(x)}{dx} = \frac{d}{dx}\left[ \int_{-\infty}^{x} \int_{-\infty}^{\infty} f_{Y,X}(y, u)\, dy\, du \right] \\
&= \int_{-\infty}^{\infty} f_{Y,X}(y, x)\, dy,
\end{aligned}
$$
using $\frac{d}{dx}\int_{-\infty}^{x} h(u)\, du = h(x)$.

- The converse fails: knowing $f_X$ and $f_Y$ does **not** determine $f_{Y,X}$. The marginals say nothing about how $Y$ and $X$ move together.

## Conditional density

- The density of $Y$ **given** $X = x$, defined where $f_X(x) > 0$:
$$
f_{Y \mid X}(y \mid x) = \frac{f_{Y,X}(y, x)}{f_X(x)}, \qquad y \in \mathbb{R}.
$$

- This has the same shape as conditional probability for events, $\mathrm{P}(A \mid B) = \mathrm{P}(A \cap B)/\mathrm{P}(B)$: the joint, rescaled by the marginal density of what we condition on.

- It is a proper density in $y$:
$$
\int_{-\infty}^{\infty} f_{Y \mid X}(y \mid x)\, dy
= \frac{1}{f_X(x)} \underbrace{\int_{-\infty}^{\infty} f_{Y,X}(y, x)\, dy}_{=\, f_X(x)} = 1.
$$

## Independence

- $Y$ and $X$ are **independent** if conditioning changes nothing:
$$
f_{Y \mid X}(y \mid x) = f_Y(y) \quad \text{for all } y \text{ and all } x \text{ with } f_X(x) > 0.
$$

- Equivalently the joint factorises:
$$
\begin{aligned}
f_{Y,X}(y, x) &= f_Y(y)\, f_X(x) \quad \text{for all } x, y, \\
\text{or equivalently}\qquad F_{Y,X}(y, x) &= F_Y(y)\, F_X(x) \quad \text{for all } x, y.
\end{aligned}
$$

- Independence is a strong statement: it involves the whole conditional distribution. Regression will need much less.

## Expectation of a function of two variables

- For a function $h$ of both arguments,
$$
\E{h(Y, X)} = \int_{-\infty}^{\infty}\!\!\int_{-\infty}^{\infty} h(y, x)\, f_{Y,X}(y, x)\, dy\, dx.
$$

- With $h(y, x) = yx$ this is the joint moment $\E{YX}$. Centring both arguments gives the **covariance**:
$$
\begin{aligned}
\Cov{Y, X} &= \E{\big(Y - \E{Y}\big)\big(X - \E{X}\big)} \\
&= \E{YX} - \E{Y}\,\E{X}.
\end{aligned}
$$

- When $\Var{Y}$ and $\Var{X}$ are positive and finite, dividing by the two standard deviations removes the units and gives the correlation,
$$
\mathrm{Corr}(Y, X) = \frac{\Cov{Y, X}}{\sqrt{\Var{Y}\,\Var{X}}} \in [-1, 1],
$$
where the bound is the Cauchy-Schwarz inequality.

- $Y$ and $X$ are **uncorrelated** if $\Cov{Y, X} = 0$, equivalently $\E{YX} = \E{Y}\,\E{X}$.

- Uncorrelatedness is a single restriction. It says the best straight-line predictor of $Y$ from $X$ under squared loss has zero slope, and it says nothing about any other way the two variables might be related.

## Some implications

- Under independence the expectation of a product of separate functions factorises:
$$
\begin{aligned}
\E{h_1(Y)\, h_2(X)}
&= \int\!\!\int h_1(y)\, h_2(x)\, \underbrace{f_{Y,X}(y, x)}_{=\, f_Y(y)\, f_X(x)}\, dy\, dx \\[2pt]
&= \int h_1(y)\, f_Y(y)\, dy \cdot \int h_2(x)\, f_X(x)\, dx \\[2pt]
&= \E{h_1(Y)}\, \E{h_2(X)}.
\end{aligned}
$$

- This holds for **every** pair of functions $h_1$ and $h_2$, not just for the identity.

- Taking $h_1(y) = y$ and $h_2(x) = x$ gives $\E{YX} = \E{Y}\E{X}$, so $\Cov{Y, X} = 0$: independence implies uncorrelatedness.

- The converse fails, and it fails for the reason above: uncorrelatedness is the factorisation for one pair of functions, whereas independence is the factorisation for all of them.

# Conditional expectation

## Conditional expectation: at a point

- Fix $x$ with $f_X(x) > 0$ and take the mean of the conditional density:
$$
\E{Y \mid X = x} = \int_{-\infty}^{\infty} y\, f_{Y \mid X}(y \mid x)\, dy.
$$
This is a **number**.

- Repeat for every $x$ with $f_X(x) > 0$. The result is a function of $x$,
$$
x \longmapsto \E{Y \mid X = x} =: g(x),
$$
the **conditional expectation function**.

## Conditional expectation: as a random variable

- Now plug the random variable $X$ into $g$:
$$
\E{Y \mid X} := g(X) = \int_{-\infty}^{\infty} y\, f_{Y \mid X}(y \mid X)\, dy.
$$
This is a **random variable**: a different draw of $X$ gives a different value.

- The distinction matters for every proof that follows. $\E{Y \mid X = x}$ is a number indexed by $x$; $\E{Y \mid X}$ is the random variable obtained by evaluating that function at $X$.

- The same construction applied to the squared deviation gives the **conditional variance**,
$$
\Var{Y \mid X} = \E{\big(Y - \E{Y \mid X}\big)^2 \,\Big|\, X},
$$
again a random variable, and a function of $X$.

- Since $\E{Y \mid X}$ is random, it has a mean of its own. What is it?

## Law of iterated expectations

- **Claim (LIE).**
$$
\E{\E{Y \mid X}} = \E{Y}.
$$

- Averaging the conditional mean over the distribution of $X$ gives back the unconditional mean.

- Read from right to left, it says any expectation can be computed in two stages: first conditional on $X$, then over $X$. Most proofs in this course use it that way.

## Law of iterated expectations: proof

- Write the outer expectation as an integral against $f_X$, then unpack $g$:
$$
\begin{aligned}
\E{\E{Y \mid X}} &= \int_{-\infty}^{\infty} \underbrace{\E{Y \mid X = x}}_{g(x)}\, f_X(x)\, dx \\
&= \int_{-\infty}^{\infty} \left[ \int_{-\infty}^{\infty} y\, f_{Y \mid X}(y \mid x)\, dy \right] f_X(x)\, dx \\
&= \int_{-\infty}^{\infty} \int_{-\infty}^{\infty} y\, \underbrace{f_{Y \mid X}(y \mid x)\, f_X(x)}_{=\, f_{Y,X}(y,x)}\, dy\, dx \\
&= \int_{-\infty}^{\infty} y \underbrace{\left[ \int_{-\infty}^{\infty} f_{Y,X}(y, x)\, dx \right]}_{=\, f_Y(y)} dy \\
&= \int_{-\infty}^{\infty} y\, f_Y(y)\, dy = \E{Y}.
\end{aligned}
$$

- The step from the third line to the fourth changes the order of integration.

## Taking out what is known

- Conditional on $X$, any function of $X$ is a constant and comes out of the expectation:
$$
\begin{aligned}
\E{Y \cdot h(X) \mid X} &= \int_{-\infty}^{\infty} y\, h(X)\, f_{Y \mid X}(y \mid X)\, dy \\
&= h(X) \int_{-\infty}^{\infty} y\, f_{Y \mid X}(y \mid X)\, dy = h(X)\, \E{Y \mid X}.
\end{aligned}
$$

- Two special cases used constantly:
$$
\E{h(X) \mid X} = h(X), \qquad \E{c \mid X} = c.
$$


## Question

- Let $Y$, $X$, $Z$ be random variables. Which of the following equals $\E{\E{Y \mid X, Z} \mid Z}$?
  1. $\E{Y}$
  2. $\E{Y \mid X}$
  3. $\E{Y \mid Z}$
  4. $0$


## Mean independence

- $Y$ is **mean independent** of $X$ if the conditional mean does not move with the conditioning value:
$$
\E{Y \mid X = x} = c \quad \text{for all } x \text{ with } f_X(x) > 0,
$$
where the constant $c$ does not depend on $x$.

- **Question.** The constant $c$ does not depend on $x$. What must it be?
  1. $0$
  2. $\E{X}$
  3. $\E{Y}$
  4. $42$

- Mean independence is weaker than independence:
  - Under mean independence, $\E{Y \mid X}$ does not depend on $X$, but $\E{Y^2\mid X}$, $\Var{Y \mid X}$, etc., can change with $X$.
  - Under independence, the whole conditional distribution is constant in $X$, so $\Var{Y \mid X}$, etc., are constant too.

## A zero conditional mean

- Suppose
$$
\E{U \mid X} = 0 .
$$
By the previous slide this is mean independence of $U$ from $X$, in the case where the constant is zero.

- Two consequences, by the LIE and by taking out what is known:
$$
\begin{aligned}
\E{U} &= \E{\underbrace{\E{U \mid X}}_{=\,0}} = 0, \\[2pt]
\E{U X} &= \E{\E{U X \mid X}} = \E{X \underbrace{\E{U \mid X}}_{=\,0}} = 0.
\end{aligned}
$$
The second line takes $X$ out of the inner expectation, which is allowed because $X$ is known once we condition on it.

- Putting the two together,
$$
\Cov{U, X} = \E{UX} - \E{U}\,\E{X} = 0 - 0 = 0 ,
$$
so $U$ and $X$ are uncorrelated.

## Three concepts of unrelatedness

- Ordered from strongest to weakest:
$$
\text{independence} \;\Longrightarrow\; \E{U \mid X} = \E{U} \;\Longrightarrow\; \Cov{U, X} = 0,
$$
and neither arrow reverses.

- Mean independence allows $\Var{U \mid X}$ to depend on $X$; independence does not. A variable can have a conditional mean that never moves with $X$ while its conditional variance does.

- Uncorrelatedness is weaker still: $\Cov{U, X} = 0$ is a single restriction, and it is compatible with a conditional mean $\E{U \mid X}$ that varies with $X$, because $\Cov{U, X} = \Cov{\E{U \mid X}, X}$.

## The conditional mean minimises mean squared error

- Among all functions $m$ of $X$, the conditional mean is the best predictor of $Y$ under squared loss. Condition on $X$ first:
$$
\begin{aligned}
\E{(Y - m(X))^2 \mid X}
&= \E{\big(Y - \E{Y \mid X}\big)^2 \mid X} \\
&\quad + \big(\E{Y \mid X} - m(X)\big)^2,
\end{aligned}
$$
because the cross term is $2\,(\E{Y\mid X} - m(X))\,\E{Y - \E{Y \mid X} \mid X} = 0$.

- Now average over $X$ with the LIE:
$$
\E{(Y - m(X))^2} \ge \E{\big(Y - \E{Y \mid X}\big)^2},
$$
with equality only when $m(X) = \E{Y \mid X}$ with probability one.

- This is the reason regression is about $\E{Y \mid X}$ and not some other feature of the conditional distribution. Change the loss and a different feature wins.

# The linear regression model

## The object of interest

- The relationship between the regressors and the dependent variable is summarised by the conditional expectation function:
$$
\E{Y_i \mid X_i} = \E{Y_i \mid X_{i1}, \ldots, X_{ik}},
$$
for instance
$$
\E{\ln \text{Wage}_i \mid \text{Education}_i, \text{Experience}_i, \text{Race}_i, \text{Gender}_i, \ldots}.
$$

- The practical problem: with data $\{(Y_i, X_i) : i = 1, \ldots, n\}$, estimate $\E{Y_i \mid X_i}$.

- Without restrictions this is a function of $k$ arguments to be learned from $n$ points. The **parametric** approach assumes the functional form and leaves a finite number of unknown constants.

## The linear regression model

- **Assumption.** The conditional expectation is linear in the parameters:
$$
\E{Y_i \mid X_i} = \beta_1 X_{i1} + \cdots + \beta_k X_{ik} = X_i^\top \beta,
$$
where $X_i = (X_{i1}, \ldots, X_{ik})^\top$ and $\beta = (\beta_1, \ldots, \beta_k)^\top \in \mathbb{R}^k$.

- Write $\E{Y_i \mid X_i = x} = x^\top \beta = \sum_{l=1}^{k} \beta_l x_l$. Differentiating in $x_j$,
$$
\frac{\partial \E{Y_i \mid X_i = x}}{\partial x_j}
= \frac{\partial (\beta_1 x_1 + \cdots + \beta_k x_k)}{\partial x_j}
= \beta_j .
$$

- So $\beta_j$ is the change in the conditional mean of $Y_i$ per unit change in $X_{ij}$, holding the other regressors fixed: a **marginal effect**.

- "Linear" refers to $\beta$. The regressors can be anything: squares, logs, products, indicator variables.

## The error term

- **Define** the error as the deviation from the conditional mean:
$$
U_i := Y_i - \E{Y_i \mid X_i} = Y_i - X_i^\top \beta.
$$

- Then, by construction,
$$
\begin{aligned}
Y_i &= X_i^\top \beta + U_i, \\
\E{U_i \mid X_i} &= \underbrace{\E{Y_i \mid X_i}}_{=\, X_i^\top\beta} - \underbrace{\E{X_i^\top \beta \mid X_i}}_{=\, X_i^\top\beta} = 0.
\end{aligned}
$$

- $U_i$ is not observable: the conditional expectation function is unknown, so the deviation from it is unknown.

- If the pairs $(Y_i, X_i)$ are iid, then so are the $U_i$, since each is the same function of its own pair.

## Two ways to write the same model

- Either start from the conditional mean and define the error,
$$
\left.\begin{aligned}
\E{Y_i \mid X_i} &= X_i^\top \beta \\
U_i &:= Y_i - X_i^\top \beta
\end{aligned}\right\}
\quad\Longrightarrow\quad
\E{U_i \mid X_i} = 0,
$$

- or start from the equation and the restriction on the error,
$$
\left.\begin{aligned}
Y_i &= X_i^\top \beta + U_i \\
\E{U_i \mid X_i} &= 0
\end{aligned}\right\}
\quad\Longrightarrow\quad
\E{Y_i \mid X_i} = X_i^\top \beta.
$$


## The intercept

- Consider an example with a single regressor $X_i^\ast\in \mathbb{R}$.
- Suppose $Y_i = \beta^\ast X_i^\ast + U_i^\ast$ with $\E{U_i^\ast\mid X_i^\ast} = \alpha \neq 0$. 
- We can add and subtract $\alpha$ to get:
$$
Y_i = \alpha + \beta^\ast X_i^\ast + \underbrace{(U_i^\ast - \alpha)}_{U_i}, \qquad \E{U_i\mid X_i^\ast} = 0.
$$

- Define vectors $X_i= (1, X_i^\ast)^\top$ and $\beta = (\alpha, \beta^\ast)^\top$. Then 
$$
Y_i = X_i^{\top} \beta + U_i, \qquad \E{U_i\mid X_i} = \E{U_i\mid 1, X_i^\ast}= 0,
$$
as conditioning on $1$ in addition to $X_i^\ast$ does not change the conditional expectation.

- Here $\alpha$ is the intercept: a coefficient on a constant regressor (i.e., a regressor equal to $1$ for all observations $i$).
- We include the intercept to absorb constant but non-zero conditional means of the error. This does not change the slope coefficients. 
- However, if the conditional mean of the error is not constant, i.e., $\E{U_i^\ast\mid X_i^\ast} = \alpha+\gamma X_i^\ast$, then 
$$
Y_i = \alpha + (\beta^\ast+\gamma) X_i^\ast + \underbrace{(U_i^\ast - \alpha-\gamma X_i^\ast)}_{U_i}, \qquad \E{U_i\mid X_i^\ast} = 0,
$$
and the slope coefficient is affected.


# Identification

## Identification: the question
- Consider the model $Y_i = X_i^\top \beta + U_i$ with $\E{U_i \mid X_i} = 0$:
  - $Y_i,X_i$ are observed,
  - $U_i$ is unobserved,
  - $\beta\in\mathbb R^k$ is unknown. 
- Suppose the joint distribution of the observables $F_{Y, X_1, \ldots, X_k}(y, x_1, \ldots, x_k)$
is **known** (we are at the population level). You have no information about $U_i$ beyond $\E{U_i \mid X_i} = 0$.

- Can you recover $\beta$?

- If the answer is no with the whole distribution in hand, no amount of data will help, as the joint distribution of the data does not pin down $\beta$.

## Identification: the definition

- **Definition.** $\beta$ is **identified** if it is uniquely determined by the joint distribution of the observed variables $(Y_i, X_{i1}, \ldots, X_{ik})$.

- Two things can go wrong:
  - the distribution of the observables is consistent with more than one value of $\beta$;
  - it is consistent with none, because the model is wrong.

- We ask: what does $\E{U_i \mid X_i} = 0$ let us compute from the distribution of $(Y_i, X_i)$, and does it pin $\beta$ down?

## What the model gives us

- By $\E{U_i \mid X_i} = 0$ and the LIE, $\E{X_i U_i}=0$.
- Rewrite $U_i$ in terms of the observables and $\beta$:
  $$
  U_i = Y_i - X_i^\top \beta
  $$
- Obtain the **population normal equations**:
$$
\boxed{\ \E{X_i \big(Y_i - X_i^\top \beta\big)} = 0\ }.
$$

- Multiply out, one step at a time:
$$
\begin{aligned}
0 = \E{X_i U_i}
&= \E{X_i \big(Y_i - X_i^\top \beta\big)} \\
&= \E{X_i Y_i - X_i X_i^\top \beta} \\
&= \E{X_i Y_i} - \E{X_i X_i^\top \beta} \\
&= \E{X_i Y_i} - \E{X_i X_i^\top}\,\beta ,
\end{aligned}
$$
the last step because $\beta$ is a vector of constants and comes out of the expectation.

- Rearranging,
$$
\E{X_i X_i^\top}\,\beta = \E{X_i Y_i}.
$$

- Knowns: $\E{X_i Y_i}$ ($k\times1$) and $\E{X_i X_i^\top}$ ($k \times k$), both features of the distribution of the observables. Unknown: $\beta$.

- If $\E{X_i X_i^\top}$ is invertible, the system has exactly one solution. Whether it is invertible is the whole question.

## The second-moment matrix

- Written out,
$$
\E{X_i X_i^\top}
= \mathrm{E}\begin{bmatrix}
X_{i1}^2 & X_{i1} X_{i2} & \cdots & X_{i1} X_{ik} \\
X_{i2} X_{i1} & X_{i2}^2 & \cdots & X_{i2} X_{ik} \\
\vdots & \vdots & \ddots & \vdots \\
X_{ik} X_{i1} & X_{ik} X_{i2} & \cdots & X_{ik}^2
\end{bmatrix}.
$$
- It is **symmetric**: using $(AB)^\top = B^\top A^\top$,
$$
\big(\E{X_i X_i^\top}\big)^\top = \E{(X_i X_i^\top)^\top} = \E{X_i X_i^\top}.
$$



## The second-moment matrix is positive semidefinite

- **Definitions.** A $k\times k$ matrix $A$ is **positive semidefinite** (p.s.d.) if $c^\top A c \ge 0$ for all $c \in \mathbb{R}^k$, and **positive definite** (p.d.) if $c^\top A c > 0$ for all $c \neq 0$.

- Take any $c \in \mathbb{R}^k$ and define the scalar
$$
W_i = X_i^\top c = c_1 X_{i1} + \cdots + c_k X_{ik}, \qquad c^\top X_i = (X_i^\top c)^\top = W_i.
$$

- Then the quadratic form is the expectation of a square:
$$
c^\top\, \E{X_i X_i^\top}\, c = \E{c^\top X_i X_i^\top c} = \E{\underbrace{W_i^2}_{\ge 0}} \ge 0.
$$

- So $\E{X_i X_i^\top}$ is p.s.d. whatever the distribution of the regressors. Positive **definiteness** is the extra step, and it can fail.

## When positive definiteness fails

- Suppose some $\tilde c \neq 0$ gives $\tilde c^\top \E{X_i X_i^\top}\, \tilde c = 0$. With $\tilde W_i = X_i^\top \tilde c$,
$$
\begin{aligned}
0 = \tilde c^\top \E{X_i X_i^\top}\, \tilde c = \E{\tilde W_i^2}
&\;\Longleftrightarrow\; \tilde W_i = 0 \text{ with probability } 1 \\
&\;\Longleftrightarrow\; \boxed{\ \tilde c_1 X_{i1} + \cdots + \tilde c_k X_{ik} = 0\ }
\end{aligned}
$$

- With $\tilde c \ne 0$, say $\tilde c_1 \neq 0$, one regressor equals, with probability one, an exact linear combination of the others:
$$
X_{i1} = -\frac{\tilde c_2}{\tilde c_1} X_{i2} - \cdots - \frac{\tilde c_k}{\tilde c_1} X_{ik}.
$$
This is **perfect multicollinearity**.

## Positive definiteness and the rank condition

- Putting the two directions together,
$$
\begin{aligned}
\E{X_i X_i^\top} \text{ is p.d.}
&\;\Longleftrightarrow\;
X_i^\top c = 0 \text{ with probability } 1 \text{ only if } c = 0 \\
&\;\Longleftrightarrow\;
\operatorname{rank}\big(\E{X_i X_i^\top}\big) = k.
\end{aligned}
$$

- The relation has to be exact and linear. A relation $c_1 X_{i1} + c_2 X_{i2} = 0$ holding with probability one for some $(c_1, c_2) \neq (0,0)$ is ruled out; $X_{i2} = X_{i1}^2$ is not linear and is fine, unless $X_{i1}$ is binary, when $X_{i1}^2 = X_{i1}$.

## Identification: the result

- **Assume**
  1. $Y_i = X_i^\top \beta + U_i$ with $\beta \in \mathbb{R}^k$;
  2. $\E{U_i \mid X_i} = 0$;
  3. No perfect multicollinearity: $\operatorname{rank}\big(\E{X_i X_i^\top}\big) = k$.

- Then $\beta$ is identified, and
$$
\boxed{\ \beta = \big(\E{X_i X_i^\top}\big)^{-1}\, \E{X_i Y_i}\ }
$$

- Assumptions 1 and 2 give the population normal equations $\E{X_i (Y_i - X_i^\top \beta)} = 0$, that is $\E{X_i X_i^\top}\beta = \E{X_i Y_i}$; assumption 3 makes the matrix invertible.

- With a single regressor and no intercept ($k = 1$),
$$
\E{X_i Y_i} = \E{X_i^2}\,\beta \quad\Longrightarrow\quad \beta = \frac{\E{X_i Y_i}}{\E{X_i^2}}.
$$

## Identification under a weaker assumption

- Replace $\E{U_i \mid X_i} = 0$ with a weaker condition:
$$
\E{X_i U_i} = 0.
$$

- Assume $Y_i= X_i^\top \beta + U_i$ and $\operatorname{rank}\big(\E{X_i X_i^\top}\big) = k$. 
- Then we can proceed as before: $\beta = (\E{X_i X_i^\top})^{-1}\E{X_i Y_i}$, so $\beta$ is still identified.

- What changes is the meaning of $\beta$. Without $\E{U_i \mid X_i} = 0$ we can no longer conclude that $X_i^\top \beta$ is the conditional mean, since in general
  $$
  \E{Y_i \mid X_i} = X_i^\top \beta + \E{U_i \mid X_i}.
  $$

# Estimation and least squares

## From identification to an estimator

- Identification wrote $\beta$ as a function of population moments:
$$
\beta = \big(\E{X_i X_i^\top}\big)^{-1}\, \E{X_i Y_i}.
$$

- Everything below assumes the sample counterpart of the rank condition, $\operatorname{rank}(X) = k$, so that the inverse exists. We return to it once the matrix notation is in place.

- **Method of moments.** Replace each expectation $\E{\cdot}$ by the sample average $\frac{1}{n}\sum_{i=1}^n (\cdot)$:
$$
\hat\beta = \left( \frac{1}{n}\sum_{i=1}^{n} X_i X_i^\top \right)^{-1} \frac{1}{n}\sum_{i=1}^{n} X_i Y_i
= \left( \sum_{i=1}^{n} X_i X_i^\top \right)^{-1} \sum_{i=1}^{n} X_i Y_i.
$$
The two factors of $\frac{1}{n}$ cancel.

- An **estimator** is any function of the sample $\{(Y_i, X_i)\}$. It may not depend on $U_i$ or on $\beta$, which are unobserved.

## The same estimator from the moment condition

- Alternatively, start from $\E{X_i (Y_i - X_i^\top \beta)} = 0$ and replace the expectation by an average, evaluated at the estimator:
$$
\frac{1}{n} \sum_{i=1}^{n} X_i \big(\underbrace{Y_i - X_i^\top \hat\beta}_{=:\ \hat U_i}\big) = 0
\quad\Longrightarrow\quad
\sum_{i=1}^{n} X_i X_i^\top\, \hat\beta = \sum_{i=1}^{n} X_i Y_i,
$$
which is the same $\hat\beta$.

- $\hat U_i = Y_i - X_i^\top \hat\beta$ is the **residual**: the sample counterpart of the error $U_i = Y_i - X_i^\top \beta$. Unlike $U_i$, it is computable.

- $\hat\beta$ is built to make the residuals satisfy the sample version of $\E{X_i U_i} = 0$.

## The estimator is not the parameter

- Sample averages are random; expectations are fixed numbers. In general
$$
\frac{1}{n}\sum_{i=1}^{n} X_i X_i^\top \neq \E{X_i X_i^\top},
\qquad
\frac{1}{n}\sum_{i=1}^{n} X_i Y_i \neq \E{X_i Y_i},
$$
and so in general $\hat\beta \neq \beta$.

- Flip a fair coin ten times: the sample proportion of heads is rarely one half, although the population proportion is exactly one half.

- If the errors $U_1, \ldots, U_n$ are continuously distributed, $\mathrm{P}(\hat\beta = \beta) = 0$.

- $\hat\beta$ is a random vector: a different sample gives a different value. It has a distribution, and the next lecture is about where that distribution sits relative to $\beta$ and how spread out it is.

## Sample orthogonality

- Since $\hat\beta$ solves $\sum_i X_i (Y_i - X_i^\top \hat\beta) = 0$,
$$
\sum_{i=1}^{n} X_i \hat U_i = 0
\qquad\Longleftrightarrow\qquad
\begin{pmatrix} \sum_{i=1}^{n} X_{i1} \hat U_i \\ \vdots \\ \sum_{i=1}^{n} X_{ik} \hat U_i \end{pmatrix}
= \begin{pmatrix} 0 \\ \vdots \\ 0 \end{pmatrix}.
$$

- One equation per regressor: each regressor is orthogonal to the residual vector,
$$
\begin{pmatrix} X_{1j} \\ \vdots \\ X_{nj} \end{pmatrix}^{\top}
\begin{pmatrix} \hat U_1 \\ \vdots \\ \hat U_n \end{pmatrix} = \sum_{i=1}^{n} X_{ij}\, \hat U_i = 0, \qquad j = 1, \ldots, k.
$$

- If the first regressor is the constant, $X_{i1} = 1$, the first equation says $\sum_{i=1}^n \hat U_i = 0$: the residuals average to zero.

## Matrix notation

- Stack the $n$ observations:
$$
\underset{n\times1}{Y} = \begin{pmatrix} Y_1 \\ \vdots \\ Y_n \end{pmatrix},\qquad
\underset{n\times1}{U} = \begin{pmatrix} U_1 \\ \vdots \\ U_n \end{pmatrix},\qquad
\underset{n\times k}{X} = \begin{pmatrix} X_1^\top \\ \vdots \\ X_n^\top \end{pmatrix}.
$$

- Row $i$ of $X$ is the transpose of the regressor vector $X_i$; column $j$ of $X$ holds regressor $j$ for all $n$ observations:
$$
X = \begin{pmatrix} X_{11} & X_{12} & \cdots & X_{1k} \\ \vdots & \vdots & & \vdots \\ X_{n1} & X_{n2} & \cdots & X_{nk} \end{pmatrix}.
$$

- The $n$ equations $Y_i = X_i^\top \beta + U_i$ become one:
$$
Y = \begin{pmatrix} X_1^\top \beta \\ \vdots \\ X_n^\top \beta \end{pmatrix} + \begin{pmatrix} U_1 \\ \vdots \\ U_n \end{pmatrix}
\qquad\Longleftrightarrow\qquad
\underset{n\times1}{Y} = \underset{n\times k}{X}\ \underset{k\times1}{\beta} + \underset{n\times1}{U}.
$$

## Two identities

- **Claim.**
$$
X^\top Y = \sum_{i=1}^{n} X_i Y_i \quad (k \times 1),
\qquad
X^\top X = \sum_{i=1}^{n} X_i X_i^\top \quad (k \times k).
$$

- **Proof of the second.** Transposing the stack of rows gives a row of columns:
$$
\begin{aligned}
X^\top X &= \begin{pmatrix} X_1^\top \\ \vdots \\ X_n^\top \end{pmatrix}^{\top} \begin{pmatrix} X_1^\top \\ \vdots \\ X_n^\top \end{pmatrix}
= \begin{pmatrix} X_1 & \cdots & X_n \end{pmatrix} \begin{pmatrix} X_1^\top \\ \vdots \\ X_n^\top \end{pmatrix} \\
&= \underbrace{X_1 X_1^\top}_{k\times k} + \cdots + X_n X_n^\top = \sum_{i=1}^{n} X_i X_i^\top.
\end{aligned}
$$
The first identity is the same computation with the second factor $X$ replaced by $Y$, whose $i$th row is the scalar $Y_i$.

## The estimator in matrix form

- Substitute the two identities into the method-of-moments formula:
$$
\begin{aligned}
\hat\beta &= \left( \sum_{i=1}^{n} X_i X_i^\top \right)^{-1} \sum_{i=1}^{n} X_i Y_i \\
&= \left( \frac{1}{n} X^\top X \right)^{-1} \left( \frac{1}{n} X^\top Y \right).
\end{aligned}
$$

- The $\frac{1}{n}$ factors cancel:
$$
\boxed{\ \hat\beta = (X^\top X)^{-1} X^\top Y\ }
$$

- $X^\top X$ is $k \times k$ and $X^\top Y$ is $k \times 1$, so $\hat\beta$ is $k \times 1$, as it should be.

## Residuals in matrix form

- Residual vector and its defining property:
$$
\begin{aligned}
\hat U &= Y - X \hat\beta, \\
X^\top \hat U &= X^\top (Y - X\hat\beta) = X^\top Y - X^\top X\, \underbrace{(X^\top X)^{-1} X^\top Y}_{\hat\beta} \\
&= X^\top Y - X^\top Y = 0.
\end{aligned}
$$

- These are the **sample normal equations**, $X^\top (Y - X\hat\beta) = X^\top \hat U = 0$, the sample counterpart of $\E{X_i (Y_i - X_i^\top \beta)} = 0$. They are the same $k$ equations as before:
$$
X^\top \hat U = \sum_{i=1}^{n} X_i \hat U_i = \begin{pmatrix} \sum_{i=1}^{n} X_{i1} \hat U_i \\ \vdots \\ \sum_{i=1}^{n} X_{ik} \hat U_i \end{pmatrix}.
$$

- Population: $\E{X_i U_i} = 0$. Sample: $\frac{1}{n} \sum_{i=1}^n X_i \hat U_i = 0$. The estimator is the value of the coefficient vector that makes the sample mimic the population.

## The sample rank condition

- $(X^\top X)^{-1}$ exists exactly when the columns of $X$ are linearly independent:
$$
X c = c_1 \begin{pmatrix} X_{11} \\ \vdots \\ X_{n1} \end{pmatrix} + \cdots + c_k \begin{pmatrix} X_{1k} \\ \vdots \\ X_{nk} \end{pmatrix} = 0 \quad\text{only if}\quad c = 0,
$$
that is, $\operatorname{rank}(X) = k$. In particular $n \ge k$: at least as many observations as coefficients.

- If it fails, $\operatorname{rank}(X^\top X) = \operatorname{rank}(X) < k$, and the normal equations have infinitely many solutions.

- This is the sample counterpart of $\operatorname{rank}\big(\E{X_i X_i^\top}\big) = k$.

## Least squares

- Take a candidate value $b \in \mathbb{R}^k$ and measure how badly the fitted values $X_i^\top b$ fit the data by the **sum of squared residuals**:
$$
\begin{aligned}
S(b) = \sum_{i=1}^{n} \big(Y_i - X_i^\top b\big)^2
&= \begin{pmatrix} Y_1 - X_1^\top b \\ \vdots \\ Y_n - X_n^\top b \end{pmatrix}^{\top}
\begin{pmatrix} Y_1 - X_1^\top b \\ \vdots \\ Y_n - X_n^\top b \end{pmatrix} \\
&= (Y - Xb)^\top (Y - Xb) = \|Y - Xb\|^2,
\end{aligned}
$$
where for $a \in \mathbb{R}^n$, $\|a\|^2 = a^\top a = \sum_{i=1}^n a_i^2$.

- The **ordinary least squares** estimator picks the $b$ that fits best:
$$
\hat\beta^{\,\text{OLS}} = \underset{b \in \mathbb{R}^k}{\arg\min}\ \sum_{i=1}^{n} \big(Y_i - X_i^\top b\big)^2
= \underset{b \in \mathbb{R}^k}{\arg\min}\ \|Y - Xb\|^2.
$$

- **Claim.** $\hat\beta^{\,\text{OLS}} = \hat\beta = (X^\top X)^{-1} X^\top Y$: least squares and the method of moments deliver the same estimator.

## Proof: completing the square

- Add and subtract $X\hat\beta$ inside the criterion, and use $(u + v)^\top (u + v) = u^\top u + v^\top v + 2 u^\top v$:
$$
\begin{aligned}
(Y - Xb)^\top (Y - Xb)
&= \big(\underbrace{Y - X\hat\beta}_{\hat U} + X(\hat\beta - b)\big)^\top \big(\hat U + X(\hat\beta - b)\big) \\
&= \hat U^\top \hat U + (\hat\beta - b)^\top X^\top X (\hat\beta - b) \\
&\qquad + 2\, \underbrace{\hat U^\top X}_{=\,(X^\top \hat U)^\top =\, 0}\, (\hat\beta - b) \\
&= \underbrace{\hat U^\top \hat U}_{\text{free of } b} + (\hat\beta - b)^\top X^\top X (\hat\beta - b).
\end{aligned}
$$

- The cross term vanishes by the normal equations $X^\top \hat U = 0$. That is the only place the definition of $\hat\beta$ enters.

- So minimising $S(b)$ over $b$ is the same as minimising the quadratic form $(\hat\beta - b)^\top X^\top X (\hat\beta - b)$.

## Proof: the quadratic form

- For any $a \in \mathbb{R}^k$, with $z = Xa \in \mathbb{R}^n$,
$$
a^\top X^\top X a = (Xa)^\top (Xa) = z^\top z = \sum_{i=1}^{n} z_i^2 \ge 0,
$$
so $X^\top X$ is p.s.d. It is also symmetric, since $(X^\top X)^\top = X^\top X$.

- Under the rank condition, $Xa = 0$ only if $a = 0$, so $\sum_i z_i^2 = 0$ only if $a = 0$: $X^\top X$ is **positive definite**.

- Apply this with $a = \hat\beta - b$:
$$
S(b) = S(\hat\beta) + \underbrace{(\hat\beta - b)^\top X^\top X (\hat\beta - b)}_{\ge\, 0,\ \text{and}\ =\,0 \text{ iff } b = \hat\beta} \;\ge\; S(\hat\beta).
$$

- Hence $\hat\beta = (X^\top X)^{-1} X^\top Y$ is the unique minimiser of the sum of squared residuals: it **is** the OLS estimator.

## The same result from the first-order condition

- Expand the criterion, $S(b) = Y^\top Y - 2\, b^\top X^\top Y + b^\top X^\top X b$, and use $\partial (x^\top c)/\partial x = c$ together with $\partial (x^\top A x)/\partial x = 2Ax$ for symmetric $A$:
$$
\frac{\partial S(b)}{\partial b} = -2 X^\top Y + 2 X^\top X b = 0
\quad\Longrightarrow\quad
b = (X^\top X)^{-1} X^\top Y.
$$

- The Hessian $\partial^2 S / \partial b\, \partial b^\top = 2 X^\top X$ is positive definite under the rank condition, so this is a minimum.

- The first-order condition, rewritten as $X^\top (Y - Xb) = 0$, **is** the normal equations. Least squares and the method of moments are two routes to one estimator.

## Population least squares

- The population has its own least-squares problem, and $\beta$ solves it.

- **Claim.** Suppose $Y_i = X_i^\top \beta + U_i$ with $\E{U_i \mid X_i} = 0$, and $\operatorname{rank}\big(\E{X_i X_i^\top}\big) = k$. Then
$$
\beta = \underset{b \in \mathbb{R}^k}{\arg\min}\ \E{(Y_i - X_i^\top b)^2},
$$
and the minimiser is unique.

- So OLS is the sample analogue of $\beta$ in two ways: replace population moments with sample moments, or replace expected squared error with average squared error.

## Population least squares: proof

- Condition on $X_i$, use that the conditional mean minimises conditional mean squared error, then apply the LIE:
$$
\begin{aligned}
\E{(Y_i - X_i^\top b)^2}
&= \E{\,\E{(Y_i - X_i^\top b)^2 \mid X_i}\,} \\
&\ge \E{\,\E{(Y_i - \underbrace{\E{Y_i \mid X_i}}_{X_i^\top \beta})^2 \mid X_i}\,} \\
&= \E{(Y_i - X_i^\top \beta)^2}.
\end{aligned}
$$

- Equality at some $b^\ast$ requires $X_i^\top b^\ast = \E{Y_i \mid X_i} = X_i^\top \beta$ with probability one, so $X_i^\top (b^\ast - \beta) = 0$ with probability one, and with no multicollinearity $b^\ast = \beta$.

## Other linear estimators, for contrast

- Nothing forces the weights in the normal equations to be equal across observations. For a known symmetric positive definite $n \times n$ matrix $W$,
$$
\tilde\beta = (X^\top W X)^{-1} X^\top W Y
$$
is also a function of the sample, and $W = I_n$ gives back $\hat\beta$.

- In general $\tilde\beta$ minimises $(Y - Xb)^\top W (Y - Xb)$. With $W$ diagonal, $W = \operatorname{diag}(w_1, \ldots, w_n)$ and $w_i > 0$, this is **weighted least squares**: it minimises $\sum_i w_i (Y_i - X_i^\top b)^2$. Taking $W$ to be the inverse of the variance matrix of the errors gives **generalised least squares**.

- Which of these is a good estimator, and in what sense, is the subject of the next lecture: bias, variance, and the Gauss-Markov theorem.

## Summary: the model and identification

- The object of interest is the conditional expectation $\E{Y_i \mid X_i}$; the linear model assumes $\E{Y_i \mid X_i} = X_i^\top \beta$, which is the same as $Y_i = X_i^\top \beta + U_i$ with $\E{U_i \mid X_i} = 0$.

- Probability supplied what was needed on the way: densities and conditional densities, the law of iterated expectations, taking out what is known, and mean independence.

- **Identification.** $\E{U_i \mid X_i} = 0$ gives $\E{X_i U_i} = 0$, hence the population normal equations $\E{X_i (Y_i - X_i^\top \beta)} = 0$, that is $\E{X_i X_i^\top}\beta = \E{X_i Y_i}$. If $\E{X_i X_i^\top}$ is positive definite, which means no exact linear relation among the regressors, then
$$
\beta = \big(\E{X_i X_i^\top}\big)^{-1} \E{X_i Y_i}.
$$

## Summary: the estimator

- **Estimation.** Under the sample rank condition $\operatorname{rank}(X) = k$, replacing expectations by sample averages gives
$$
\hat\beta = \left( \sum_{i=1}^{n} X_i X_i^\top \right)^{-1} \sum_{i=1}^{n} X_i Y_i = (X^\top X)^{-1} X^\top Y,
$$
which is also the unique minimiser of $\sum_i (Y_i - X_i^\top b)^2$: the OLS estimator.

- $\hat\beta$ solves the sample normal equations $X^\top (Y - X\hat\beta) = 0$, that is $X^\top \hat U = 0$: the sample counterpart of the population normal equations $\E{X_i (Y_i - X_i^\top \beta)} = 0$.
