---
# _quarto.yml in this folder sets two things for every deck: `output-dir: output`,
# which sends the rendered files to output/, and `embed-resources: true` on the
# page and on the slide show, which keeps both of them self-contained. Quarto
# finds that file by walking up from this one, so both settings hold wherever the
# render is started from. Do not repeat either setting here.
title: "Lecture 3: Geometry of OLS and partitioned regression"
subtitle: "Economics 527: Econometric Methods"
author:
  - name: "Vadim Marmer, UBC"
format:
  html:
    output-file: 527_03_ols_geometry.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: katex
  pdf:
    output-file: 527_03_ols_geometry.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
  revealjs:
    output-file: 527_03_ols_geometry_slides.html
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: katex
    navigation-mode: linear
    scrollable: true
    chalkboard: false
    transition: none
---

# Geometry of OLS

## Least squares as a distance

::: {.hidden}
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
\gdef\rank#1{\operatorname{rank}\left(#1\right)}
:::

```{r}
#| include: false
# Figures are drawn through tikzDevice, so every label is typeset by LaTeX
# itself and matches the mathematics in the slides. Needs tikzDevice, png and
# a working pdflatex and pdftoppm.
suppressPackageStartupMessages(library(tikzDevice))
# Keep the string-metrics cache out of the source folder.
options(tikzMetricsDictionary = file.path(tempdir(), "tikzMetrics"))

tikz_fig <- function(stem, width, height, expr, draw = TRUE) {
  d <- file.path(tempdir(), "tikzfig")
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
  tex <- file.path(d, paste0(stem, ".tex"))
  tikz(tex, width = width, height = height, pointsize = 11, standAlone = TRUE)
  expr
  dev.off()
  system2("pdflatex", c("-interaction=batchmode", "-halt-on-error",
                        paste0("-output-directory=", shQuote(d)), shQuote(tex)),
          stdout = FALSE, stderr = FALSE)
  system2("pdftoppm", c("-r", "300", "-png", "-singlefile",
                        shQuote(file.path(d, paste0(stem, ".pdf"))),
                        shQuote(file.path(d, stem))),
          stdout = FALSE, stderr = FALSE)
  img_path <- file.path(d, paste0(stem, ".png"))
  if (!draw) return(img_path)
  img <- png::readPNG(img_path)
  op <- par(mar = rep(0, 4))
  plot.new(); plot.window(c(0, 1), c(0, 1), xaxs = "i", yaxs = "i")
  rasterImage(img, 0, 0, 1, 1, interpolate = TRUE)
  par(op)
  invisible(NULL)
}

# The picture the class drew for k = 1, n = 2: S(X) is the line through X and
# the fitted vector is the foot of the perpendicular dropped from Y.
# full = TRUE adds the coordinates and the angle; full = FALSE labels the two
# projections instead. Steps 0:5 add axes, X, S(X), Y, the perpendicular and
# fitted vector, then the residual label. The default is the complete picture.
proj_picture <- function(full, step = 5L) {
  X <- c(1.2, 0.6); Y <- c(1.9, 3.7)
  Yh <- X * sum(X * Y) / sum(X * X)
  d <- X / sqrt(sum(X^2)); p <- c(-d[2], d[1])
  red <- "#d7301f"; green <- "#1b9e77"; blue <- "#1f78b4"
  op <- par(mar = rep(0.1, 4))
  plot.new(); plot.window(xlim = c(-0.7, 5.3), ylim = c(-0.5, 4.3), asp = 1)
  arrows(-0.4, 0, 5.1, 0, length = 0.08)
  arrows(0, -0.4, 0, 4.2, length = 0.08)
  if (step >= 2L) {
    segments(-0.45 * d[1], -0.45 * d[2], 4.9 * d[1], 4.9 * d[2],
             col = red, lwd = 1.5)
    text(4.9 * d[1] + 0.1, 4.9 * d[2] + 0.25, "$S(X)$", col = red)
  }
  if (full) {
    for (j in which(c(step >= 1L, step >= 3L))) {
      v <- list(X, Y)[[j]]
      segments(v[1], v[2], v[1], 0, lty = 3)
      segments(v[1], v[2], 0, v[2], lty = 3)
    }
    if (step >= 4L) {
      segments(Yh[1], Yh[2], Yh[1], 0, lty = 3, col = green)
      segments(Yh[1], Yh[2], 0, Yh[2], lty = 3, col = green)
    }
    shown <- c(step >= 1L, step >= 3L)
    if (any(shown)) {
      text(c(X[1], Y[1])[shown], -0.25, c("$X_1$", "$Y_1$")[shown])
      text(-0.3, c(X[2], Y[2])[shown], c("$X_2$", "$Y_2$")[shown])
    }
    if (step >= 4L) {
      text(Yh[1], -0.25, "$\\hat Y_1$", col = green)
      text(-0.3, Yh[2], "$\\hat Y_2$", col = green)
    }
    if (step >= 3L) {
      a <- seq(atan2(d[2], d[1]), atan2(Y[2], Y[1]), length.out = 40)
      lines(0.9 * cos(a), 0.9 * sin(a))
      text(1.15 * cos(mean(a)), 1.15 * sin(mean(a)), "$\\theta$")
    }
  }
  if (step >= 4L)
    arrows(0, 0, Yh[1], Yh[2], col = green, lwd = 3, length = 0.1)
  if (step >= 1L)
    arrows(0, 0, X[1], X[2], lwd = 2, length = 0.1)
  if (step >= 3L)
    arrows(0, 0, Y[1], Y[2], lwd = 2, length = 0.1)
  if (step >= 4L) {
    segments(Yh[1], Yh[2], Y[1], Y[2], col = blue, lwd = 2,
             lty = if (full) 2 else 1)
    s <- 0.2
    lines(Yh[1] + s * c(-d[1], -d[1] + p[1], p[1]),
          Yh[2] + s * c(-d[2], -d[2] + p[2], p[2]), col = blue)
  }
  if (step >= 3L) text(Y[1] + 0.05, Y[2] + 0.22, "$Y$")
  if (step >= 1L) text(X[1] + 0.2, X[2] - 0.2, "$X$")
  m <- (Y + Yh) / 2
  if (full) {
    if (step >= 4L)
      text(Yh[1] + 0.25, Yh[2] - 0.2, "$\\hat Y$", col = green)
    if (step >= 5L)
      text(m[1] + 0.3, m[2] + 0.15, "$\\hat U$", col = blue)
  } else {
    if (step >= 4L)
      text(Yh[1] + 0.15, Yh[2] - 0.25, "$\\hat Y = P_X Y$", col = green, adj = 0)
    if (step >= 5L)
      text(m[1] + 0.15, m[2] + 0.1, "$M_X Y = \\hat U$", col = blue, adj = 0)
  }
  par(op)
}
```

- OLS, first lecture:
  $$
  \hat\beta = \underset{b \in \mathbb{R}^k}{\arg\min}\ \underbrace{(Y - Xb)^\top (Y - Xb)}_{=\, \|Y - Xb\|^2}
  $$

- Norm of $a \in \mathbb{R}^n$:
  $$
  \begin{aligned}
  \|a\| &= \sqrt{\sum_{i=1}^n a_i^2} \\
  &= \sqrt{a^\top a}
  \end{aligned}
  $$

## Column space $S(X)$

- Column space of $X$:
  $$
  S(X) = \big\{\, v \in \mathbb{R}^n : v = \underset{n\times k}{X}\, \underbrace{b}_{k\times 1} \ \text{ for some } b \in \mathbb{R}^k \,\big\}
  $$

- Partition $X$ column-wise:
  $$
  \underbrace{X}_{n\times k} = \big[\, \underbrace{\underline{X}_1}_{n\times 1} \ \cdots \ \underbrace{\underline{X}_k}_{n\times 1} \,\big], 
  $$
  where $\underline{X}_j$ is the $n \times 1$ vector of the observations on the $j$-th regressor.

- What the elements of $S(X)$ look like, in terms of the columns of $X$:
  $$
  \begin{aligned}
  \underbrace{X}_{n\times k}\underbrace{b}_{k\times 1} &= \underbrace{\big[\, \underline{X}_1 \ \cdots \ \underline{X}_k \,\big]}_{=X} \begin{bmatrix} b_1 \\ \vdots \\ b_k \end{bmatrix} \\
  &= b_1 \underline{X}_1 + \cdots + b_k \underline{X}_k \\
  &= \sum_{j=1}^{k} b_j \underline{X}_j
  \end{aligned}
  $$

- Each $\underline{X}_j \in S(X)$, since we can take $b_j = 1$, $b_i = 0$ for $i \ne j$:
  $$
  \begin{aligned}
  Xb &= \big[\, \underline{X}_1 \ \cdots \ \underline{X}_k \,\big] \begin{bmatrix} 0 \\ \vdots \\ 1 \\ \vdots \\ 0 \end{bmatrix} \\
  &= \underline{X}_j \in S(X)
  \end{aligned}
  $$

- $S(X) \subset \mathbb{R}^n$: $k$-dimensional subspace; $\rank{X} = k < n$.

## $\hat Y$: closest to $Y$ in $S(X)$

- We can re-state the OLS problem in terms of $S(X)$:
  $$
  \begin{aligned}
  & \hat\beta = \underset{b \in \mathbb{R}^k}{\arg\min}\ \|Y - Xb\|^2, \qquad \hat Y = X\hat\beta \\
  & \Longleftrightarrow \\
  \hat Y &= \underset{v \in S(X)}{\arg\min}\ \|Y - v\|^2
  \end{aligned}
  $$

- When $k = 1$, $n = 2$, we can visualize the solution: You can find the closest element by dropping a perpendicular from $Y$ to $S(X)$.

- $Y = (Y_1, Y_2)^\top$, $X = (X_1, X_2)^\top$, and $S(X)=\{(bX_1, bX_2)^\top : b \in \mathbb{R}\}$.

  ```{r}
  #| echo: false
  #| results: asis
  #| fig-align: center
  #| fig-width: 6
  #| fig-height: 4.6
  #| fig-dpi: 300
  if (identical(knitr::opts_knit$get("rmarkdown.pandoc.to"), "revealjs")) {
    # Cumulative, opaque frames share one grid cell, so advancing or going
    # back uses Reveal's normal fragment controls without shifting the figure.
    # Only this slide gets these styles; ordinary HTML/PDF uses one full plot.
    cat('```{=html}\n<style>
    .reveal .ols-projection-build {
      display: grid; width: 576px; max-width: 100%; margin: 1em auto;
    }
    .reveal .ols-projection-build > img {
      grid-area: 1 / 1; width: 100%; max-width: 100%; max-height: none;
      margin: 0; background: white; transition: none;
    }
    </style>\n<div class="ols-projection-build">\n')
    descriptions <- c("Coordinate axes", "Vector X and its coordinates",
                      "The span S(X)", "Vector Y and its coordinates",
                      "Perpendicular from Y to S(X) and fitted vector Y-hat",
                      "Complete projection diagram with residual U-hat")
    for (step in 0:5) {
      frame <- tikz_fig(paste0("geometry-step-", step), 6, 4.6, {
        proj_picture(full = TRUE, step = step)
      }, draw = FALSE)
      cat(sprintf('<img class="%s" data-projection-step="%d" src="%s" alt="%s">\n',
                  if (step == 0L) "" else "fragment", step,
                  knitr::image_uri(frame), descriptions[step + 1L]))
    }
    cat('</div>
    <script>
    document.addEventListener("DOMContentLoaded", function () {
      const diagram = document.querySelector(".ols-projection-build");
      const parentFragment = diagram.closest(".fragment");
      function keepDiagramVisible(event) {
        if (diagram.contains(event.fragment) ||
            (event.type === "fragmentshown" && event.fragment === parentFragment)) {
          diagram.scrollIntoView({ block: "center", behavior: "instant" });
        }
      }
      Reveal.on("fragmentshown", keepDiagramVisible);
      Reveal.on("fragmenthidden", keepDiagramVisible);
    });
    </script>\n```\n')
  } else {
    tikz_fig("geometry", 6, 4.6, {
      proj_picture(full = TRUE)
    })
  }
  ```

- The coefficient $\hat\beta$ is determined by the angle $\theta$ between $X$ and $Y$:
  $$
  \begin{aligned}
  \cos\theta &= \frac{X^\top Y}{\|X\|\,\|Y\|} \\
  \hat\beta &= \frac{X^\top Y}{\|X\|^2} \\
  &= \cos\theta \cdot \frac{\|Y\|}{\|X\|}
  \end{aligned}
  $$

- Residuals, perpendicular to $S(X)$:
  $$
  \begin{aligned}
  \hat U &= Y - \hat Y = Y - X\hat\beta
  \end{aligned}
  $$

## Dropping the perpendicular: the algebra

- Fitted values:
  $$
  \begin{aligned}
  \hat Y &= X\hat\beta \\
  &= \underbrace{X (X^\top X)^{-1} X^\top}_{P_X}\, Y \\
  &= P_X Y
  \end{aligned}
  $$

- Residuals:
  $$
  \begin{aligned}
  \hat U &= Y - \hat Y \\
  &= Y - X\hat\beta \\
  &= Y - X (X^\top X)^{-1} X^\top Y \\
  &= Y - P_X Y \\
  &= \underbrace{(I_n - P_X)}_{=\, M_X}\, Y \\
  &= M_X Y
  \end{aligned}
  $$

- Both $n \times n$:
  $$
  \boxed{\ P_X = \underset{n\times k}{X}\ \underset{k\times k}{(X^\top X)}^{-1} \underset{k\times n}{X}^\top \qquad M_X = \underset{n\times n}{I_n} - \underset{n\times n}{P_X}\ }
  $$

# Projection matrices

## $P_X Y$ and $M_X Y$

- For $Y \in \mathbb{R}^n$:
  $$
  \begin{aligned}
  P_X + M_X &= I_n \\
  \Longrightarrow\quad P_X Y + M_X Y &= Y
  \end{aligned}
  $$

  ```{r}
  #| echo: false
  #| fig-align: center
  #| fig-width: 6
  #| fig-height: 4.6
  #| fig-dpi: 300
  tikz_fig("projections", 6, 4.6, {
    proj_picture(full = FALSE)
  })
  ```

## Properties of $P_X$

- Symmetric, by $(AB)^\top = B^\top A^\top$:
  $$
  \begin{aligned}
  P_X^\top &= \big( X (X^\top X)^{-1} X^\top \big)^\top \\
  &= X (X^\top X)^{-1} X^\top \\
  &= P_X
  \end{aligned}
  $$

- $P_X X = X$:
  $$
  \begin{aligned}
  P_X X &= \underbrace{X (X^\top X)^{-1} X^\top}_{P_X}\, X \\
  &= X
  \end{aligned}
  $$

- Column by column, $P_X \underline{X}_j = \underline{X}_j$ for $j = 1, \ldots, k$:
  $$
  \begin{aligned}
  P_X X &= P_X \big[\, \underset{n\times 1}{\underline{X}_1} \ \cdots \ \underset{n\times 1}{\underline{X}_k} \,\big] \\
  &= \big[\, P_X \underline{X}_1 \ \cdots \ P_X \underline{X}_k \,\big] \\
  &= \big[\, \underline{X}_1 \ \cdots \ \underline{X}_k \,\big],
  \end{aligned}
  $$
  where the last equality holds because $P_X X=X$.

- Idempotent:
  $$
  \begin{aligned}
  P_X^2 &= P_X P_X \\
  &= \underbrace{P_X X}_{=\, X}\, (X^\top X)^{-1} X^\top \\
  &= P_X
  \end{aligned}
  $$

## $P_X$: p.s.d., trace $k$

- For $a \in \mathbb{R}^n$:
  $$
  \begin{aligned}
  a^\top P_X a &= \underbrace{a^\top X}_{w^\top}\ \underbrace{(X^\top X)^{-1}}_{\text{p.d.}}\ \underbrace{X^\top a}_{w} \\
  &\ge 0
  \end{aligned}
  $$

- We will use: $\operatorname{tr}(AB) = \operatorname{tr}(BA)$.
  $$
  \begin{aligned}
  \operatorname{tr}(P_X) &= \operatorname{tr}\big( X (X^\top X)^{-1} X^\top \big) \\
  &= \operatorname{tr}\big( (X^\top X)^{-1} \underbrace{\underset{k\times n}{X^\top}\, \underset{n\times k}{X}}_{k\times k} \big) \\
  &= \operatorname{tr}(I_k) \\
  &= k
  \end{aligned}
  $$

- $P_X$: $n \times n$, symmetric, p.s.d., idempotent, rank $k$.

- $P_X: \mathbb{R}^n \longrightarrow S(X)$.

## Properties of $M_X$

- Symmetric:
  $$
  \begin{aligned}
  M_X^\top &= I_n^\top - P_X^\top \\
  &= I_n - P_X \\
  &= M_X
  \end{aligned}
  $$

- $M_X X = 0$:
  $$
  \begin{aligned}
  M_X X &= (I_n - P_X) X \\
  &= I_n X - P_X X \\
  &= X - \underbrace{P_X X}_{=\, X} \\
  &= X - X \\
  &= 0
  \end{aligned}
  $$

- $v = Xb \in S(X)$, $b \in \mathbb{R}^k$:
  $$
  \begin{aligned}
  M_X v &= M_X X b \\
  &= 0
  \end{aligned}
  $$

## $M_X P_X = 0$ and idempotency

- $M_X P_X = 0$ and $P_X M_X = 0$:
  $$
  \begin{aligned}
  M_X P_X &= \underbrace{M_X X}_{=\, 0}\, (X^\top X)^{-1} X^\top \\
  &= 0 \\
  P_X M_X &= (M_X P_X)^\top \\
  &= 0
  \end{aligned}
  $$

- Idempotent:
  $$
  \begin{aligned}
  M_X M_X &= (I_n - P_X) M_X \\
  &= M_X - \underbrace{P_X M_X}_{=\, 0} \\
  &= M_X
  \end{aligned}
  $$

## Orthogonal complement $S^\perp(X)$

- Orthogonal complement of $S(X)$:
  $$
  S^\perp(X) = \big\{\, v \in \mathbb{R}^n : v^\top X = 0 \,\big\}
  $$

- Normal equations, first lecture:
  $$
  \begin{aligned}
  X^\top \hat U &= X^\top (Y - X\hat\beta) \\
  &= X^\top Y - X^\top X \underbrace{\hat\beta}_{=\, (X^\top X)^{-1} X^\top Y} \\
  &= X^\top Y - X^\top Y \\
  &= 0 \\
  \Longrightarrow\quad \hat U^\top X &= 0 \\
  \Longrightarrow\quad \hat U &= M_X Y \in S^\perp(X)
  \end{aligned}
  $$

- $M_X: \mathbb{R}^n \longrightarrow S^\perp(X)$.

## $M_X$: rank $n - k$, p.s.d.

- Rank:
  $$
  \begin{aligned}
  \rank{M_X} &= \operatorname{tr}(M_X) \\
  &= \operatorname{tr}(I_n - P_X) \\
  &= \operatorname{tr}(I_n) - \operatorname{tr}(P_X) \\
  &= n - \operatorname{tr}(P_X) \\
  &= n - \operatorname{tr}\big( X (X^\top X)^{-1} X^\top \big) \\
  &= n - \operatorname{tr}\big( \underbrace{(X^\top X)^{-1} X^\top X}_{I_k} \big) \\
  &= n - k
  \end{aligned}
  $$

- $M_X$: $n \times n$; no inverse.

- **Claim.** $M_X$: p.s.d. For $a \in \mathbb{R}^n$ and $w = M_X a$:
  $$
  \begin{aligned}
  a^\top M_X a &= a^\top M_X M_X a \quad \text{(idempotent)} \\
  &= \underbrace{a^\top M_X^\top}_{w^\top}\ \underbrace{M_X a}_{w} \quad \text{(symmetric)} \\
  &= w^\top w \\
  &= \|w\|^2 \\
  &\ge 0
  \end{aligned}
  $$

## $Y = \hat Y + \hat U$

- Decomposition:
  $$
  \underset{\in\, \mathbb{R}^n}{Y} = \underset{\in\, S(X)}{\hat Y} + \underset{\in\, S^\perp(X)}{\hat U}, \qquad S(X) + S^\perp(X) = \mathbb{R}^n
  $$

- Fitted values:
  $$
  \begin{aligned}
  \hat Y &= X\hat\beta \\
  &= P_X Y \in S(X), \quad \text{dimension } k
  \end{aligned}
  $$

- Residuals:
  $$
  \begin{aligned}
  \hat U &= Y - \hat Y \\
  &= M_X Y \in S^\perp(X), \quad \text{dimension } n - k
  \end{aligned}
  $$

- Orthogonality:
  $$
  \begin{aligned}
  \hat Y^\top \hat U &= Y^\top P_X^\top M_X Y \\
  &= Y^\top \underbrace{P_X M_X}_{=\, 0}\, Y \\
  &= 0
  \end{aligned}
  $$

# Estimation of $\sigma^2$

## Two estimators of $\sigma^2$

- Assumptions 2 and 3:
  $$
  \begin{aligned}
  \Var{U \mid X} &= \E{U U^\top \mid X} \\
  &= \sigma^2 I_n
  \end{aligned}
  $$

- Unknown parameter:
  $$
  \begin{aligned}
  \sigma^2 &= \E{U_i^2 \mid X} \\
  &= \E{U_i^2}
  \end{aligned}
  $$

- Estimators:
  $$
  \begin{aligned}
  \hat\sigma^2 &= \frac{1}{n} \sum_{i=1}^n \hat U_i^2 \\
  &= \frac{1}{n} \sum_{i=1}^n (Y_i - X_i^\top \underset{k\times 1}{\hat\beta})^2 \\
  &= \frac{1}{n}\, \hat U^\top \hat U \\
  s^2 &= \frac{1}{n-k} \sum_{i=1}^n \hat U_i^2
  \end{aligned}
  $$

- **Claim.** Under Assumptions 1 to 4:
  $$
  \E{s^2 \mid X} = \sigma^2
  $$

## Proof, step 1: residuals to errors

- Residuals:
  $$
  \begin{aligned}
  \hat U &= M_X Y \\
  &= M_X (X\beta + U) \quad \text{(Assumption 1)} \\
  &= M_X U \quad (M_X X = 0)
  \end{aligned}
  $$

- Dimensions:
  $$
  \underset{n\times 1}{\hat U} = \underset{n\times n}{M_X}\ \underset{n\times 1}{U}, \qquad \rank{M_X} = n - k
  $$

- Sum of squared residuals:
  $$
  \begin{aligned}
  \sum_{i=1}^n \hat U_i^2 &= \|\hat U\|^2 \\
  &= \hat U^\top \hat U \\
  &= (M_X U)^\top (M_X U) \\
  &= U^\top M_X^\top M_X U \\
  &= U^\top M_X U \quad \text{(symmetric, idempotent)}
  \end{aligned}
  $$

## Proof, step 2: a quadratic form

- To find:
  $$
  \E{\hat U^\top \hat U \mid X} = \E{\overbrace{\underset{1\times n}{U^\top}\ \underset{n\times n}{M_X}\ \underset{n\times 1}{U}}^{1\times 1} \mid X}
  $$

- Does not factor:
  $$
  \E{U^\top M_X U \mid X} \ \ne\ \E{U^\top \mid X}\, M_X\, \E{U \mid X}
  $$

- Quadratic form:
  $$
  \underset{1\times n}{U^\top}\ \underset{n\times n}{A}\ \underset{n\times 1}{U} = \sum_{i=1}^n \sum_{j=1}^n A_{ij} U_i U_j
  $$

- $\E{U_i U_j} \ne \E{U_i}\E{U_j}$ unless uncorrelated; $\E{U_i^2} \ne \E{U_i}\E{U_i}$.

## Proof, step 3: the trace trick

- We will use:
  - $1 \times 1$ matrix: its own trace;
  - $\operatorname{tr}(AB) = \operatorname{tr}(BA)$;
  - $\E{\sum \cdot} = \sum \E{\cdot}$;
  - $M_X$: known given $X$.
- We have:
  $$
  \begin{aligned}
  \E{\hat U^\top \hat U \mid X} &= \E{U^\top M_X U \mid X} \\
  &= \E{\operatorname{tr}\big({\color{red} U^\top} M_X U\big) \mid X} \\
  &= \E{\operatorname{tr}\big(M_X U {\color{red} U^\top}\big) \mid X} \\
  &= \operatorname{tr}\big(\E{M_X U U^\top \mid X}\big) \\
  &= \operatorname{tr}\big(M_X \underbrace{\E{U U^\top \mid X}}_{=\, \sigma^2 I_n}\big) \\
  &= \sigma^2 \operatorname{tr}(M_X) \\
  &= \sigma^2 (n - k)
  \end{aligned}
  $$

## $\hat\sigma^2$: biased

- Residuals:
  $$
  \begin{aligned}
  \E{\sum_{i=1}^n \hat U_i^2 \mid X} &= \sigma^2 (n - k) \\
  \Longrightarrow\quad \E{\sum_{i=1}^n \hat U_i^2} &= \sigma^2 (n - k) \quad \text{(LIE)}
  \end{aligned}
  $$

- Errors:
  $$
  \begin{aligned}
  \E{U^\top U} &= \E{\sum_{i=1}^n U_i^2} \\
  &= n \sigma^2
  \end{aligned}
  $$

- Biased, by a known factor:
  $$
  \begin{aligned}
  \E{\hat\sigma^2 \mid X} &= \frac{1}{n}\, \E{\hat U^\top \hat U \mid X} \\
  &= \frac{n-k}{n}\, \sigma^2 \\
  &\ne \sigma^2
  \end{aligned}
  $$

## $s^2$: unbiased

- Correction:
  $$
  \begin{aligned}
  \frac{n}{n-k}\, \E{\hat\sigma^2 \mid X} &= \sigma^2 \\
  s^2 &= \frac{n}{n-k}\, \hat\sigma^2 \\
  &= \frac{\hat U^\top \hat U}{n-k} \\
  &= \frac{\sum_{i=1}^n \hat U_i^2}{n-k} \\
  &= \frac{\sum_{i=1}^n (Y_i - X_i^\top \hat\beta)^2}{n-k} \\
  &= \frac{\|Y - X\hat\beta\|^2}{n-k}
  \end{aligned}
  $$

- Unbiased:
  $$
  \begin{aligned}
  & \boxed{\ \E{s^2 \mid X} = \sigma^2\ } \\
  \overset{\text{LIE}}{\Longrightarrow}\quad & \E{s^2} = \sigma^2
  \end{aligned}
  $$

## Standard errors

- Second lecture: $\Var{\hat\beta \mid X} = \sigma^2 (X^\top X)^{-1}$. Estimated:
  $$
  \widehat{\mathrm{Var}}\big(\hat\beta \mid X\big) = s^2 (X^\top X)^{-1}
  $$

- Standard error against standard deviation:
  $$
  \begin{aligned}
  \text{std. err}\big(\hat\beta_j\big) &= \sqrt{s^2 \big[(X^\top X)^{-1}\big]_{jj}} \\
  &= \sqrt{\widehat{\mathrm{Var}}\big(\hat\beta_j \mid X\big)} \\
  \text{std. dev}\big(\hat\beta_j\big) &= \sqrt{\Var{\hat\beta_j \mid X}}
  \end{aligned}
  $$

# Partitioned regression

## Partition

- Two blocks of regressors, $k_1 + k_2 = k$:
  $$
  \underset{n\times k}{X} = \big[\ \underset{n\times k_1}{X_1} \quad \underset{n\times k_2}{X_2}\ \big], \qquad
  \beta = \begin{bmatrix} \beta_1 \\ \beta_2 \end{bmatrix}
  \begin{matrix} k_1 \times 1 \\ k_2 \times 1 \end{matrix}
  $$

- Model:
  $$
  \begin{aligned}
  Y &= \big[\ X_1 \quad X_2\ \big] \begin{bmatrix} \beta_1 \\ \beta_2 \end{bmatrix} + U \\
  &= X_1 \beta_1 + X_2 \beta_2 + U
  \end{aligned}
  $$

- OLS:
  $$
  \begin{aligned}
  \underset{k\times 1}{\hat\beta} &= \begin{bmatrix} \underset{k_1\times 1}{\hat\beta_1} \\ \underset{k_2\times 1}{\hat\beta_2} \end{bmatrix} \\
  &= (X^\top X)^{-1} X^\top Y
  \end{aligned}
  $$

## Frisch-Waugh-Lovell theorem

- **Claim.**
  $$
  \hat\beta_1 = (X_1^\top M_2 X_1)^{-1} X_1^\top M_2 Y, \qquad M_2 = I_n - P_2, \quad P_2 = X_2 (X_2^\top X_2)^{-1} X_2^\top
  $$

- $M_2$: symmetric, idempotent, $M_2 X_2 = 0$.

- $M_2: \mathbb{R}^n \longrightarrow S^\perp(X_2)$.

- $M_1 = I_n - X_1 (X_1^\top X_1)^{-1} X_1^\top$:
  $$
  \hat\beta_2 = (X_2^\top M_1 X_2)^{-1} X_2^\top M_1 Y
  $$

## Proof

- We will use:
  $$
  \begin{aligned}
  \hat U &= Y - X\hat\beta \\
  &= Y - X_1 \hat\beta_1 - X_2 \hat\beta_2 \\
  \Longleftrightarrow\quad Y &= X\hat\beta + \hat U \\
  &= X_1 \hat\beta_1 + X_2 \hat\beta_2 + \hat U
  \end{aligned}
  $$

- We have:
  $$
  \begin{aligned}
  & (X_1^\top M_2 X_1)^{-1} X_1^\top M_2 Y \\
  &= (X_1^\top M_2 X_1)^{-1} X_1^\top M_2 \big( X_1 \hat\beta_1 + X_2 \hat\beta_2 + \hat U \big) \\
  &= \underbrace{(X_1^\top M_2 X_1)^{-1} X_1^\top M_2 X_1}_{=\, I_{k_1}}\, \hat\beta_1 \\
  &\qquad + (X_1^\top M_2 X_1)^{-1} X_1^\top {\color{red} \underbrace{M_2 X_2}_{=\, 0}}\, \hat\beta_2 \\
  &\qquad + (X_1^\top M_2 X_1)^{-1} X_1^\top M_2 \hat U \\
  &= \hat\beta_1 + (X_1^\top M_2 X_1)^{-1} X_1^\top \underbrace{M_2 \hat U}_{=\, \hat U} \\
  &= \hat\beta_1 + (X_1^\top M_2 X_1)^{-1} {\color{red} \underbrace{X_1^\top \hat U}_{=\, 0}} \\
  &= \boxed{\hat\beta_1}
  \end{aligned}
  $$

## $M_2 \hat U = \hat U$

- Normal equations, block by block:
  $$
  \begin{aligned}
  0 &= X^\top \hat U \\
  &= \big[\ X_1 \quad X_2\ \big]^\top \hat U \\
  &= \begin{bmatrix} X_1^\top \\ X_2^\top \end{bmatrix} \hat U \\
  &= \begin{bmatrix} X_1^\top \hat U \\ X_2^\top \hat U \end{bmatrix} \\
  &= \begin{bmatrix} 0 \\ 0 \end{bmatrix}
  \end{aligned}
  $$

- Hence:
  $$
  \begin{aligned}
  M_2 \hat U &= (I_n - P_2) \hat U \\
  &= \hat U - P_2 \hat U \\
  &= \hat U - X_2 (X_2^\top X_2)^{-1} \underbrace{X_2^\top \hat U}_{=\, 0} \\
  &= \hat U
  \end{aligned}
  $$

- Hence:
  $$
  \begin{aligned}
  X_1^\top M_2 \hat U &= X_1^\top \hat U \\
  &= 0
  \end{aligned}
  $$

## Residuals on residuals

- We will use: $M_2^\top = M_2$ and $M_2 M_2 = M_2$.

- We have:
  $$
  \begin{aligned}
  \hat\beta_1 &= (X_1^\top M_2 X_1)^{-1} X_1^\top M_2 Y \\
  &= (X_1^\top M_2^\top M_2 X_1)^{-1} X_1^\top M_2^\top M_2 Y \\
  &= \big( (M_2 X_1)^\top (M_2 X_1) \big)^{-1} (M_2 X_1)^\top (M_2 Y) \\
  &= (\tilde X_1^\top \tilde X_1)^{-1} \tilde X_1^\top \tilde Y
  \end{aligned}
  $$

- $\tilde X_1 = M_2 X_1$, $\tilde Y = M_2 Y$, $\tilde U = M_2 U$:
  $$
  \begin{aligned}
  Y &= X_1 \beta_1 + X_2 \beta_2 + U \\
  M_2 Y &= M_2 X_1 \beta_1 + \underbrace{M_2 X_2}_{=\, 0}\, \beta_2 + M_2 U \\
  &= M_2 X_1 \beta_1 + M_2 U \\
  \tilde Y &= \tilde X_1 \beta_1 + \tilde U
  \end{aligned}
  $$

## $\tilde X_1$ and $\tilde Y$: OLS residuals

- Residuals from the regression of the columns of $X_1$ against $X_2$:
  $$
  \begin{aligned}
  \tilde X_1 &= M_2 X_1 \\
  &= (I_n - P_2) X_1 \\
  &= X_1 - P_2 X_1 \\
  &= X_1 - X_2 \underbrace{(X_2^\top X_2)^{-1} X_2^\top X_1}_{\hat\gamma} \\
  &= X_1 - X_2 \hat\gamma \\
  \Longleftrightarrow\quad X_1 &= X_2 \hat\gamma + \tilde X_1
  \end{aligned}
  $$

- $\tilde X_1^\top X_2 = 0$.

- Residuals from the regression of $Y$ against $X_2$:
  $$
  \begin{aligned}
  \tilde Y &= M_2 Y \\
  &= Y - X_2 \underbrace{(X_2^\top X_2)^{-1} X_2^\top Y}_{\hat\eta} \\
  \Longleftrightarrow\quad Y &= X_2 \hat\eta + \tilde Y
  \end{aligned}
  $$

- In general: $\hat\eta \ne \hat\beta_2$, $\tilde Y \ne \hat U$.

## Regression in three steps

1. Remove from $X_1$ the part collinear with $X_2$: $\tilde X_1$.
2. Remove from $Y$ the part collinear with $X_2$: $\tilde Y$.
3. $\hat\beta_1$: the OLS coefficient from the regression of $\tilde Y$ against $\tilde X_1$.

- $\hat\beta_1$: the estimated effect of $X_1$ after controlling for $X_2$.

- Step 2 optional:
  $$
  \begin{aligned}
  \hat\beta_1 &= (X_1^\top M_2 X_1)^{-1} X_1^\top M_2 Y \\
  &= (X_1^\top M_2^\top M_2 X_1)^{-1} X_1^\top M_2^\top Y \\
  &= \big( (M_2 X_1)^\top (M_2 X_1) \big)^{-1} (M_2 X_1)^\top Y \\
  &= (\tilde X_1^\top \tilde X_1)^{-1} \tilde X_1^\top Y
  \end{aligned}
  $$

- Step 1 is not optional:
  $$
  (X_1^\top X_1)^{-1} X_1^\top \tilde Y \ \ne\ (\tilde X_1^\top \tilde X_1)^{-1} \tilde X_1^\top \tilde Y \quad \text{in general}
  $$

# Partitioned regression and the intercept

## Intercept as a regressor

- One regressor and an intercept:
  $$
  Y_i = \underset{1\times 1}{\beta_1} + \underset{1\times 1}{\beta_2}\, \underset{1\times 1}{X_{i,2}} + U_i
  $$

- Matrix form, with $\ell$ the $n \times 1$ vector of ones:
  $$
  \begin{aligned}
  X &= \begin{bmatrix} 1 & X_{1,2} \\ \vdots & \vdots \\ 1 & X_{n,2} \end{bmatrix} \\
  &= \big[\ \ell \quad X_2\ \big] \\
  Y &= \underset{1\times 1}{\beta_1}\, \underset{n\times 1}{\ell} + \underset{1\times 1}{\beta_2}\, \underset{n\times 1}{X_2} + U
  \end{aligned}
  $$

- Partitioned regression:
  $$
  \begin{aligned}
  \underset{1\times 1}{\hat\beta_2} &= \big( \underset{1\times 1}{X_2^\top M_\ell X_2} \big)^{-1}\, \underset{1\times n}{X_2^\top}\ \underset{n\times 1}{M_\ell Y} \\
  M_\ell &= I_n - \ell \big( \underset{1\times 1}{\ell^\top \ell} \big)^{-1} \ell^\top
  \end{aligned}
  $$

## $P_\ell$ and $M_\ell$

- $\ell^\top \ell$:
  $$
  \begin{aligned}
  \ell^\top \ell &= \begin{bmatrix} 1 \\ \vdots \\ 1 \end{bmatrix}^\top \begin{bmatrix} 1 \\ \vdots \\ 1 \end{bmatrix} \\
  &= \sum_{i=1}^n 1 \cdot 1 \\
  &= n
  \end{aligned}
  $$

- Hence:
  $$
  \begin{aligned}
  P_\ell &= \ell (\ell^\top \ell)^{-1} \ell^\top \\
  &= \frac{\ell \ell^\top}{n} \\
  M_\ell &= I_n - \frac{\ell \ell^\top}{n}
  \end{aligned}
  $$

- For $v = (v_1, \ldots, v_n)^\top$:
  $$
  \begin{aligned}
  (\ell^\top \ell)^{-1} \ell^\top v &= \frac{\ell^\top v}{n} \\
  &= \frac{\sum_{i=1}^n \ell_i v_i}{n} \\
  &= \frac{\sum_{i=1}^n v_i}{n} \\
  &= \bar v
  \end{aligned}
  $$

## $P_\ell$: averaging

- For $v \in \mathbb{R}^n$:
  $$
  \begin{aligned}
  P_\ell v &= \ell (\ell^\top \ell)^{-1} \ell^\top v \\
  &= \frac{\ell \ell^\top}{n}\, v \\
  &= \ell \left( \frac{\sum_{i=1}^n 1 \cdot v_i}{n} \right) \\
  &= \ell\, \bar v \\
  &= \begin{pmatrix} \bar v \\ \vdots \\ \bar v \end{pmatrix}
  \end{aligned}
  $$

- Column space of $\ell$:
  $$
  \begin{aligned}
  S(\ell) &= \big\{\, w \in \mathbb{R}^n : w = c\, \ell,\ c \in \mathbb{R} \,\big\} \\
  &= \left\{\, w \in \mathbb{R}^n : w = \begin{pmatrix} c \\ \vdots \\ c \end{pmatrix},\ c \in \mathbb{R} \,\right\}
  \end{aligned}
  $$

## $M_\ell$: deviations from the mean

- For $v \in \mathbb{R}^n$:
  $$
  \begin{aligned}
  M_\ell v &= v - P_\ell v \\
  &= v - \ell\, \bar v \\
  &= \begin{pmatrix} v_1 - \bar v \\ \vdots \\ v_n - \bar v \end{pmatrix}
  \end{aligned}
  $$

- Orthogonal complement of $S(\ell)$:
  $$
  \begin{aligned}
  S^\perp(\ell) &= \Big\{\, v \in \mathbb{R}^n : \underbrace{v^\top \ell}_{=\, \sum_{i=1}^n v_i} = 0 \,\Big\} \\
  &= \Big\{\, v \in \mathbb{R}^n : \sum_{i=1}^n v_i = 0 \,\Big\} \\
  &= \big\{\, v \in \mathbb{R}^n : \bar v = 0 \,\big\}
  \end{aligned}
  $$

## $\tilde X_2 = M_\ell X_2$

- Mean of the regressor:
  $$
  \begin{aligned}
  (\ell^\top \ell)^{-1} \ell^\top X_2 &= \frac{\ell^\top X_2}{n} \\
  &= \frac{1}{n} \begin{bmatrix} 1 \\ \vdots \\ 1 \end{bmatrix}^\top \begin{bmatrix} X_{1,2} \\ \vdots \\ X_{n,2} \end{bmatrix} \\
  &= \frac{1}{n} \sum_{i=1}^n 1 \cdot X_{i,2} \\
  &= \frac{1}{n} \sum_{i=1}^n X_{i,2} \\
  &= \bar X_2
  \end{aligned}
  $$

- The regressor:
  $$
  \begin{aligned}
  \tilde X_2 &= M_\ell X_2 \\
  &= X_2 - \ell\, \frac{\ell^\top X_2}{n} \\
  &= X_2 - \bar X_2\, \ell \\
  &= \begin{pmatrix} X_{1,2} - \bar X_2 \\ \vdots \\ X_{n,2} - \bar X_2 \end{pmatrix}
  \end{aligned}
  $$

- The dependent variable: $M_\ell Y = (Y_1 - \bar Y, \ldots, Y_n - \bar Y)^\top$.

## Slope: deviations from the mean

- We will use: $M_\ell^\top = M_\ell$, $M_\ell M_\ell = M_\ell$, and $X_2^\top M_\ell X_2$: $1 \times 1$.

- We have:
  $$
  \begin{aligned}
  \hat\beta_2 &= (X_2^\top M_\ell X_2)^{-1} X_2^\top M_\ell Y \\
  &= \frac{X_2^\top M_\ell Y}{X_2^\top M_\ell X_2} \\
  &= \frac{(M_\ell X_2)^\top Y}{(M_\ell X_2)^\top (M_\ell X_2)} \\
  &= \frac{\tilde X_2^\top Y}{\tilde X_2^\top \tilde X_2} \\
  &= \frac{(X_2 - \bar X_2\, \ell)^\top Y}{(X_2 - \bar X_2\, \ell)^\top (X_2 - \bar X_2\, \ell)} \\
  &= \frac{\sum_{i=1}^n (X_{i,2} - \bar X_2)\, Y_i}{\sum_{i=1}^n (X_{i,2} - \bar X_2)^2} \\
  &= \boxed{\frac{\sum_{i=1}^n (X_{i,2} - \bar X_2)(Y_i - \bar Y)}{\sum_{i=1}^n (X_{i,2} - \bar X_2)^2}} \quad (X_2^\top M_\ell Y = X_2^\top M_\ell^\top M_\ell Y)
  \end{aligned}
  $$

## Several regressors and an intercept

- Intercept $\beta_1$ and slope coefficients $\beta_2$:
  $$
  \underset{n\times k}{X} = \big[\ \underset{n\times 1}{\ell} \quad \underset{n\times (k-1)}{X_2}\ \big], \qquad
  \beta = \begin{pmatrix} \beta_1 \\ \beta_2 \end{pmatrix}, \qquad
  Y = \beta_1 \ell + X_2 \beta_2 + U
  $$

- Each regressor minus its mean:
  $$
  \begin{aligned}
  \tilde X_2 &= M_\ell X_2 \\
  &= M_\ell \big[\, \underline{X}_2 \ \cdots \ \underline{X}_k \,\big] \\
  &= \big[\, M_\ell \underline{X}_2 \ \cdots \ M_\ell \underline{X}_k \,\big] \\
  &= \big[\, \underline{\tilde X}_2 \ \cdots \ \underline{\tilde X}_k \,\big]
  \end{aligned}
  $$

- Slope coefficients:
  $$
  \begin{aligned}
  \hat\beta_2 &= (X_2^\top M_\ell X_2)^{-1} X_2^\top M_\ell Y \\
  &= (\tilde X_2^\top \tilde X_2)^{-1} \tilde X_2^\top Y
  \end{aligned}
  $$

- Variation of $X_2$ in the sample, with $X_{i,2}$ now $(k-1) \times 1$ and $\bar X_2 = \frac{1}{n} \sum_{i=1}^n X_{i,2}$:
  $$
  \begin{aligned}
  \tilde X_2^\top \tilde X_2 &= \sum_{i=1}^n \tilde X_{i,2} \tilde X_{i,2}^\top \\
  &= \sum_{i=1}^n (X_{i,2} - \bar X_2)(X_{i,2} - \bar X_2)^\top
  \end{aligned}
  $$

## Intercept, 1: deviations from means

- In the sample:
  $$
  \begin{aligned}
  Y_i &= \beta_1 + X_{i,2}^\top \beta_2 + U_i \\
  \bar Y &= \beta_1 + \bar X_2^\top \beta_2 + \bar U \\
  \underbrace{Y_i - \bar Y}_{\tilde Y_i} &= \underbrace{(X_{i,2} - \bar X_2)^\top}_{\tilde X_{i,2}^\top} \beta_2 + \underbrace{(U_i - \bar U)}_{\tilde U_i} \\
  \tilde Y_i &= \tilde X_{i,2}^\top \beta_2 + \tilde U_i
  \end{aligned}
  $$

- In the population:
  $$
  \begin{aligned}
  Y_i &= \beta_1 + X_{i,2}^\top \beta_2 + U_i \\
  \E{Y_i} &= \beta_1 + \E{X_{i,2}}^\top \beta_2 \\
  Y_i - \E{Y_i} &= (X_{i,2} - \E{X_{i,2}})^\top \beta_2 + U_i
  \end{aligned}
  $$

## Intercept, 2: non-zero error mean

- $\E{U_i \mid X_i} = \mu \ne 0$:
  $$
  \begin{aligned}
  \E{Y_i} &= \beta_1 + \E{X_{i,2}}^\top \beta_2 + \mu \\
  Y_i - \E{Y_i} &= (X_{i,2} - \E{X_{i,2}})^\top \beta_2 + (U_i - \mu)
  \end{aligned}
  $$

- Then:
  $$
  \begin{aligned}
  Y_i &= \beta_1 + X_{i,2}^\top \beta_2 + U_i \\
  &= \beta_1 + X_{i,2}^\top \beta_2 + (U_i - \mu) + \mu \\
  &= \underbrace{(\beta_1 + \mu)}_{=\, \beta_1^*} + X_{i,2}^\top \beta_2 + \underbrace{(U_i - \mu)}_{=\, U_i^*} \\
  &= \beta_1^* + X_{i,2}^\top \beta_2 + U_i^*
  \end{aligned}
  $$

- $\E{U_i^* \mid X_i} = 0$.

- Estimable: $\beta_2$ and $\beta_1^* = \beta_1 + \mu$.

# Summary

## Projections

- $P_X = X (X^\top X)^{-1} X^\top$ and $M_X = I_n - P_X$: symmetric, idempotent, p.s.d.

- $P_X X = X$, $M_X X = 0$, $P_X M_X = 0$.

- $\operatorname{tr}(P_X) = k$ and $\rank{M_X} = \operatorname{tr}(M_X) = n - k$.

- $Y = \hat Y + \hat U$ with $\hat Y = P_X Y \in S(X)$, $\hat U = M_X Y \in S^\perp(X)$, $\hat Y^\top \hat U = 0$.

## $s^2$ and partitioned regression

- $\hat U = M_X U$, $\hat U^\top \hat U = U^\top M_X U$, $\E{\hat U^\top \hat U \mid X} = \sigma^2 \operatorname{tr}(M_X) = \sigma^2 (n - k)$.

- $\E{\hat\sigma^2 \mid X} = \frac{n-k}{n}\, \sigma^2$; $s^2 = \hat U^\top \hat U / (n - k)$ with $\E{s^2 \mid X} = \sigma^2$; $\widehat{\mathrm{Var}}\big(\hat\beta \mid X\big) = s^2 (X^\top X)^{-1}$.

- Frisch-Waugh-Lovell: $\hat\beta_1 = (X_1^\top M_2 X_1)^{-1} X_1^\top M_2 Y = (\tilde X_1^\top \tilde X_1)^{-1} \tilde X_1^\top \tilde Y$, with $\tilde X_1 = M_2 X_1$ and $\tilde Y = M_2 Y$.

- Intercept, one regressor: $\hat\beta_2 = \sum_i (X_{i,2} - \bar X_2) Y_i \big/ \sum_i (X_{i,2} - \bar X_2)^2$.
