---
# _quarto.yml in this folder sets two things for every deck: `output-dir: output`,
# which sends the rendered files to output/, and `embed-resources: true` on the
# page and on the slide show, which keeps both of them self-contained. Quarto
# finds that file by walking up from this one, so both settings hold wherever the
# render is started from. Do not repeat either setting here.
title: "Lecture 2: Properties of the OLS estimator"
subtitle: "Economics 527: Econometric Methods"
author:
  - name: "Vadim Marmer, UBC"
format:
  html:
    output-file: 527_02_ols_properties.html
    toc: true
    toc-depth: 3
    toc-location: right
    toc-title: "Table of Contents"
    theme: cosmo
    smooth-scroll: true
    html-math-method: katex
  pdf:
    output-file: 527_02_ols_properties.pdf
    pdf-engine: xelatex
    geometry: margin=0.75in
    fontsize: 10pt
    number-sections: false
    toc: false
    classoption: fleqn
  revealjs:
    output-file: 527_02_ols_properties_slides.html
    theme: solarized
    css: slides_no_caps.css
    smaller: true
    slide-number: c/t
    incremental: true
    html-math-method: katex
    navigation-mode: linear
    scrollable: true
    chalkboard: false
    transition: none
---

# The model and the assumptions

## Where the first lecture ended

::: {.hidden}
\gdef\E#1{\mathrm{E}\left[#1\right]}
\gdef\Var#1{\mathrm{Var}\left(#1\right)}
\gdef\Cov#1{\mathrm{Cov}\left(#1\right)}
:::

```{r}
#| include: false
# Figures are drawn through tikzDevice, so every label is typeset by LaTeX
# itself and matches the mathematics in the slides. Needs tikzDevice, png and
# a working pdflatex and pdftoppm.
suppressPackageStartupMessages(library(tikzDevice))
# Keep the string-metrics cache out of the source folder.
options(tikzMetricsDictionary = file.path(tempdir(), "tikzMetrics"))

tikz_fig <- function(stem, width, height, expr) {
  d <- file.path(tempdir(), "tikzfig")
  dir.create(d, showWarnings = FALSE, recursive = TRUE)
  tex <- file.path(d, paste0(stem, ".tex"))
  tikz(tex, width = width, height = height, pointsize = 11, standAlone = TRUE)
  expr
  dev.off()
  system2("pdflatex", c("-interaction=batchmode", "-halt-on-error",
                        paste0("-output-directory=", shQuote(d)), shQuote(tex)),
          stdout = FALSE, stderr = FALSE)
  system2("pdftoppm", c("-r", "300", "-png", "-singlefile",
                        shQuote(file.path(d, paste0(stem, ".pdf"))),
                        shQuote(file.path(d, stem))),
          stdout = FALSE, stderr = FALSE)
  img <- png::readPNG(file.path(d, paste0(stem, ".png")))
  op <- par(mar = rep(0, 4))
  plot.new(); plot.window(c(0, 1), c(0, 1), xaxs = "i", yaxs = "i")
  rasterImage(img, 0, 0, 1, 1, interpolate = TRUE)
  par(op)
  invisible(NULL)
}
```

- The first lecture produced the estimator and stopped there:
$$
\hat\beta = (X^\top X)^{-1} X^\top Y
= \left(\sum_{i=1}^n X_i X_i^\top\right)^{-1}\sum_{i=1}^n X_i Y_i,
$$
the unique minimiser of $\sum_{i=1}^n (Y_i - X_i^\top b)^2$.

- Nothing so far says whether it is a good estimator.

- $\hat\beta$ is a function of the sample, so it is random, and in general it will not equal $\beta$.

- Three questions, taken in this order: whether $\hat\beta$ is right on average, how far it moves from sample to sample, and whether a better estimator of the same kind exists.

## The model in matrix form

- Stack the $n$ observations, as in the first lecture:
$$
\underset{n\times1}{Y} = \begin{pmatrix} Y_1 \\ \vdots \\ Y_n \end{pmatrix},\qquad
\underset{n\times k}{X} = \begin{pmatrix} X_1^\top \\ \vdots \\ X_n^\top \end{pmatrix},\qquad
\underset{n\times1}{U} = \begin{pmatrix} U_1 \\ \vdots \\ U_n \end{pmatrix}.
$$

- The $n$ scalar equations $Y_i = X_i^\top\beta + U_i$ become one:
$$
\underset{n\times1}{Y} = \underset{n\times k}{X}\ \underset{k\times1}{\beta} + \underset{n\times1}{U}.
$$

- $\beta$ is fixed and unknown. $X$, $Y$ and $U$ are random, and $U$ is not observed.

## The four assumptions

- **Assumption 1.** $Y = X\beta + U$ for some $\beta \in \mathbb{R}^k$.

- **Assumption 2.** $\E{U \mid X} = 0$.

- **Assumption 3.** $\Var{U \mid X} = \sigma^2 I_n$ for some $\sigma^2 > 0$.

- **Assumption 4.** $\operatorname{rank}(X) = k$.


# Unbiasedness

## What unbiased means

- $\hat\beta$ is a function of the sample, hence a random vector:
$$
\hat\beta(\omega) = \left(\sum_{i=1}^n X_i(\omega) X_i(\omega)^\top\right)^{-1}\sum_{i=1}^n X_i(\omega) Y_i(\omega).
$$

- With a continuously distributed error, $\mathrm{P}(\hat\beta = \beta) = 0$. Hitting $\beta$ exactly is not the standard.

- **Definition.** An estimator $\hat\theta$ of $\theta$ is **unbiased** if $\E{\hat\theta} = \theta$, and **conditionally unbiased given $X$** if
$$
\E{\hat\theta \mid X} = \theta,
$$
whatever the true value of $\theta$: the estimator is a rule that does not know $\theta$, and the equality must hold for every value $\theta$ could take.

- The conditional statement is the stronger one, and it is the one this lecture proves.



## The sampling error

- Substitute Assumption 1 into the estimator; Assumption 4 is what makes $(X^\top X)^{-1}$ exist:
$$
\begin{aligned}
\hat\beta &= (X^\top X)^{-1} X^\top Y \\
&= (X^\top X)^{-1} X^\top (X\beta + U) \\
&= \underbrace{(X^\top X)^{-1} X^\top X}_{=\, I_k}\, \beta + (X^\top X)^{-1} X^\top U.
\end{aligned}
$$

- The first term produces $\beta$, and we have signal+noise decomposition:
$$
\boxed{\ \hat\beta = \underbrace{\beta}_{\text{signal}} + \underbrace{{\color{blue}(X^\top X)^{-1} X^\top U}}_{\text{noise}}\ }
$$

- Everything in this section and the next is a statement about the noise term.

## Taking expectations, first attempt

- Take expectations of the sampling error:
$$
\E{\hat\beta} = \beta + \E{{\color{blue}(X^\top X)^{-1} X^\top U}}.
$$

- Suppose only that $\E{X_i U_i} = 0$ for every $i$. Then
$$
\E{X^\top U} = \E{\sum_{i=1}^n X_i U_i} = \sum_{i=1}^n \E{X_i U_i} = 0.
$$

- That is not enough, because in general
$$
\E{(X^\top X)^{-1} X^\top U} \;\text{ need not equal }\; \E{(X^\top X)^{-1}}\,\E{X^\top U}.
$$

- The two factors are built from the same $X$, and once $\E{U \mid X}$ is not zero, the expectation of their product need not factor.

## The expectation of a ratio

- The one-regressor case makes the failure visible:
$$
\E{\frac{\sum_i X_i U_i}{\sum_i X_i^2}} \;\text{ need not equal }\; \frac{\E{\sum_i X_i U_i}}{\E{\sum_i X_i^2}} = 0.
$$

- In general the expectation of a ratio is not the ratio of expectations. For any two scalar random variables $Z$ and $S$,
$$
\E{\frac{Z}{S}} = \iint \frac{z}{s}\, f_{Z,S}(z,s)\, dz\, ds
\;\text{ need not equal }\;
\frac{\int z f_Z(z)\, dz}{\int s f_S(s)\, ds} = \frac{\E{Z}}{\E{S}}.
$$

- The way out is to condition on $X$, which turns $(X^\top X)^{-1}X^\top$ into a known matrix.

## Conditional unbiasedness: the claim

- **Claim.** Under Assumptions 1, 2 and 4,
$$
\E{\hat\beta \mid X} = \beta.
$$

- Assumption 3 is not used. Unbiasedness does not depend on the variance of the errors.

## Proof of conditional unbiasedness

- Condition the sampling error on $X$:
$$
\begin{aligned}
\E{\hat\beta \mid X} &= \E{\beta + (X^\top X)^{-1}X^\top U \mid X} \\
&= \beta + \E{(X^\top X)^{-1}X^\top U \mid X} \\
&= \beta + (X^\top X)^{-1}X^\top \underbrace{\E{U \mid X}}_{=\,0 \text{ by Assumption 2}} \\
&= \beta.
\end{aligned}
$$

- The third line takes out what is known, as in the first lecture. Given $X$, the matrix $(X^\top X)^{-1}X^\top$ is known and comes out of the expectation.


## From conditional to unconditional

- The law of iterated expectations, from the first lecture, finishes the proof:
$$
\E{\hat\beta} = \E{\E{\hat\beta \mid X}} = \E{\beta} = \beta.
$$

- Conditional unbiasedness implies unbiasedness whenever the unconditional mean exists. The converse does not hold.



## A weaker condition

- Replace Assumption 2 by
$$
\E{X_i U_i} = 0 \quad \text{for all } i,
$$
which, together with $\E{U_i} = 0$, says that each regressor is uncorrelated with the error.

- $\beta$ is still identified under this condition.

- It is not enough for unbiasedness:
$$
\E{\hat\beta} = \beta + \underbrace{\E{(X^\top X)^{-1}X^\top U}}_{\text{need not be } 0}.
$$

- So $\hat\beta$ need not be unbiased. The noise term is a nonlinear function of $X$ and $U$, and once $\E{U \mid X}$ is not zero, nothing forces its mean to vanish.




# The variance of a random vector

## The definition

- For a random vector $V$ that is $n \times 1$ with finite second moments, and with $\E{V} = (\E{V_1}, \ldots, \E{V_n})^\top$,
$$
\Var{V} = \E{\underset{n\times1}{(V - \E{V})}\ \underset{1\times n}{(V - \E{V})^\top}} \qquad (n \times n).
$$

- The product inside is an outer product, not an inner product, so the result is a matrix rather than a number.

## The matrix entry by entry

- Entry $(i,j)$ is $\E{(V_i - \E{V_i})(V_j - \E{V_j})}$, so
$$
\Var{V} = \begin{pmatrix}
\Var{V_1} & \Cov{V_1, V_2} & \cdots & \Cov{V_1, V_n} \\
\Cov{V_2, V_1} & \Var{V_2} & & \vdots \\
\vdots & & \ddots & \\
\Cov{V_n, V_1} & \cdots & & \Var{V_n}
\end{pmatrix}.
$$

- The variances sit on the diagonal and the covariances off it.

## Symmetry

- $\Var{V}^\top = \Var{V}$, because $\Cov{V_i, V_j} = \Cov{V_j, V_i}$.

- The same statement from the definition, using that the transpose passes through the expectation, and then $(AB)^\top = B^\top A^\top$:
$$
\Var{V}^\top = \E{\big((V - \E{V})(V - \E{V})^\top\big)^\top} = \E{(V - \E{V})(V - \E{V})^\top} = \Var{V}.
$$

## Positive semidefiniteness

- **Claim.** $\Var{V}$ is positive semidefinite.

- **Proof.** Take any non-random $a \in \mathbb{R}^n$ and set $W = a^\top (V - \E{V})$, a scalar. Then
$$
\begin{aligned}
a^\top \Var{V} a &= a^\top \E{(V - \E{V})(V - \E{V})^\top} a \\
&= \E{\underbrace{a^\top (V - \E{V})}_{W}\ \underbrace{(V - \E{V})^\top a}_{W}} \\
&= \E{W^2} \;\ge\; 0.
\end{aligned}
$$

- The constant $a$ moves inside the expectation because expectation is linear.

- This is the argument the first lecture used for $\E{X_i X_i^\top}$, with $W_i = X_i^\top c$.

## When it is positive definite

- For $a \neq 0$, with $W = a^\top (V - \E{V})$ as above, track the equality case:
$$
a^\top \Var{V} a = 0 \iff \E{W^2} = 0 \iff a^\top (V - \E{V}) = 0,
$$
that is,
$$
a_1 (V_1 - \E{V_1}) + \cdots + a_n (V_n - \E{V_n}) = 0.
$$

- So $\Var{V}$ is positive definite exactly when no non-zero linear combination of $V_1, \ldots, V_n$ is constant, that is, when no exact linear relation holds among the centred variables $V_i - \E{V_i}$.

## Variance under a linear map

- **Claim.** For a non-random $m \times n$ matrix $\Gamma$,
$$
\boxed{\ \Var{\Gamma V} = \Gamma \Var{V} \Gamma^\top \ } \qquad (m \times m).
$$

- **Proof.** $\E{\Gamma V} = \Gamma \E{V}$, so
$$
\begin{aligned}
\Var{\Gamma V} &= \E{\big(\Gamma V - \Gamma\E{V}\big)\big(\Gamma V - \Gamma\E{V}\big)^\top} \\
&= \E{\Gamma (V - \E{V}) (V - \E{V})^\top \Gamma^\top} \\
&= \Gamma \E{(V - \E{V})(V - \E{V})^\top} \Gamma^\top = \Gamma \Var{V} \Gamma^\top,
\end{aligned}
$$
using $(AB)^\top = B^\top A^\top$ on the second factor, and linearity to move the non-random $\Gamma$ and $\Gamma^\top$ outside the expectation.

- Check against the scalar case: with $m = n = 1$ this is $\Var{\gamma V} = \gamma^2 \Var{V}$.

## Adding a constant

- For a non-random $\alpha$ that is $m \times 1$,
$$
\Var{\alpha + \Gamma V} = \Var{\Gamma V} = \Gamma \Var{V} \Gamma^\top.
$$

- The constant shifts the mean, so the deviations from the mean are unchanged and it drops out.

- This is why $\beta$ will drop out of the variance of the sampling error.

## The conditional version

- The same definition with conditional expectations throughout:
$$
\Var{V \mid X} = \E{\big(V - \E{V \mid X}\big)\big(V - \E{V \mid X}\big)^\top \,\Big|\, X}.
$$

- Everything above holds conditionally, with $\Gamma$ allowed to be a function of $X$. Given $X$, such a $\Gamma$ is known and comes out of the conditional expectation, by taking out what is known, as in the proof of unbiasedness.

- When the conditional mean is zero, the definition simplifies. Under Assumption 2,
$$
\Var{U \mid X} = \E{U U^\top \mid X}.
$$

# The variance of the OLS estimator

## The variance matrix of $\hat\beta$

- $\hat\beta$ is $k \times 1$, so $\Var{\hat\beta \mid X}$ is $k \times k$: the conditional variance of each estimated coefficient on the diagonal, the conditional covariances between them off it.

- Conditioning on $X$ again, for the same reason it worked for unbiasedness: it makes $(X^\top X)^{-1}X^\top$ a known matrix.

## The sandwich formula

- Start from the sampling error and name the matrix that multiplies $U$:
$$
\hat\beta = \beta + \underbrace{(X^\top X)^{-1} X^\top}_{\Gamma_X}\, U.
$$

- $\beta$ is a constant and drops out, and $\Gamma_X$ is known given $X$, so the rule for a linear map applies:
$$
\begin{aligned}
\Var{\hat\beta \mid X} &= \Var{\beta + \Gamma_X U \mid X} \\
&= \Var{\Gamma_X U \mid X} \\
&= \Gamma_X \Var{U \mid X} \Gamma_X^\top.
\end{aligned}
$$

## The sandwich formula written out

- Substituting $\Gamma_X = (X^\top X)^{-1}X^\top$, with $\Gamma_X^\top = X (X^\top X)^{-1}$ because $X^\top X$ and its inverse are symmetric:
$$
\boxed{\ \Var{\hat\beta \mid X} = (X^\top X)^{-1} X^\top\ \Var{U \mid X}\ X (X^\top X)^{-1}\ }
$$

- The box uses Assumptions 1 and 4, and needs the errors to have finite conditional second moments, so that $\Var{U \mid X}$ exists. Assumption 2 enters only to write the middle factor as $\E{U U^\top \mid X}$.

- Apart from requiring $\Var{U \mid X}$ to exist, nothing yet has been assumed about the errors' conditional variances or covariances.

## Assumption 3: homoskedasticity and no autocorrelation

- **Assumption 3.** $\Var{U \mid X} = \sigma^2 I_n$ for some $\sigma^2 > 0$, that is,
$$
\Var{U \mid X} = \begin{pmatrix} \sigma^2 & & 0 \\ & \ddots & \\ 0 & & \sigma^2 \end{pmatrix}.
$$

- Read the diagonal: every error has the same conditional variance,
$$
\Var{U_1 \mid X} = \cdots = \Var{U_n \mid X} = \sigma^2.
$$

- This is **homoskedasticity**. It rules out errors that are more variable at some values of the regressors than at others.

- Read the off-diagonal entries: for $i \neq j$, with the first equality by Assumption 2,
$$
\Cov{U_i, U_j \mid X} = \E{U_i U_j \mid X} = 0.
$$

- The errors are conditionally uncorrelated across observations. This is **no autocorrelation**, also called no serial correlation.

- Assumption 3 is these two statements together, and nothing more. It does not say the errors are independent, and it does not say they are normal.

## The variance of OLS under Assumption 3

- Put $\Var{U \mid X} = \sigma^2 I_n$ into the sandwich:
$$
\begin{aligned}
\Var{\hat\beta \mid X} &= (X^\top X)^{-1} X^\top\ \sigma^2 I_n\ X (X^\top X)^{-1} \\
&= \sigma^2 (X^\top X)^{-1} \underbrace{X^\top X (X^\top X)^{-1}}_{=\, I_k} \\
&= \sigma^2 (X^\top X)^{-1}.
\end{aligned}
$$

- Under Assumptions 1 to 4:
$$
\boxed{\ \E{\hat\beta \mid X} = \beta, \qquad \Var{\hat\beta \mid X} = \sigma^2 (X^\top X)^{-1}\ }
$$

- This pair is the central result of the lecture.

## What $\sigma^2$ is

- From Assumptions 2 and 3 together,
$$
\sigma^2 = \Var{U_i \mid X} = \E{U_i^2 \mid X},
$$
and the law of iterated expectations gives
$$
\sigma^2 = \E{U_i^2} = \Var{U_i},
$$
the last step because $\E{U_i} = \E{\E{U_i \mid X}} = 0$ by Assumption 2.

- $\sigma^2$ is one unknown number, the same for every observation. Estimating it is a question for the next lecture.

## One regressor

- Take $k = 1$ and no intercept, so $Y_i = \beta X_i + U_i$ and $X$ is $n \times 1$. Then $X^\top X = \sum_i X_i^2$ and
$$
\Var{\hat\beta \mid X} = \frac{\sigma^2}{\sum_{i=1}^n X_i^2}.
$$

- Two things move the variance. Noisier errors raise it. A larger $\sum_i X_i^2$ lowers it: with no intercept this is the sum of squares of the regressor about zero. With an intercept the same formula gives the variance of the slope, with $\sum_i (X_i - \bar X)^2$, the spread of the regressor about its mean, in place of $\sum_i X_i^2$.

## Why spread in the regressor helps

- With the regressor bunched together, many quite different slopes fit the data about equally well, so the estimate moves a lot from sample to sample.

- With the regressor spread out, the slope is pinned down.

  ```{r}
  #| echo: false
  #| fig-align: center
  #| fig-width: 9
  #| fig-height: 3.6
  #| fig-dpi: 300
  tikz_fig("spread", 9, 3.6, {
    set.seed(5)
    fit_cols <- c("#d95f02", "#1b9e77", "#7570b3")
    op <- par(mfrow = c(1, 2), mar = c(2.6, 2.6, 2.2, 0.6),
              mgp = c(1.4, 1, 0), font.main = 1)
    xl <- rnorm(40, sd = 0.35); yl <- xl + rnorm(40, sd = 0.5)
    plot(xl, yl, pch = 19, cex = 0.6, col = "gray40", axes = FALSE,
         xlim = c(-4, 4), ylim = c(-5, 5),
         xlab = "$X_i$", ylab = "$Y_i$",
         main = "$\\sum_{i=1}^{n} X_i^2$ small")
    box()
    slopes <- c(-0.7, 1.0, 2.8)
    for (j in seq_along(slopes)) {
      abline(a = mean(yl) - slopes[j] * mean(xl), b = slopes[j],
             lwd = 2, col = fit_cols[j])
    }
    xr <- runif(40, -3.6, 3.6); yr <- xr + rnorm(40, sd = 0.5)
    plot(xr, yr, pch = 19, cex = 0.6, col = "gray40", axes = FALSE,
         xlim = c(-4, 4), ylim = c(-5, 5),
         xlab = "$X_i$", ylab = "$Y_i$",
         main = "$\\sum_{i=1}^{n} X_i^2$ large")
    box()
    abline(lm(yr ~ xr), lwd = 2, col = fit_cols[1])
    par(op)
  })
  ```

## Without homoskedasticity

- If the errors are heteroskedastic but still conditionally uncorrelated across observations, so that $\Var{U \mid X}$ is diagonal with entries that may depend on $X$,
$$
\Var{U \mid X} = \Omega = \begin{pmatrix} \sigma_1^2 & & 0 \\ & \ddots & \\ 0 & & \sigma_n^2 \end{pmatrix},
$$
the sandwich does not collapse. The middle factor becomes
$$
X^\top \Omega X = \begin{pmatrix} X_1 & \cdots & X_n \end{pmatrix} \Omega \begin{pmatrix} X_1^\top \\ \vdots \\ X_n^\top \end{pmatrix} = \sum_{i=1}^n \sigma_i^2 X_i X_i^\top,
$$
so
$$
\Var{\hat\beta \mid X} = (X^\top X)^{-1} \left(\sum_{i=1}^n \sigma_i^2 X_i X_i^\top\right) (X^\top X)^{-1}.
$$

- Unbiasedness is untouched. Only the variance formula changes.

## Where each assumption was used

| Assumptions | What they deliver |
|---|---|
| 1, 2, 4 | $\E{\hat\beta \mid X} = \beta$, and $\E{\hat\beta} = \beta$ whenever that mean exists |
| 1, 4, finite conditional second moments | $\Var{\hat\beta \mid X} = \Gamma_X \Var{U \mid X} \Gamma_X^\top$, with $\Gamma_X = (X^\top X)^{-1}X^\top$ |
| 1, 3, 4 | $\Var{\hat\beta \mid X} = \sigma^2 (X^\top X)^{-1}$ |

# The Gauss-Markov theorem

## Linear estimators

- **Definition.** An estimator of $\beta$ is **linear** if it can be written
$$
\tilde\beta = A Y
$$
for a matrix $A$ that is $k \times n$ and depends on $X$ alone, not on $Y$.

- OLS is linear, with $A = (X^\top X)^{-1} X^\top$.

- Given $X$, each entry of a linear estimator is a fixed linear combination of the observations on the dependent variable.

## Two more linear estimators

- **The two-point slope.** With $k = 1$, $Y_i = \beta X_i + U_i$ and $X_1 \neq X_n$,
$$
\tilde\beta = \frac{Y_n - Y_1}{X_n - X_1} = -\frac{1}{X_n - X_1} Y_1 + \frac{1}{X_n - X_1} Y_n.
$$
It uses two observations and throws the rest away.

- **Weighted least squares.** For a known $n \times n$ matrix $W$ that is symmetric and positive definite, and that may depend on $X$ but not on $Y$,
$$
\tilde\beta = (X^\top W X)^{-1} X^\top W\, Y.
$$
OLS is the case $W = I_n$. With $W$ diagonal, this estimator weights observation $i$ by the $i$th diagonal entry $w_i$, as in the first lecture's slide on other linear estimators. The inverse $(X^\top W X)^{-1}$ exists: for $c \neq 0$, $Xc \neq 0$ by Assumption 4, so $c^\top X^\top W X c = (Xc)^\top W (Xc) > 0$.

## Both are unbiased

- The two-point slope, substituting $Y_i = \beta X_i + U_i$ and using Assumption 2:
$$
\tilde\beta = \beta + \frac{U_n - U_1}{X_n - X_1}, \qquad \E{\tilde\beta \mid X} = \beta.
$$

- Weighted least squares, substituting $Y = X\beta + U$ and using Assumption 2:
$$
\tilde\beta = \beta + (X^\top W X)^{-1} X^\top W\, U, \qquad \E{\tilde\beta \mid X} = \beta.
$$

- Being unbiased does not single OLS out. When $n > k$, there are many conditionally unbiased linear estimators.

## The variance of the weighted estimator

- The route that gave the sandwich formula applies, with $A = (X^\top W X)^{-1} X^\top W$ in place of $\Gamma_X$. Since $W^\top = W$, the matrix $X^\top W X$ and its inverse are symmetric too, so $A^\top = W X (X^\top W X)^{-1}$. Under Assumptions 1, 3 and 4,
$$
\Var{\tilde\beta \mid X} = \sigma^2 (X^\top W X)^{-1} X^\top W\, W X (X^\top W X)^{-1},
$$
which reduces to $\sigma^2 (X^\top X)^{-1}$ when $W = I_n$.

- Which of these matrices is smallest, and what smallest means for matrices, are the next questions.

## The comparison to settle

- For every linear conditionally unbiased $\tilde\beta$, compare the two conditional variance matrices:
$$
\underset{k\times k}{\Var{\tilde\beta \mid X}} - \underset{k\times k}{\Var{\hat\beta \mid X}} \ \overset{?}{\ge}\ 0.
$$

- Here $\ge 0$ means positive semidefinite.

## The Gauss-Markov theorem

- **Theorem.** Under Assumptions 1 to 4, let $\tilde\beta$ satisfy
  1. **linearity**: $\tilde\beta = A Y$, with $A$ a $k \times n$ matrix depending on $X$ alone;
  2. **conditional unbiasedness**: $\E{\tilde\beta \mid X} = \beta$ for every value of $\beta$.

- Then
$$
\Var{\tilde\beta \mid X} - \Var{\hat\beta \mid X} \ \text{ is positive semidefinite.}
$$

- OLS is the **best linear unbiased estimator** (BLUE): best in the sense above, linear in $Y$, conditionally unbiased, and an estimator of $\beta$.

- The theorem is about conditional unbiasedness, because it compares conditional variances.

## Reading the matrix ordering

- If a $k \times k$ matrix $M$ is positive semidefinite, then $c^\top M c \ge 0$ for **every** $c \in \mathbb{R}^k$.

- Take $c$ to be the $j$th coordinate vector $e_j$, the vector with a $1$ in position $j$ and a $0$ everywhere else:
$$
e_j = (0, \ldots, 0, 1, 0, \ldots, 0)^\top, \qquad \text{the } 1 \text{ sitting in position } j.
$$

- Writing the quadratic form out entry by entry, and using $(e_j)_l = 0$ for every $l \neq j$ and $(e_j)_j = 1$,
$$
e_j^\top M e_j = \sum_{l=1}^k \sum_{m=1}^k (e_j)_l\, M_{lm}\, (e_j)_m = M_{jj}.
$$

- So every diagonal entry of a positive semidefinite matrix is non-negative.

- Apply this to $M = \Var{\tilde\beta \mid X} - \Var{\hat\beta \mid X}$, whose $j$th diagonal entry is the difference of the two variances of the $j$th coefficient:
$$
\Var{\tilde\beta_j \mid X} \ \ge\ \Var{\hat\beta_j \mid X}, \qquad j = 1, \ldots, k.
$$

- A general $c$ says the same about every linear combination $c^\top\beta$ of the coefficients, since $\Var{c^\top V \mid X} = c^\top \Var{V \mid X} c$:
$$
\Var{c^\top \tilde\beta \mid X} \ \ge\ \Var{c^\top \hat\beta \mid X} \qquad \text{for every } c \in \mathbb{R}^k.
$$

## Proof, step 1: linearity and unbiasedness force $A X = I_k$

- Take conditional expectations of $\tilde\beta = A Y$:
$$
\begin{aligned}
\beta = \E{\tilde\beta \mid X} &= \E{A(X\beta + U) \mid X} \\
&= A X \beta + A \underbrace{\E{U \mid X}}_{=\,0 \text{ by Assumption 2}} \\
&= A X \beta.
\end{aligned}
$$

- This must hold for every value of $\beta$, which is what unbiasedness means, so
$$
\boxed{\ A X = I_k\ }
$$

- Linearity plus conditional unbiasedness is exactly this one condition.

## Step 1 checked on the examples

- OLS: $A = (X^\top X)^{-1} X^\top$, so $A X = (X^\top X)^{-1} X^\top X = I_k$.

- Weighted least squares: $A = (X^\top W X)^{-1} X^\top W$, so $A X = (X^\top W X)^{-1} X^\top W X = I_k$.

- The two-point slope, with $k = 1$: $A$ has $-1/(X_n - X_1)$ in position 1 and $1/(X_n - X_1)$ in position $n$, so
$$
A X = \frac{X_n - X_1}{X_n - X_1} = 1.
$$

## Proof, step 2: the two sampling errors

- Using $A X = I_k$ for the general estimator, and the sampling error for OLS,
$$
\hat\beta = \beta + (X^\top X)^{-1} X^\top U, \qquad
\tilde\beta = A X \beta + A U = \beta + A U.
$$

- Both estimators differ from $\beta$ only through a known matrix times $U$. Everything that follows is an expectation of $U U^\top$.

## The transpose rule for covariance matrices

- For a $k \times 1$ random vector $V$ and an $\ell \times 1$ random vector $Z$,
$$
\Cov{V, Z} = \E{(V - \E{V})(Z - \E{Z})^\top} \qquad (k \times \ell),
$$
and transposing gives
$$
\Cov{V, Z}^\top = \E{(Z - \E{Z})(V - \E{V})^\top} = \Cov{Z, V}.
$$

- The same definition and the same rule hold conditionally, with $\E{\cdot}$ replaced by $\E{\cdot \mid X}$ throughout. Step 4 needs this, because both orders appear there.

## Proof, step 3: the covariance equals the variance of OLS

- Both estimators are conditionally unbiased, so their deviations from their conditional means are the two noise terms of step 2:
$$
\begin{aligned}
\Cov{\tilde\beta, \hat\beta \mid X}
&= \E{\big(\tilde\beta - \beta\big)\big(\hat\beta - \beta\big)^\top \,\Big|\, X} \\
&= \E{A U\, U^\top X (X^\top X)^{-1} \,\big|\, X} \\
&= A\, \underbrace{\E{U U^\top \mid X}}_{=\,\sigma^2 I_n \text{ by Assumptions 2 and 3}}\, X (X^\top X)^{-1} \\
&= \sigma^2 \underbrace{A X}_{=\, I_k} (X^\top X)^{-1} \\
&= \sigma^2 (X^\top X)^{-1}.
\end{aligned}
$$

- The right-hand side is $\Var{\hat\beta \mid X}$, so
$$
\boxed{\ \Cov{\tilde\beta, \hat\beta \mid X} = \Var{\hat\beta \mid X}\ }
$$

- Every linear conditionally unbiased estimator has the same covariance with OLS, namely the variance of OLS itself. This is the one surprising step in the proof.

## Proof, step 4: the difference of the variances

- As for a scalar, expand the variance of the difference using the definition; the cross terms are the two covariance matrices:
$$
\begin{aligned}
\Var{\tilde\beta - \hat\beta \mid X} = \Var{\tilde\beta \mid X} &+ {\color{blue}\Var{\hat\beta \mid X}} \\
&- {\color{blue}\Cov{\tilde\beta, \hat\beta \mid X}} - {\color{blue}\Cov{\hat\beta, \tilde\beta \mid X}}.
\end{aligned}
$$

- Step 3 gives the first covariance term, the second is its transpose by the rule above, and $\Var{\hat\beta \mid X}$ is symmetric. All three terms in blue equal $\Var{\hat\beta \mid X}$, and they combine into one with a minus sign:
$$
\Var{\tilde\beta - \hat\beta \mid X} = \Var{\tilde\beta \mid X} - \Var{\hat\beta \mid X}.
$$

- The left-hand side is a conditional variance matrix, hence positive semidefinite. Therefore
$$
\Var{\tilde\beta \mid X} - \Var{\hat\beta \mid X} \ \ge\ 0,
$$
which is the theorem.

## Uniqueness

- Suppose some linear conditionally unbiased $\tilde\beta$ attains the same variance matrix. By the identity of step 4,
$$
\Var{\tilde\beta \mid X} = \Var{\hat\beta \mid X}
\quad\Longrightarrow\quad
\Var{\tilde\beta - \hat\beta \mid X} = 0.
$$

- A random vector with zero conditional variance equals its conditional mean (the equality case of positive semidefiniteness, one coordinate at a time), so
$$
\tilde\beta - \hat\beta = d(X)
$$
for some function of $X$ alone.

- Both are conditionally unbiased, so taking conditional expectations,
$$
\underbrace{\E{\tilde\beta \mid X}}_{=\,\beta} = \underbrace{\E{\hat\beta \mid X}}_{=\,\beta} + d(X)
\quad\Longrightarrow\quad d(X) = 0.
$$

- Therefore $\tilde\beta = \hat\beta$. A linear conditionally unbiased competitor that is also best attains the same variance matrix: both $M = \Var{\tilde\beta \mid X} - \Var{\hat\beta \mid X}$ and $-M$ are then positive semidefinite, so $c^\top M c = 0$ for every $c$, and a symmetric matrix with that property is zero. So OLS is not just a best linear unbiased estimator; it is the only one.

## Where Assumption 3 was used

- Assumption 3 entered in step 3, where $\E{U U^\top \mid X}$ was replaced by $\sigma^2 I_n$, and again when $\sigma^2 (X^\top X)^{-1}$ was recognised as $\Var{\hat\beta \mid X}$. Without it the cancellation that produced $\sigma^2 A X$ fails, and the covariance need not equal the variance of OLS.

- Assumptions 1, 2 and 4 give unbiasedness, and Assumptions 1 and 4 with finite conditional second moments give the sandwich formula. Assumption 3 is what makes OLS efficient.

- If Assumption 3 fails, OLS is still unbiased but need not be best: Assumption 3 is sufficient for efficiency, not necessary.

# Two extensions, and the summary

## Adding normality

- **Assumption 5.** $U \mid X \sim N(0, \sigma^2 I_n)$. Assumptions 1 to 5 define the classical normal regression model.

- Given $X$, $\hat\beta = \beta + (X^\top X)^{-1}X^\top U$ is a constant plus a linear function of $U$, and Assumption 5 fixes the conditional distribution of $U$. For non-random $\alpha$ and $\Gamma$, such a function of a normal vector is normal:
$$
V \sim N(\mu, \Sigma) \quad\Longrightarrow\quad \alpha + \Gamma V \sim N\big(\alpha + \Gamma \mu,\ \Gamma \Sigma \Gamma^\top\big).
$$

- Under Assumptions 1, 4 and 5, since Assumption 5 implies Assumptions 2 and 3, the conditional mean and variance are the ones already computed:
$$
\boxed{\ \hat\beta \mid X \ \sim\ N\big(\beta,\ \sigma^2 (X^\top X)^{-1}\big)\ }
$$

- Nothing before this slide needed normality. It is what confidence intervals and tests will be built on.

## If Assumption 3 fails: generalised least squares

- Suppose now that $\Var{U \mid X} = \Omega$ for a known positive definite $\Omega$, not necessarily diagonal, with $\Omega \neq \sigma^2 I_n$ for every $\sigma^2 > 0$. Let $\Omega^{-1/2}$ be the symmetric positive definite matrix with $\Omega^{-1/2}\Omega^{-1/2} = \Omega^{-1}$, built from the eigenvalue decomposition $\Omega = C \Lambda C^\top$, and write $\Omega^{1/2}$ for the inverse of $\Omega^{-1/2}$.

- Multiply the model through by $\Omega^{-1/2}$:
$$
\underbrace{\Omega^{-1/2} Y}_{Y^*} = \underbrace{\Omega^{-1/2} X}_{X^*} \beta + \underbrace{\Omega^{-1/2} U}_{U^*},
\qquad \E{U^* \mid X} = 0, \qquad \Var{U^* \mid X} = \Omega^{-1/2} \Omega \Omega^{-1/2} = I_n,
$$
by taking out what is known and by the rule for a linear map.

- The transformed model satisfies all four assumptions ($X^*$ has rank $k$ because $\Omega^{-1/2}$ is nonsingular), so Gauss-Markov applies to it. Since $A \mapsto A\,\Omega^{1/2}$ matches the linear conditionally unbiased estimators of the two models one to one, OLS on the transformed model is also best among estimators linear in $Y$ and conditionally unbiased, and it is
$$
\hat\beta^* = (X^{*\top} X^*)^{-1} X^{*\top} Y^* = (X^\top \Omega^{-1} X)^{-1} X^\top \Omega^{-1} Y,
$$
which is weighted least squares with $W = \Omega^{-1}$.

- In practice $\Omega$ is not known. Later lectures return to this.

## Summary: unbiasedness and variance

- Substituting the model into the estimator gives the sampling error $\hat\beta = \beta + (X^\top X)^{-1} X^\top U$. Everything follows from it.

- **Unbiasedness.** Under Assumptions 1, 2 and 4, conditioning on $X$ makes $(X^\top X)^{-1}X^\top$ known, so $\E{\hat\beta \mid X} = \beta$, and the law of iterated expectations gives $\E{\hat\beta} = \beta$ whenever that mean exists. The weaker condition $\E{X_i U_i} = 0$ is not enough, because once $\E{U \mid X}$ is not zero, the expectation of the noise term need not vanish, and the parabola shows that zero correlation does not make the conditional mean zero.

- **Variance.** Under Assumptions 1 and 4, given finite conditional second moments,
$$
\Var{\hat\beta \mid X} = (X^\top X)^{-1} X^\top \Var{U \mid X} X (X^\top X)^{-1},
$$
and adding Assumption 3, that $\Var{U \mid X} = \sigma^2 I_n$, gives
$$
\Var{\hat\beta \mid X} = \sigma^2 (X^\top X)^{-1}.
$$
With one regressor and no intercept this is $\sigma^2 / \sum_i X_i^2$: noisier errors raise it; a larger $\sum_i X_i^2$ lowers it.

## Summary: efficiency

- **Gauss-Markov.** Under Assumptions 1 to 4, OLS has the smallest conditional variance among all estimators that are linear in $Y$ and conditionally unbiased, whatever the value of $\beta$. For every such $\tilde\beta$, the difference of the two conditional variance matrices is positive semidefinite:
$$
\Var{\tilde\beta \mid X} - \Var{\hat\beta \mid X} \ \ge\ 0,
$$
so in particular $\Var{\tilde\beta_j \mid X} \ge \Var{\hat\beta_j \mid X}$ for every coefficient.

- **The proof.** Linearity and conditional unbiasedness force $A X = I_k$. Assumption 3 then makes $\Cov{\tilde\beta, \hat\beta \mid X} = \Var{\hat\beta \mid X}$. Expanding $\Var{\tilde\beta - \hat\beta \mid X}$, which is positive semidefinite, leaves exactly the difference of the two variances.

- **Uniqueness.** Equality of the variances forces $\tilde\beta = \hat\beta$.

- Assumption 3 is what buys efficiency. Assumptions 1, 2 and 4 give unbiasedness on their own.
